---
title: "Foxj1a sensitivity on the global cell-type graph"
output: html_document
---

This figure asks a simple question: are the cell types that normally express
more `foxj1a` also the cell types that lose more motile-cilia activity after
foxj1a perturbation?

We keep the two measurements separate:

- node color shows normal foxj1a expression;
- node size shows motile-cilia score loss after perturbation;
- a black outline marks cell types passing BH FDR.

The concurrence helper in `0_shared/effect_concurrence.R` is intentionally not
used here. It compares two effects measured in the same units. Baseline gene
expression and a module-score change are different quantities, so combining
them into one number would be difficult to interpret.

```{r setup, message=FALSE, warning=FALSE}
library(igraph)
library(platt)
library(dplyr)
library(tibble)
library(readr)
library(ggplot2)
library(ggrepel)
library(stringr)
library(patchwork)

knitr::opts_chunk$set(
  dev = "ragg_png",
  fig.bg = "white"
)

PROJECT_DIR <- paste0(
  "/net/trapnell/vol1/home/mdcolon/proj/morphseq/results/",
  "mcolon/20260727_gene14_clean"
)
HERE <- file.path(
  PROJECT_DIR,
  "cilia_gene_module",
  "foxj1a_48hpf_inspection"
)
DATA_DIR <- file.path(HERE, "data")
FIGURE_DIR <- file.path(HERE, "figures")

CELLTYPE_RESULTS_FILE <- file.path(
  DATA_DIR,
  "celltype_baseline_and_motile_response.tsv"
)
CELLTYPE_LINEAGE_FILE <- file.path(
  PROJECT_DIR,
  "0_shared",
  "cell_type_lineage.tsv"
)
STATE_GRAPH_FILE <- paste0(
  "/net/seahub_zfish/vol1/data/graphs/",
  "v3.1.0/combined_graph.rds"
)

dir.create(FIGURE_DIR, recursive = TRUE, showWarnings = FALSE)
```

## Step 1: Join the measurements to the global graph

Most graph nodes are not present in this 48-hpf contrast. We keep those nodes
and draw them in light gray so the developmental map stays intact.

```{r prepare-graph}
celltype_results <- readr::read_tsv(
  CELLTYPE_RESULTS_FILE,
  show_col_types = FALSE
) |>
  dplyr::mutate(
    motile_score_loss = pmax(-.data$score_difference, 0)
  )

celltype_lineage <- readr::read_tsv(
  CELLTYPE_LINEAGE_FILE,
  show_col_types = FALSE
)

state_graph <- readRDS(STATE_GRAPH_FILE)

# Platt's layout keeps related cell types together in the familiar atlas map.
node_metadata <- tibble::tibble(
  id = igraph::V(state_graph)$name
) |>
  dplyr::left_join(
    dplyr::select(celltype_lineage, cell_type, sulston_group),
    by = c("id" = "cell_type")
  ) |>
  dplyr::mutate(
    group_nodes_by = dplyr::coalesce(.data$sulston_group, "Unknown")
  )

graph_layout <- platt:::layout_state_graph(
  G = state_graph,
  node_metadata = node_metadata
)

graph_nodes <- as.data.frame(graph_layout$gvizl_coords)
names(graph_nodes) <- c("x", "y")
graph_nodes <- graph_nodes |>
  tibble::rownames_to_column("cell_type") |>
  dplyr::left_join(celltype_results, by = "cell_type") |>
  dplyr::mutate(
    # Untested atlas nodes stay visible as small gray reference points.
    motile_score_loss_for_plot = dplyr::coalesce(.data$motile_score_loss, 0),
    fdr_outline_for_plot = dplyr::coalesce(.data$is_fdr_significant, FALSE)
  )

graph_edges <- graph_layout$bezier_df

readr::write_tsv(
  graph_nodes,
  file.path(DATA_DIR, "global_graph_node_metrics.tsv")
)

results_missing_from_graph <- celltype_results |>
  dplyr::anti_join(graph_nodes, by = "cell_type")
readr::write_tsv(
  results_missing_from_graph,
  file.path(DATA_DIR, "celltype_results_missing_from_global_graph.tsv")
)

message(
  sum(!is.na(graph_nodes$score_difference)),
  " tested cell types mapped onto ",
  nrow(graph_nodes),
  " global graph nodes."
)
```

## Step 2: Measure the relationship

The all-cell-type correlation answers the global question. The top-ten result
shows whether the relationship is concentrated in cell types that normally
express foxj1a, rather than spread uniformly across the atlas.

```{r correlations}
tested_nodes <- graph_nodes |>
  dplyr::filter(
    !is.na(.data$control_foxj1a_expression),
    !is.na(.data$score_difference)
  ) |>
  dplyr::arrange(dplyr::desc(.data$control_foxj1a_expression))

top_ten_nodes <- tested_nodes |>
  dplyr::slice_head(n = 10)

correlation_summary <- tibble::tibble(
  cell_types = c("all tested", "top 10 foxj1a-expressing"),
  n_cell_types = c(nrow(tested_nodes), nrow(top_ten_nodes)),
  pearson = c(
    cor(
      tested_nodes$control_foxj1a_expression,
      -tested_nodes$score_difference,
      method = "pearson"
    ),
    cor(
      top_ten_nodes$control_foxj1a_expression,
      -top_ten_nodes$score_difference,
      method = "pearson"
    )
  ),
  spearman = c(
    cor(
      tested_nodes$control_foxj1a_expression,
      -tested_nodes$score_difference,
      method = "spearman"
    ),
    cor(
      top_ten_nodes$control_foxj1a_expression,
      -top_ten_nodes$score_difference,
      method = "spearman"
    )
  )
)

readr::write_tsv(
  correlation_summary,
  file.path(DATA_DIR, "global_graph_expression_loss_correlations.tsv")
)

correlation_summary
```

## Step 3: Draw the global map

The first two panels show each measurement alone. The last panel overlays them:
large, brightly colored nodes are the cell types that both express foxj1a and
lose motile-cilia activity after perturbation. Color scales are capped at the
95th percentile so a few outliers do not flatten the rest of the graph. Node
size uses a looser 97.5th-percentile cap. The underlying values are not changed.

```{r draw-graph, fig.width=24, fig.height=9}
edge_layer <- geom_path(
  data = graph_edges,
  aes(x = .data$x, y = .data$y, group = .data$edge_name),
  color = "grey84",
  linewidth = 0.18,
  lineend = "round"
)

graph_theme <- theme_void(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10),
    legend.position = "bottom",
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8),
    plot.margin = margin(8, 8, 8, 8)
  )

# Robust display limits: only the plotting scales are capped. The node table
# written above always contains the original, uncapped measurements.
expression_color_limit <- stats::quantile(
  graph_nodes$control_foxj1a_expression,
  0.95,
  na.rm = TRUE
)
change_color_limit <- stats::quantile(
  abs(graph_nodes$score_difference),
  0.95,
  na.rm = TRUE
)
loss_size_limit <- stats::quantile(
  graph_nodes$motile_score_loss,
  0.975,
  na.rm = TRUE
)
graph_nodes <- graph_nodes |>
  dplyr::mutate(
    motile_score_loss_capped = pmin(
      .data$motile_score_loss_for_plot,
      .env$loss_size_limit
    )
  )

expression_graph <- ggplot() +
  edge_layer +
  geom_point(
    data = graph_nodes,
    aes(x = .data$x, y = .data$y, fill = .data$control_foxj1a_expression),
    shape = 21,
    size = 2.7,
    stroke = 0.2,
    color = "grey35"
  ) +
  scale_fill_viridis_c(
    option = "C",
    limits = c(0, expression_color_limit),
    oob = scales::squish,
    breaks = scales::breaks_pretty(n = 4),
    labels = scales::label_number(accuracy = 0.001),
    na.value = "grey90",
    name = "mean control foxj1a (95th-pct cap)",
    guide = guide_colorbar(
      title.position = "top",
      barwidth = grid::unit(2.8, "in"),
      barheight = grid::unit(0.15, "in")
    )
  ) +
  labs(
    title = "A. Normal foxj1a expression",
    subtitle = "Robust color scale; gray nodes were not tested at 48 hpf"
  ) +
  graph_theme

change_graph <- ggplot() +
  edge_layer +
  geom_point(
    data = graph_nodes,
    aes(x = .data$x, y = .data$y, fill = .data$score_difference),
    shape = 21,
    size = 2.7,
    stroke = 0.2,
    color = "grey35"
  ) +
  scale_fill_gradient2(
    low = "#3B6FB6",
    mid = "white",
    high = "#D95F59",
    midpoint = 0,
    limits = c(-change_color_limit, change_color_limit),
    oob = scales::squish,
    breaks = scales::breaks_pretty(n = 4),
    labels = scales::label_number(accuracy = 0.001),
    na.value = "grey90",
    name = "motile-score change (95th-pct cap)",
    guide = guide_colorbar(
      title.position = "top",
      barwidth = grid::unit(2.8, "in"),
      barheight = grid::unit(0.15, "in")
    )
  ) +
  labs(
    title = "B. Motile-cilia response",
    subtitle = "Blue means lower activity; robust color scale"
  ) +
  graph_theme

# Label the strongest linked responses, plus the ependymoglial result that
# extends the radial-glia pattern. Keeping this list explicit makes it auditable.
cell_types_to_label <- c(
  "glial cell, floor plate (hpgd+)",
  "ciliated olfactory receptor neuron, olfactory organ",
  "radial glial cell, spinal cord",
  "radial glial cell, midbrain",
  "endodermal cell, intestinal bulb",
  "ependymoglial cell, central nervous system"
)

overlay_labels <- graph_nodes |>
  dplyr::filter(.data$cell_type %in% cell_types_to_label)

overlay_graph <- ggplot() +
  edge_layer +
  geom_point(
    data = graph_nodes,
    aes(
      x = .data$x,
      y = .data$y,
      fill = .data$control_foxj1a_expression,
      size = .data$motile_score_loss_capped,
      color = .data$fdr_outline_for_plot
    ),
    shape = 21,
    stroke = 0.7
  ) +
  ggrepel::geom_text_repel(
    data = overlay_labels,
    aes(x = .data$x, y = .data$y, label = .data$cell_type),
    size = 2.5,
    max.overlaps = Inf,
    min.segment.length = 0,
    box.padding = 0.6,
    point.padding = 0.25,
    force = 8,
    seed = 1
  ) +
  scale_fill_viridis_c(
    option = "C",
    limits = c(0, expression_color_limit),
    oob = scales::squish,
    breaks = scales::breaks_pretty(n = 4),
    labels = scales::label_number(accuracy = 0.001),
    na.value = "grey90",
    name = "mean control foxj1a (95th-pct cap)",
    guide = guide_colorbar(
      title.position = "top",
      barwidth = grid::unit(2.8, "in"),
      barheight = grid::unit(0.15, "in"),
      order = 1
    )
  ) +
  scale_size_continuous(
    # Match Panel A at zero loss; module loss can only enlarge a node.
    range = c(2.7, 8),
    limits = c(0, loss_size_limit),
    breaks = c(0, loss_size_limit / 2, loss_size_limit),
    labels = scales::label_number(accuracy = 0.001),
    name = "motile-score loss (97.5th-pct cap)",
    guide = guide_legend(
      title.position = "top",
      nrow = 1,
      order = 2
    )
  ) +
  scale_color_manual(
    values = c(`TRUE` = "black", `FALSE` = "grey45"),
    na.value = "grey75",
    guide = "none"
  ) +
  labs(
    title = "C. Foxj1a-linked sensitivity",
    subtitle = paste0(
      "Expression versus loss across tested nodes: Pearson r = ",
      sprintf("%.2f", correlation_summary$pearson[1]),
      "; Spearman rho = ",
      sprintf("%.2f", correlation_summary$spearman[1]),
      ". Top 10: r = ",
      sprintf("%.2f", correlation_summary$pearson[2]),
      ", rho = ",
      sprintf("%.2f", correlation_summary$spearman[2]),
      ". Black outline = BH q < 0.05."
    )
  ) +
  graph_theme +
  theme(legend.box = "vertical")

global_graph_figure <- expression_graph | change_graph | overlay_graph

ggsave(
  file.path(FIGURE_DIR, "09_foxj1a_global_celltype_graph.pdf"),
  global_graph_figure,
  width = 24,
  height = 9,
  limitsize = FALSE
)
ggsave(
  file.path(FIGURE_DIR, "09_foxj1a_global_celltype_graph.png"),
  global_graph_figure,
  device = ragg::agg_png,
  width = 24,
  height = 9,
  units = "in",
  dpi = 200,
  bg = "white",
  limitsize = FALSE
)

global_graph_figure
```
