---
title: "Inspect foxj1a cilia-module expression at 48 hpf"
output: html_document
---

This report examines the first differential cilia-module result from several
angles. It keeps the embryo-level observations visible and treats component
genes as descriptive validation rather than a second significance screen.

```{r setup, message=FALSE, warning=FALSE}
library(monocle3)
library(Matrix)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(ggplot2)
library(jsonlite)
library(patchwork)

knitr::opts_chunk$set(
  dev = "ragg_png",
  fig.bg = "white"
)

PROJECT_DIR <- paste0(
  "/net/trapnell/vol1/home/mdcolon/proj/morphseq/results/",
  "mcolon/20260727_gene14_clean"
)
MODULE_DIR <- file.path(PROJECT_DIR, "cilia_gene_module")
HERE <- file.path(MODULE_DIR, "foxj1a_48hpf_inspection")
DATA_DIR <- file.path(HERE, "data")
FIGURE_DIR <- file.path(HERE, "figures")

dir.create(DATA_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(FIGURE_DIR, recursive = TRUE, showWarnings = FALSE)

PROJECT_CONFIG <- jsonlite::read_json(
  file.path(PROJECT_DIR, "project_config.json"),
  simplifyVector = TRUE
)

GENE14_CDS <- PROJECT_CONFIG$mcclintock_cds
EMBRYO_MODULE_FILE <- file.path(
  MODULE_DIR,
  "output",
  "embryo_celltype_cilia_module_scores.tsv"
)
MODEL_RESULT_FILE <- file.path(
  MODULE_DIR,
  "output",
  "differential_module_results",
  "foxj1a_48hpf.tsv"
)
HUMAN_MODULE_FILE <- file.path(
  MODULE_DIR,
  "data",
  "human_cilia_modules.tsv"
)
ORTHOLOGY_FILE <- file.path(
  MODULE_DIR,
  "data",
  "human_to_zebrafish_cilia_map.tsv"
)

CONTRAST_NAME <- "foxj1a_48hpf"
PERTURBATION <- "foxj1a"
CONTROL <- "ctrl-inj"
TIMEPOINT <- 48
MINIMUM_CELLS <- 10L
FOCUS_CELL_TYPE <- "radial glial cell, spinal cord"
FOCUS_MODULE <- "gold_motile_cilia"
FOXJ1A_GENE <- "foxj1a"
MAX_COMPONENT_GENES <- 40L

source(file.path(MODULE_DIR, "cilia_gene_module_policy.R"))

save_plot_pair <- function(file_stem, plot, width, height) {
  ggsave(
    file.path(FIGURE_DIR, paste0(file_stem, ".pdf")),
    plot,
    width = width,
    height = height,
    limitsize = FALSE
  )
  ggsave(
    file.path(FIGURE_DIR, paste0(file_stem, ".png")),
    plot,
    device = ragg::agg_png,
    width = width,
    height = height,
    dpi = 200,
    bg = "white",
    limitsize = FALSE
  )
}

arm_colors <- c(
  control = "#4C78A8",
  perturbation = "#E45756"
)
```

## Step 1: Read the completed model and embryo scores

```{r read-results}
module_results <- readr::read_tsv(
  MODEL_RESULT_FILE,
  show_col_types = FALSE
)
embryo_module_scores <- readr::read_tsv(
  EMBRYO_MODULE_FILE,
  show_col_types = FALSE
)

if (!all(module_results$contrast_name == CONTRAST_NAME)) {
  stop("The model result file contains the wrong contrast.")
}

contrast_embryo_scores <- embryo_module_scores |>
  dplyr::filter(
    .data$timepoint == TIMEPOINT,
    .data$perturbation %in% c(PERTURBATION, CONTROL),
    .data$n_cells >= MINIMUM_CELLS
  ) |>
  dplyr::mutate(
    arm = dplyr::if_else(
      .data$perturbation == PERTURBATION,
      "perturbation",
      "control"
    ),
    arm = factor(.data$arm, levels = c("control", "perturbation")),
    hash_plate = stringr::str_extract(.data$embryo_ID, "P\\d+")
  )

module_results |>
  dplyr::count(.data$module_id, .data$is_fdr_significant)
```

## Step 2: Show every embryo for the significant results

One dot is one embryo. The black bar is the group mean.

```{r significant-embryo-values, fig.width=11, fig.height=10}
significant_results <- module_results |>
  dplyr::filter(.data$is_fdr_significant) |>
  dplyr::select(
    module_id,
    cell_type,
    score_difference,
    standardized_effect,
    q_value
  )

significant_embryo_values <- contrast_embryo_scores |>
  dplyr::inner_join(
    significant_results,
    by = c("module_id", "cell_type")
  ) |>
  dplyr::mutate(
    panel_label = paste0(
      .data$module_id,
      "\n",
      .data$cell_type,
      "\nq = ",
      format(.data$q_value, digits = 2, scientific = TRUE)
    )
  )

readr::write_tsv(
  dplyr::select(significant_embryo_values, -panel_label),
  file.path(DATA_DIR, "significant_embryo_values.tsv")
)

significant_plot <- ggplot(
  significant_embryo_values,
  aes(
    x = .data$arm,
    y = .data$mean_module_score,
    color = .data$arm
  )
) +
  geom_jitter(width = 0.10, height = 0, size = 2, alpha = 0.85) +
  stat_summary(
    fun = mean,
    geom = "crossbar",
    width = 0.5,
    color = "black",
    linewidth = 0.35
  ) +
  facet_wrap(vars(.data$panel_label), scales = "free_y", ncol = 2) +
  scale_color_manual(values = arm_colors, guide = "none") +
  theme_minimal(base_size = 10) +
  theme(
    axis.title.x = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 8)
  ) +
  labs(
    title = "Foxj1a 48-hpf significant cilia-module results",
    subtitle = "Each point is one embryo; black bar is the group mean",
    y = "mean module score"
  )

save_plot_pair(
  "01_significant_embryo_values",
  significant_plot,
  width = 11,
  height = 10
)

significant_plot
```

## Step 3: Compare all five modules in spinal-cord radial glia

Free y-axes are intentional because the five curated modules have different
score ranges.

```{r focus-module-profile, fig.width=12, fig.height=7}
focus_module_profile <- contrast_embryo_scores |>
  dplyr::filter(
    .data$cell_type == FOCUS_CELL_TYPE,
    .data$module_id %in% CHOSEN_CILIA_MODULES
  )

readr::write_tsv(
  focus_module_profile,
  file.path(DATA_DIR, "spinal_radial_glia_five_modules.tsv")
)

focus_module_plot <- ggplot(
  focus_module_profile,
  aes(
    x = .data$arm,
    y = .data$mean_module_score,
    color = .data$arm
  )
) +
  geom_jitter(width = 0.10, height = 0, size = 2, alpha = 0.85) +
  stat_summary(
    fun = mean,
    geom = "crossbar",
    width = 0.5,
    color = "black",
    linewidth = 0.35
  ) +
  facet_wrap(vars(.data$module_id), scales = "free_y", ncol = 3) +
  scale_color_manual(values = arm_colors, guide = "none") +
  theme_minimal(base_size = 10) +
  theme(
    axis.title.x = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Five cilia programs in spinal-cord radial glia",
    subtitle = "Each point is one embryo",
    y = "mean module score"
  )

save_plot_pair(
  "02_spinal_radial_glia_five_modules",
  focus_module_plot,
  width = 12,
  height = 7
)

focus_module_plot
```

## Step 4: Ask whether the motile-cilia effect is cell-type specific

The forest plot shows every eligible cell type. Filled red points pass the BH
threshold; open gray points do not.

```{r motile-forest, fig.width=10, fig.height=18}
motile_results <- module_results |>
  dplyr::filter(.data$module_id == FOCUS_MODULE) |>
  dplyr::arrange(.data$score_difference) |>
  dplyr::mutate(
    cell_type = factor(.data$cell_type, levels = .data$cell_type)
  )

readr::write_tsv(
  motile_results,
  file.path(DATA_DIR, "gold_motile_all_cell_types.tsv")
)

motile_forest_plot <- ggplot(
  motile_results,
  aes(
    x = .data$score_difference,
    y = .data$cell_type
  )
) +
  geom_segment(
    aes(
      x = .data$confidence_interval_low,
      xend = .data$confidence_interval_high,
      yend = .data$cell_type
    ),
    color = "gray55",
    linewidth = 0.35
  ) +
  geom_point(
    aes(fill = .data$is_fdr_significant),
    shape = 21,
    size = 2,
    color = "black"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray35") +
  scale_fill_manual(
    values = c(`FALSE` = "white", `TRUE` = "#E45756"),
    name = "BH q < 0.05"
  ) +
  theme_minimal(base_size = 8) +
  theme(
    axis.title.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    title = "Foxj1a effect on the gold motile-cilia module",
    subtitle = "Perturbation minus control; lines are 95% confidence intervals",
    x = "module-score difference"
  )

save_plot_pair(
  "03_gold_motile_forest_all_cell_types",
  motile_forest_plot,
  width = 10,
  height = 18
)

motile_forest_plot
```

## Step 5: Relate normal foxj1a expression to the motile-cilia response

The large CDS is loaded once. We measure baseline `foxj1a` expression only in
control embryos, with each embryo contributing equally to its cell-type mean.
The response remains the embryo-level model estimate from Steps 1–4.

```{r baseline-response, fig.width=15, fig.height=5.5}
message("Loading GENE14 CDS...")
cds <- monocle3::load_monocle_objects(GENE14_CDS)

cell_metadata <- as.data.frame(SummarizedExperiment::colData(cds))
cell_metadata$cell <- rownames(cell_metadata)

# Use exactly the control embryo and cell-type combinations that passed the
# module model's minimum-cell requirement.
eligible_control_pairs <- contrast_embryo_scores |>
  dplyr::filter(
    .data$module_id == FOCUS_MODULE,
    .data$arm == "control",
    .data$cell_type %in% as.character(motile_results$cell_type)
  ) |>
  dplyr::distinct(.data$embryo_ID, .data$cell_type) |>
  dplyr::mutate(
    pair_key = paste(.data$embryo_ID, .data$cell_type, sep = "___")
  )

cell_pair_key <- paste(
  cell_metadata$embryo_ID,
  cell_metadata$cell_type,
  sep = "___"
)
keep_baseline_cells <- (
  !is.na(cell_metadata$timepoint) &
    cell_metadata$timepoint == TIMEPOINT &
    !is.na(cell_metadata$perturbation) &
    cell_metadata$perturbation == CONTROL &
    cell_pair_key %in% eligible_control_pairs$pair_key
)

gene_metadata <- as.data.frame(SummarizedExperiment::rowData(cds))
foxj1a_rows <- (
  !is.na(gene_metadata$gene_short_name) &
    gene_metadata$gene_short_name == FOXJ1A_GENE
)
if (sum(foxj1a_rows) != 1L) {
  stop("Expected exactly one foxj1a row in the CDS.")
}

foxj1a_cds <- cds[foxj1a_rows, keep_baseline_cells]
foxj1a_counts <- as.numeric(
  as.matrix(SingleCellExperiment::counts(foxj1a_cds))[1, ]
)
foxj1a_expression <- as.numeric(
  as.matrix(
    monocle3::normalized_counts(
      foxj1a_cds,
      norm_method = "size_only"
    )
  )[1, ]
)

foxj1a_by_embryo <- cell_metadata[keep_baseline_cells, ] |>
  tibble::as_tibble() |>
  dplyr::transmute(
    embryo_ID = .data$embryo_ID,
    cell_type = .data$cell_type,
    foxj1a_expression = log1p(foxj1a_expression),
    foxj1a_detected = foxj1a_counts > 0
  ) |>
  dplyr::group_by(.data$embryo_ID, .data$cell_type) |>
  dplyr::summarise(
    mean_foxj1a_expression = mean(.data$foxj1a_expression),
    foxj1a_detection_fraction = mean(.data$foxj1a_detected),
    .groups = "drop"
  )

foxj1a_by_cell_type <- foxj1a_by_embryo |>
  dplyr::group_by(.data$cell_type) |>
  dplyr::summarise(
    control_foxj1a_expression = mean(.data$mean_foxj1a_expression),
    control_foxj1a_detection_fraction = mean(
      .data$foxj1a_detection_fraction
    ),
    .groups = "drop"
  )

control_motile_by_cell_type <- contrast_embryo_scores |>
  dplyr::filter(
    .data$module_id == FOCUS_MODULE,
    .data$arm == "control",
    .data$cell_type %in% as.character(motile_results$cell_type)
  ) |>
  dplyr::group_by(.data$cell_type) |>
  dplyr::summarise(
    control_motile_score = mean(.data$mean_module_score),
    .groups = "drop"
  )

baseline_response <- motile_results |>
  dplyr::mutate(cell_type = as.character(.data$cell_type)) |>
  dplyr::left_join(foxj1a_by_cell_type, by = "cell_type") |>
  dplyr::left_join(control_motile_by_cell_type, by = "cell_type")

if (anyNA(baseline_response$control_foxj1a_expression)) {
  stop("Some modeled cell types are missing baseline foxj1a expression.")
}

readr::write_tsv(
  foxj1a_by_embryo,
  file.path(DATA_DIR, "control_foxj1a_expression_by_embryo.tsv")
)
readr::write_tsv(
  baseline_response,
  file.path(DATA_DIR, "celltype_baseline_and_motile_response.tsv")
)

top_foxj1a_cell_types <- baseline_response |>
  dplyr::slice_max(
    .data$control_foxj1a_expression,
    n = 10,
    with_ties = FALSE
  ) |>
  dplyr::pull("cell_type")

volcano_labels <- union(
  baseline_response |>
    dplyr::filter(.data$is_fdr_significant) |>
    dplyr::pull("cell_type"),
  baseline_response |>
    dplyr::slice_max(
      abs(.data$score_difference),
      n = 3,
      with_ties = FALSE
    ) |>
    dplyr::pull("cell_type")
)

response_theme <- theme_minimal(base_size = 9) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

pearson_result <- stats::cor.test(
  baseline_response$control_foxj1a_expression,
  baseline_response$control_motile_score,
  method = "pearson"
)
spearman_result <- stats::cor.test(
  baseline_response$control_foxj1a_expression,
  baseline_response$control_motile_score,
  method = "spearman",
  exact = FALSE
)
correlation_label <- paste0(
  "Pearson r = ",
  format(unname(pearson_result$estimate), digits = 2),
  "\nSpearman rho = ",
  format(unname(spearman_result$estimate), digits = 2)
)

baseline_correlation_plot <- ggplot(
  baseline_response,
  aes(
    x = .data$control_foxj1a_expression,
    y = .data$control_motile_score,
    color = .data$control_foxj1a_expression
  )
) +
  geom_point(size = 2.5, alpha = 0.9) +
  ggrepel::geom_text_repel(
    data = dplyr::filter(
      baseline_response,
      .data$cell_type %in% top_foxj1a_cell_types
    ),
    aes(label = .data$cell_type),
    size = 2.5,
    seed = 1,
    force = 2,
    box.padding = 0.4,
    min.segment.length = 0,
    max.time = 2,
    max.overlaps = Inf,
    show.legend = FALSE
  ) +
  annotate(
    "label",
    x = Inf,
    y = -Inf,
    label = correlation_label,
    hjust = 1.05,
    vjust = -0.1,
    size = 3
  ) +
  scale_color_viridis_c(
    name = "control foxj1a expression",
    guide = guide_colorbar(
      direction = "horizontal",
      title.position = "top",
      barwidth = 5,
      barheight = 0.5
    )
  ) +
  response_theme +
  labs(
    title = "A. Foxj1a marks the motile-cilia program",
    x = "mean control foxj1a expression",
    y = "mean control gold-motile score"
  )

volcano_data <- baseline_response |>
  dplyr::mutate(
    minus_log10_q = -log10(pmax(.data$q_value, .Machine$double.xmin))
  )

volcano_plot <- ggplot(
  volcano_data,
  aes(
    x = .data$score_difference,
    y = .data$minus_log10_q,
    color = .data$control_foxj1a_expression
  )
) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed",
    color = "gray45"
  ) +
  geom_vline(xintercept = 0, color = "gray65") +
  geom_point(size = 2.5, alpha = 0.9) +
  ggrepel::geom_text_repel(
    data = dplyr::filter(
      volcano_data,
      .data$cell_type %in% volcano_labels
    ),
    aes(label = .data$cell_type),
    size = 2.5,
    seed = 1,
    force = 2,
    box.padding = 0.4,
    min.segment.length = 0,
    max.time = 2,
    max.overlaps = Inf,
    show.legend = FALSE
  ) +
  scale_color_viridis_c(
    name = "control foxj1a expression",
    guide = guide_colorbar(
      direction = "horizontal",
      title.position = "top",
      barwidth = 5,
      barheight = 0.5
    )
  ) +
  response_theme +
  labs(
    title = "B. Motile-score changes across cell types",
    x = "motile-score change (foxj1a - control)",
    y = "-log10(BH q)"
  )

top_validation_cell_types <- baseline_response |>
  dplyr::slice_min(.data$q_value, n = 6, with_ties = FALSE) |>
  dplyr::arrange(.data$q_value) |>
  dplyr::pull("cell_type")
top_validation_labels <- stringr::str_wrap(top_validation_cell_types, 24)

embryo_validation <- contrast_embryo_scores |>
  dplyr::filter(
    .data$module_id == FOCUS_MODULE,
    .data$cell_type %in% top_validation_cell_types
  ) |>
  dplyr::mutate(
    short_cell_type = stringr::str_wrap(.data$cell_type, 24),
    short_cell_type = factor(
      .data$short_cell_type,
      levels = top_validation_labels
    )
  )

embryo_validation_plot <- ggplot(
  embryo_validation,
  aes(
    x = .data$arm,
    y = .data$mean_module_score,
    color = .data$arm
  )
) +
  geom_jitter(width = 0.10, height = 0, size = 2, alpha = 0.85) +
  stat_summary(
    fun = mean,
    geom = "crossbar",
    width = 0.5,
    color = "black",
    linewidth = 0.35
  ) +
  facet_wrap(vars(.data$short_cell_type), scales = "free_y", ncol = 3) +
  scale_color_manual(values = arm_colors, guide = "none") +
  response_theme +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(size = 7),
    strip.text = element_text(size = 7),
    panel.grid = element_blank()
  ) +
  labs(
    title = "C. Top six results across embryos",
    y = "mean gold-motile score"
  )

baseline_response_figure <- (
  baseline_correlation_plot |
    volcano_plot |
    embryo_validation_plot
) +
  patchwork::plot_layout(
    widths = c(1, 1, 1.3),
    guides = "collect"
  ) +
  patchwork::plot_annotation(
    title = "Foxj1a perturbation reduces the motile-cilia program in radial glia",
    subtitle = paste(
      "Cell types define Panels A-B; embryos define Panel C;",
      "Panel B labels the FDR hits and largest changes"
    )
  ) &
  theme(legend.position = "bottom")

save_plot_pair(
  "04_foxj1a_motile_baseline_and_response",
  baseline_response_figure,
  width = 16,
  height = 6
)

baseline_response_figure
```

## All cell types ordered by evidence and effect

This overview keeps every eligible cell type. Panels are ordered by BH q-value
and then by motile-score change, with the strongest losses first when q-values
tie. `Fox rank` orders the same cell types by control `foxj1a` expression.
Each point remains one embryo; the black bar is the arm mean.

```{r all-celltype-embryo-grid, fig.width=16, fig.height=24}
ordered_cell_types <- baseline_response |>
  dplyr::arrange(.data$q_value, .data$score_difference) |>
  dplyr::mutate(
    result_rank = dplyr::row_number(),
    foxj1a_rank = rank(
      -.data$control_foxj1a_expression,
      ties.method = "min"
    ),
    short_cell_type = stringr::str_wrap(.data$cell_type, 22),
    panel_label = paste0(
      "#", .data$result_rank,
      " | Fox rank ", .data$foxj1a_rank,
      "\n", .data$short_cell_type,
      "\ndelta ", format(.data$score_difference, digits = 2),
      " | q ", format(.data$q_value, digits = 2, scientific = TRUE)
    )
  )

readr::write_tsv(
  dplyr::select(
    ordered_cell_types,
    -short_cell_type,
    -panel_label
  ),
  file.path(DATA_DIR, "all_celltypes_ordered_for_embryo_grid.tsv")
)

all_celltype_embryos <- contrast_embryo_scores |>
  dplyr::filter(.data$module_id == FOCUS_MODULE) |>
  dplyr::inner_join(
    ordered_cell_types |>
      dplyr::select(cell_type, panel_label, result_rank),
    by = "cell_type"
  ) |>
  dplyr::mutate(
    panel_label = factor(
      .data$panel_label,
      levels = ordered_cell_types$panel_label
    )
  )

all_celltype_grid <- ggplot(
  all_celltype_embryos,
  aes(
    x = .data$arm,
    y = .data$mean_module_score,
    color = .data$arm
  )
) +
  geom_jitter(width = 0.10, height = 0, size = 0.8, alpha = 0.8) +
  stat_summary(
    fun = mean,
    geom = "crossbar",
    width = 0.55,
    color = "black",
    linewidth = 0.25
  ) +
  facet_wrap(
    vars(.data$panel_label),
    scales = "free_y",
    ncol = 8
  ) +
  scale_color_manual(
    values = arm_colors,
    labels = c(control = "control", perturbation = "foxj1a")
  ) +
  theme_void(base_size = 7) +
  theme(
    strip.text = element_text(size = 5.5, lineheight = 0.9),
    legend.position = "bottom",
    plot.title = element_text(size = 14),
    plot.subtitle = element_text(size = 9),
    panel.spacing = grid::unit(0.8, "lines")
  ) +
  labs(
    title = "All 48-hpf foxj1a gold-motile results",
    subtitle = paste(
      "Ordered by BH q, then motile-score change;",
      "each point is one embryo and black bars are arm means"
    ),
    color = NULL
  )

ggsave(
  file.path(FIGURE_DIR, "08_all_celltypes_ordered_embryo_grid.png"),
  all_celltype_grid,
  device = ragg::agg_png,
  width = 16,
  height = 24,
  dpi = 200,
  bg = "white",
  limitsize = FALSE
)

all_celltype_grid
```

## Correlation sensitivity among the most foxj1a-expressing cell types

The full rank correlation includes many cell types with little or no detected
`foxj1a`. This descriptive sensitivity analysis shows both Pearson and
Spearman correlations as lower-ranked cell types are added. It is not used to
choose a new significance threshold.

```{r spearman-sensitivity, fig.width=8, fig.height=5}
ranked_cell_types <- baseline_response |>
  dplyr::arrange(dplyr::desc(.data$control_foxj1a_expression))

numbers_to_include <- 10:nrow(ranked_cell_types)
correlation_rows <- vector("list", length(numbers_to_include))

for (i in seq_along(numbers_to_include)) {
  n_cell_types <- numbers_to_include[[i]]
  included_cell_types <- dplyr::slice_head(
    ranked_cell_types,
    n = n_cell_types
  )

  correlation_rows[[i]] <- tibble::tibble(
    n_cell_types = n_cell_types,
    mean_foxj1a_expression = mean(
      included_cell_types$control_foxj1a_expression
    ),
    minimum_foxj1a_expression = min(
      included_cell_types$control_foxj1a_expression
    ),
    pearson_r = stats::cor(
      included_cell_types$control_foxj1a_expression,
      included_cell_types$control_motile_score,
      method = "pearson"
    ),
    spearman_rho = stats::cor(
      included_cell_types$control_foxj1a_expression,
      included_cell_types$control_motile_score,
      method = "spearman"
    )
  )
}

correlation_sensitivity <- dplyr::bind_rows(correlation_rows)
readr::write_tsv(
  correlation_sensitivity,
  file.path(DATA_DIR, "correlation_top_foxj1a_sensitivity.tsv")
)

first_zero_expression <- correlation_sensitivity |>
  dplyr::filter(.data$minimum_foxj1a_expression == 0) |>
  dplyr::slice_head(n = 1) |>
  dplyr::pull("n_cell_types")

correlation_plot_data <- correlation_sensitivity |>
  tidyr::pivot_longer(
    cols = c("pearson_r", "spearman_rho"),
    names_to = "correlation_method",
    values_to = "correlation"
  ) |>
  dplyr::mutate(
    correlation_method = dplyr::recode(
      .data$correlation_method,
      pearson_r = "Pearson r",
      spearman_rho = "Spearman rho"
    )
  )

correlations_to_label <- correlation_plot_data |>
  dplyr::filter(
    .data$n_cell_types %in% c(10L, 20L, nrow(ranked_cell_types))
  ) |>
  dplyr::mutate(
    label = paste0(
      "top ",
      .data$n_cell_types,
      ": ",
      format(.data$correlation, digits = 2)
    )
  )

correlation_sensitivity_plot <- ggplot(
  correlation_plot_data,
  aes(
    x = .data$n_cell_types,
    y = .data$correlation,
    color = .data$correlation_method
  )
) +
  geom_hline(yintercept = 0, color = "gray65") +
  geom_vline(
    xintercept = first_zero_expression,
    linetype = "dashed",
    color = "gray45"
  ) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.4) +
  ggrepel::geom_label_repel(
    data = correlations_to_label,
    aes(label = .data$label),
    size = 3,
    max.overlaps = Inf,
    show.legend = FALSE
  ) +
  annotate(
    "text",
    x = first_zero_expression,
    y = -Inf,
    label = "zero-expression cell types begin to enter",
    angle = 90,
    hjust = -0.05,
    vjust = -0.4,
    size = 3,
    color = "gray35"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    plot.margin = margin(5, 5, 0, 5)
  ) +
  scale_color_manual(
    values = c("Pearson r" = "#E45756", "Spearman rho" = "#4C78A8"),
    name = NULL
  ) +
  labs(
    x = NULL,
    y = "correlation"
  )

mean_expression_plot <- ggplot(
  correlation_sensitivity,
  aes(
    x = .data$n_cell_types,
    y = .data$mean_foxj1a_expression
  )
) +
  geom_vline(
    xintercept = first_zero_expression,
    linetype = "dashed",
    color = "gray45"
  ) +
  geom_line(linewidth = 0.8, color = "#7A5195") +
  geom_point(size = 1.4, color = "#7A5195") +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor = element_blank(),
    plot.margin = margin(0, 5, 5, 5)
  ) +
  labs(
    x = "top foxj1a-expressing cell types included",
    y = "mean control\nfoxj1a expression"
  )

correlation_sensitivity_figure <-
  correlation_sensitivity_plot /
  mean_expression_plot +
  patchwork::plot_layout(heights = c(2.2, 1)) +
  patchwork::plot_annotation(
    title = "Foxj1a-motile correlation sensitivity",
    subtitle = paste(
      "Cell types are added from highest to lowest control foxj1a expression;",
      "the lower panel shows how their mean expression falls"
    )
  )

save_plot_pair(
  "07_correlation_sensitivity",
  correlation_sensitivity_figure,
  width = 8,
  height = 7
)

correlation_sensitivity_figure
```

## Step 6: Load spinal-cord radial glia for focused diagnostics

Every subsequent calculation is restricted to 48-hpf spinal-cord radial glia
from foxj1a and control-injected embryos.

```{r load-focus-cds}
eligible_focus_embryos <- unique(focus_module_profile$embryo_ID)
keep_cells <- (
  !is.na(cell_metadata$timepoint) &
    cell_metadata$timepoint == TIMEPOINT &
    !is.na(cell_metadata$perturbation) &
    cell_metadata$perturbation %in% c(PERTURBATION, CONTROL) &
    !is.na(cell_metadata$cell_type) &
    cell_metadata$cell_type == FOCUS_CELL_TYPE &
    !is.na(cell_metadata$embryo_ID) &
    cell_metadata$embryo_ID %in% eligible_focus_embryos
)

focus_cds <- cds[, keep_cells]
rm(cds)
invisible(gc())

focus_cell_metadata <- as.data.frame(
  SummarizedExperiment::colData(focus_cds)
)
focus_cell_metadata$cell <- rownames(focus_cell_metadata)
focus_cell_metadata <- tibble::as_tibble(focus_cell_metadata) |>
  dplyr::transmute(
    cell = .data$cell,
    embryo_ID = .data$embryo_ID,
    perturbation = .data$perturbation,
    arm = dplyr::if_else(
      .data$perturbation == PERTURBATION,
      "perturbation",
      "control"
    ),
    rt_block = .data$rt_block,
    hash_plate = stringr::str_extract(.data$embryo_ID, "P\\d+")
  )

focus_cell_metadata |>
  dplyr::distinct(
    .data$perturbation,
    .data$arm,
    .data$rt_block,
    .data$hash_plate
  )
```

Controls and foxj1a share RT block `Bl1`, but their hash plates are different.
Because plate and condition are confounded, the following depth diagnostics
are important; a separate plate coefficient cannot resolve this design.

## Step 7: Check cell count and UMI depth

```{r depth-diagnostics, fig.width=10, fig.height=5}
focus_counts <- SingleCellExperiment::counts(focus_cds)
if (methods::is(focus_counts, "IterableMatrix")) {
  total_umi <- BPCells::colSums(focus_counts)
} else {
  total_umi <- Matrix::colSums(focus_counts)
}

cell_depth <- focus_cell_metadata |>
  dplyr::mutate(total_umi = as.numeric(total_umi))

embryo_depth <- cell_depth |>
  dplyr::group_by(
    .data$embryo_ID,
    .data$arm,
    .data$rt_block,
    .data$hash_plate
  ) |>
  dplyr::summarise(
    n_cells_from_cds = dplyr::n(),
    median_umi_per_cell = stats::median(.data$total_umi),
    mean_umi_per_cell = mean(.data$total_umi),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    focus_module_profile |>
      dplyr::filter(.data$module_id == FOCUS_MODULE) |>
      dplyr::select(
        embryo_ID,
        n_cells,
        mean_module_score
      ),
    by = "embryo_ID"
  )

if (any(embryo_depth$n_cells_from_cds != embryo_depth$n_cells)) {
  stop("Focused CDS cell counts do not match the embryo score artifact.")
}

readr::write_tsv(
  embryo_depth,
  file.path(DATA_DIR, "spinal_radial_glia_depth_diagnostics.tsv")
)

depth_plot_data <- embryo_depth |>
  dplyr::select(
    embryo_ID,
    arm,
    hash_plate,
    mean_module_score,
    n_cells,
    median_umi_per_cell
  ) |>
  tidyr::pivot_longer(
    cols = c(n_cells, median_umi_per_cell),
    names_to = "depth_measure",
    values_to = "depth_value"
  ) |>
  dplyr::mutate(
    depth_measure = dplyr::recode(
      .data$depth_measure,
      n_cells = "cells from this embryo",
      median_umi_per_cell = "median UMIs per cell"
    )
  )

depth_plot <- ggplot(
  depth_plot_data,
  aes(
    x = .data$depth_value,
    y = .data$mean_module_score,
    color = .data$arm,
    shape = .data$hash_plate
  )
) +
  geom_point(size = 2.5, alpha = 0.9) +
  facet_wrap(vars(.data$depth_measure), scales = "free_x") +
  scale_color_manual(values = arm_colors) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom") +
  labs(
    title = "Spinal radial-glia motile score versus sampling depth",
    x = NULL,
    y = "mean gold-motile score",
    color = "arm",
    shape = "hash plate"
  )

save_plot_pair(
  "05_depth_diagnostics",
  depth_plot,
  width = 10,
  height = 5
)

depth_plot
```

## Step 8: Decompose the motile score into component genes

Zebrafish paralogs are summed to their human ortholog exactly as in the module
score. We then retain per-cell log-normalized expression and average it within
each embryo. The heatmap shows the 40 most highly expressed component genes;
selection is based on overall expression, not the foxj1a effect.

```{r component-genes, fig.width=12, fig.height=11}
human_modules <- readr::read_tsv(
  HUMAN_MODULE_FILE,
  show_col_types = FALSE
)
human_to_zebrafish <- readr::read_tsv(
  ORTHOLOGY_FILE,
  show_col_types = FALSE
)

motile_human_genes <- human_modules |>
  dplyr::filter(.data$module_id == FOCUS_MODULE) |>
  dplyr::distinct(.data$current_human_symbol)

ortholog_groups <- motile_human_genes |>
  dplyr::inner_join(human_to_zebrafish, by = "current_human_symbol") |>
  dplyr::filter(
    !is.na(.data$zebrafish_gene_id),
    .data$zebrafish_gene_id %in% rownames(focus_cds)
  ) |>
  dplyr::distinct(
    .data$zebrafish_gene_id,
    .data$current_human_symbol
  )

human_expression <- monocle3::aggregate_gene_expression(
  focus_cds,
  gene_group_df = ortholog_groups,
  norm_method = "size_only",
  gene_agg_fun = "sum",
  scale_agg_values = FALSE
)

if (inherits(human_expression, "sparseMatrix")) {
  human_expression@x <- log1p(human_expression@x)
} else {
  human_expression <- log1p(human_expression)
}

human_cell_expression <- Matrix::t(human_expression) |>
  as.matrix() |>
  as.data.frame() |>
  tibble::rownames_to_column("cell") |>
  tibble::as_tibble() |>
  dplyr::left_join(focus_cell_metadata, by = "cell") |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(rownames(human_expression)),
    names_to = "human_gene",
    values_to = "log_expression"
  )

embryo_gene_expression <- human_cell_expression |>
  dplyr::group_by(
    .data$embryo_ID,
    .data$arm,
    .data$hash_plate,
    .data$human_gene
  ) |>
  dplyr::summarise(
    mean_log_expression = mean(.data$log_expression),
    .groups = "drop"
  )

component_gene_summary <- embryo_gene_expression |>
  dplyr::group_by(.data$human_gene) |>
  dplyr::summarise(
    overall_mean_expression = mean(.data$mean_log_expression),
    control_mean = mean(
      .data$mean_log_expression[.data$arm == "control"]
    ),
    perturbation_mean = mean(
      .data$mean_log_expression[.data$arm == "perturbation"]
    ),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    expression_difference = .data$perturbation_mean - .data$control_mean
  ) |>
  dplyr::arrange(.data$expression_difference)

readr::write_tsv(
  embryo_gene_expression,
  file.path(DATA_DIR, "motile_component_genes_by_embryo.tsv")
)
readr::write_tsv(
  component_gene_summary,
  file.path(DATA_DIR, "motile_component_gene_summary.tsv")
)

genes_to_show <- component_gene_summary |>
  dplyr::slice_max(
    .data$overall_mean_expression,
    n = MAX_COMPONENT_GENES,
    with_ties = FALSE
  ) |>
  dplyr::pull("human_gene")

gene_order <- component_gene_summary |>
  dplyr::filter(.data$human_gene %in% genes_to_show) |>
  dplyr::arrange(.data$expression_difference) |>
  dplyr::pull("human_gene")

embryo_order <- embryo_depth |>
  dplyr::arrange(.data$arm, .data$mean_module_score) |>
  dplyr::pull("embryo_ID")

component_heatmap_data <- embryo_gene_expression |>
  dplyr::filter(.data$human_gene %in% genes_to_show) |>
  dplyr::group_by(.data$human_gene) |>
  dplyr::mutate(
    expression_sd = stats::sd(.data$mean_log_expression),
    scaled_expression = dplyr::if_else(
      .data$expression_sd > 0,
      (.data$mean_log_expression - mean(.data$mean_log_expression)) /
        .data$expression_sd,
      0
    )
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    human_gene = factor(.data$human_gene, levels = rev(gene_order)),
    embryo_ID = factor(.data$embryo_ID, levels = embryo_order)
  )

component_heatmap <- ggplot(
  component_heatmap_data,
  aes(
    x = .data$embryo_ID,
    y = .data$human_gene,
    fill = .data$scaled_expression
  )
) +
  geom_tile(color = "white", linewidth = 0.15) +
  facet_grid(cols = vars(.data$arm), scales = "free_x", space = "free_x") +
  scale_fill_gradient2(
    low = "#3B4CC0",
    mid = "white",
    high = "#B40426",
    midpoint = 0
  ) +
  theme_minimal(base_size = 8) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
    panel.grid = element_blank()
  ) +
  labs(
    title = "Gold motile-cilia component genes in spinal radial glia",
    subtitle = paste(
      "Top 40 genes by overall expression;",
      "each column is one embryo"
    ),
    fill = "within-gene\nstandardized\nexpression"
  )

save_plot_pair(
  "06_motile_component_gene_heatmap",
  component_heatmap,
  width = 12,
  height = 11
)

component_heatmap
```

## Component-gene direction summary

```{r component-direction-summary}
component_gene_summary |>
  dplyr::summarise(
    n_mapped_component_genes = dplyr::n(),
    n_lower_in_foxj1a = sum(.data$expression_difference < 0),
    fraction_lower_in_foxj1a = mean(.data$expression_difference < 0)
  )

component_gene_summary
```
