---
title: "GENE14 Phase-3 — Manual walkthrough"
author: "mdcolon"
date: "`r Sys.Date()`"
output: html_document
---

<!--
  Step-by-step, run-it-yourself walkthrough of the GENE14 Phase-3 analysis.
  We fill this in one section at a time. Reference: HOWTO_load_and_group.md
-->

# 1. Load libraries

```{r libs, message=FALSE, warning=FALSE}
# The processed CDS is a monocle3 object -> we need monocle3 to load it.
library(monocle3)  # load_monocle_objects(), colData()
library(readr)     # read_tsv() — fast TSV reader (for the label mapping)
library(dplyr)     # filter/mutate/group_by/count — the verbs
library(tidyr)     # pivot_wider / pivot_longer — reshaping
library(ggplot2)   # plots
```

# 2. Define paths

```{r paths}
# The processed cells (fitted hooke v3.1.0 CDS, ~7.3G, BPCells-backed):
MCCLINTOCK <- "/net/seahub_zfish/vol1/data/seahub_rna_processing/portal_inputs/v3.1.0/mcclintock/GENE14/run_1"
CDS_PATH   <- file.path(MCCLINTOCK, "filter_embryos", "embryo_filtered_cds")

# The phenotype label mapping (per-embryo table; the merge source):
BASE    <- "/net/trapnell/vol1/home/mdcolon/proj/morphseq/results/mcolon/20260715_gene14_hooke_dact"
PHENO   <- file.path(BASE, "output", "gene14_embryos_with_phenotype.tsv")
OUT_DIR <- file.path(BASE, "output")   # where fit results are saved

# The WT reference abundance atlas — the source of the REAL percent_max_abund
# (see the quality-filter chunk below for why this file is required).
WT_ABUNDANCES <- file.path(MCCLINTOCK, "kinetics", "wt_abundances.tsv")

# CONFIRM all three exist before going further:
stopifnot(file.exists(CDS_PATH), file.exists(PHENO), file.exists(WT_ABUNDANCES))
cat("CDS path, phenotype table, and WT abundances found.\n")
```

# 3. Load the processed cells

```{r load-cds}
# WARNING: this loads the 7.3G CDS — it OOM-kills a LOGIN node.= that de=esnt have enough RAM.
cds <- load_monocle_objects(CDS_PATH)
cat("cds:", ncol(cds), "cells x", nrow(cds), "genes\n")
```

# 4. Merge the phenotype labels into the object

```{r merge-labels}
# Read the label mapping (one row per embryo).
pheno <- read_tsv(PHENO, show_col_types = FALSE)

# The merge: for each CELL, look up its embryo_ID in the label table and pull
# back that embryo's morphseq_phenotype. match() returns the row index in
# `pheno` for each cell's embryo_ID; we index the label column with it.
# This is the whole join — one line, matched on embryo_ID.
colData(cds)$morphseq_phenotype <-
  pheno$morphseq_phenotype[match(colData(cds)$embryo_ID, pheno$embryo_ID)]

# CONFIRM the labels landed: how many cells now carry a phenotype?
# (blank/NA = pre-divergence embryos that have no phenotype label yet)
cat(sum(!is.na(colData(cds)$morphseq_phenotype)),
    "cells carry a MorphSeq phenotype\n")
table(colData(cds)$morphseq_phenotype, useNA = "ifany")
```

# 5. KR1 baseline — homozygous / crispant vs control (DACTs)

Before splitting by phenotype, get the simple, label-independent baseline: each
perturbation group vs ITS OWN control. Five fits, all the SAME recipe.

The steps below MUST run top-to-bottom — each depends on the columns the previous
one wrote onto the CDS. The order that matters:

    5a  load hooke/platt
    5b  ADD every colData column the fit needs  (genotype_block, batch_one)
        -> all onto `cds`, BEFORE building ccs
    5c  build ccs   (it snapshots the CDS colData AS-IS at this moment)
    5d  define the fit + contrast list
    5e  run the fits -> DACTs

Why the order is strict: `new_cell_count_set()` copies over ONLY the colData
columns present on the CDS at build time. If you build ccs first and add a column
after, ccs won't have it -> hooke's subset is empty -> "time range: Inf to -Inf"
+ a dgCMatrix crash. So ALL colData edits happen in 5b, and ccs is built in 5c.

## 5a. Load hooke / platt

```{r hooke-libs}
# hooke fits on the CELLS (needs the CDS) — it counts cells per embryo per cell
# type, then models how those abundances shift. platt wraps hooke's genotype fit.
library(monocle3)
library(hooke)
library(platt)
library(splines)  # ns() — the natural-spline fn hooke's timepoint formula calls
```

## 5b. Add ALL the colData columns the fit reads (onto the CDS, before ccs)

```{r prep-coldata}
# ---- (1) genotype_block: the fence, straight from rt_block --------------------
# The block IS who-was-collected-with-whom, so it groups each gene with its own
# controls (mut + negsib + AB reference share the gene's blocks).
#   Bl1 -> crispant (foxj1a/ift88/sspo/ctrl-inj + AB)   Bl2-5 -> cep290   Bl6-9 -> b9d2
colData(cds)$genotype_block <- dplyr::case_when(
  colData(cds)$rt_block == "Bl1"                              ~ "crispant",
  colData(cds)$rt_block %in% c("Bl2", "Bl3", "Bl4", "Bl5")    ~ "cep290",
  colData(cds)$rt_block %in% c("Bl6", "Bl7", "Bl8", "Bl9")    ~ "b9d2",
  TRUE                                                        ~ NA_character_
)

# ---- (2) batch_one: a CONSTANT batch term -------------------------------------
# fit_genotype_ccm REQUIRES a batch_col. rt_block can't be it — some blocks hold
# only case or only control (cep290 Bl2/Bl3 negsib-only, Bl5 case-only), so it's
# collinear with the case/control split (rank-deficient). A single-value column is
# a no-op batch term that satisfies the API without distorting the fit.
colData(cds)$batch_one <- "b1"

# NOTE: no pooled-control COLUMN is needed. The fence is done by SUBSETTING the ccs
# to one genotype_block before each fit (section 5d) — inside that subset, plain
# `perturbation` values (mut / negsib / reference) fit cleanly, and a POOLED control
# is just ctrl_ids = c("<gene>-negsib", "reference") with no leak possible.

# ---- CONFIRM the fence column landed before we freeze it into ccs -------------
stopifnot(!anyNA(colData(cds)$genotype_block))            # every cell got a family
print(table(colData(cds)$genotype_block, colData(cds)$perturbation, useNA = "ifany"))
```

## 5c. Build the cell_count_set (freezes the colData above)

```{r cell-count-set}
# ONE cell_count_set: how many cells of each cell_type each embryo has — the
# object hooke actually models. Built ONCE here, AFTER 5b, so it inherits
# genotype_block + batch_one. Reused (subset per block) by every contrast.
#   sample_group = embryo (the unit)   cell_group = cell_type (what we count)
ccs <- new_cell_count_set(cds, sample_group = "embryo_ID", cell_group = "cell_type")

# GUARD: prove the fit's columns rode along onto ccs (else the fit would silently
# subset to nothing -> "Inf to -Inf" crash). If this fails, re-run 5b then 5c.
# genotype_block is what we subset the ccs by (the fence); perturbation holds the
# case + control values within each block.
stopifnot(all(c("perturbation", "genotype_block", "timepoint", "batch_one")
              %in% colnames(colData(ccs))))
ccs
```

## 5d. The unified fit function (subset ccs by block = the fence)

```{r fit-within-block}
# THE reusable fit for the whole notebook (baseline AND the phenotype splits later).
# It SUBSETS the ccs to ONE genotype_block first — that subset IS the fence: inside
# it only that gene's cells remain (mut / negsib / its own AB), so plain
# `perturbation` values fit cleanly and `reference` can never leak from other blocks.
# A POOLED control is simply ctrl_ids = c("<gene>-negsib", "reference").
#
# The two knobs per contrast: `genotype` (case) and `ctrl_ids` (control, a vector).
fit_within_block <- function(block, genotype, ctrl_ids, tag) {
  ccs_sub <- ccs[, colData(ccs)$genotype_block == block]   # <-- SUBSET = the fence
  message("\n=== ", tag, "  (", genotype, " vs ", paste(ctrl_ids, collapse = "+"),
          " | block=", block, ", ", ncol(ccs_sub), " embryos) ===")
  ccm <- platt:::fit_genotype_ccm(
    genotype         = genotype,
    ccs              = ccs_sub,
    perturbation_col = "perturbation",  # plain perturbation — the subset fenced it
    ctrl_ids         = ctrl_ids,        # one OR many control values (pool = vector)
    interval_col     = "timepoint",     # drives the timepoint spline
    batch_col        = "batch_one",     # constant no-op batch (see 5b)
    vhat_method      = "sandwich_var",  # fast analytical variance (not bootstrap)
    num_threads      = 3
  )
  # SAVE THE FITTED MODEL (for plotting: plot_contrast / estimate_abundances). LARGE
  # (~6GB — bundles the ccs), one .rds per contrast, written as the fit finishes.
  saveRDS(ccm, file.path(OUT_DIR, paste0("dact_ccm_", tag, ".rds")))
  tbl <- platt:::get_perturbation_effects(ccm, interval_col = "timepoint")
  tbl$tag <- tag
  # SAVE-AS-YOU-GO: write each contrast the moment it finishes, so a LATER crash
  # never loses the ones that already worked.
  readr::write_tsv(tbl, file.path(OUT_DIR, paste0("dact_contrast_", tag, ".tsv")))
  tbl
}
```

## 5e. The full contrast set (9 fits) + run

```{r contrast-spec}
# ALL contrasts fit + saved in ONE run. Re-running the chunk refits + re-saves all.
# cep290 & b9d2 get 3 CONTROLS each (negsib / reference / the pool of both) so we can
# see if the matched-sibling control is BLUNTING the signal (b9d2=1, cep290=9 earlier).
# ctrl_ids is a LIST column (each entry a character vector) — pool = c(negsib, ref).
contrasts_all <- tibble::tribble(
  ~block,    ~tag,                  ~genotype,     ~ctrl,
  # -- cep290: 3 controls ------------------------------------------------------
  "cep290",  "cep290_vs_negsib",    "cep290-mut",  "cep290-negsib",
  "cep290",  "cep290_vs_reference", "cep290-mut",  "reference",
  "cep290",  "cep290_vs_pool",      "cep290-mut",  list(c("cep290-negsib", "reference")),
  # -- b9d2: 3 controls --------------------------------------------------------
  "b9d2",    "b9d2_vs_negsib",      "b9d2-mut",    "b9d2-negsib",
  "b9d2",    "b9d2_vs_reference",   "b9d2-mut",    "reference",
  "b9d2",    "b9d2_vs_pool",        "b9d2-mut",    list(c("b9d2-negsib", "reference")),
  # -- IS THE SIBLING CONTROL CLEAN? negsib vs AB reference --------------------
  # If these yield DACTs, the negsibs are THEMSELVES shifted from wildtype — which
  # is exactly why mut_vs_negsib is blunted vs mut_vs_reference.
  "cep290",  "cep290_negsib_vs_ref","cep290-negsib","reference",
  "b9d2",    "b9d2_negsib_vs_ref",  "b9d2-negsib", "reference",
  # -- crispants vs injection control (their designed control) -----------------
  "crispant","foxj1a_vs_ctrl",      "foxj1a",      "ctrl-inj",
  "crispant","ift88_vs_ctrl",       "ift88",       "ctrl-inj",
  "crispant","sspo_vs_ctrl",        "sspo",        "ctrl-inj",
)
# normalize ctrl to a list-column (so scalar strings and the pooled vector coexist)
contrasts_all$ctrl <- lapply(contrasts_all$ctrl, unlist)
contrasts_all %>% select(block, tag, genotype)
```

```{r run-all-fits}
# Run all 9. Each returns a per-cell-type x per-timepoint contrast table; stacked.
dact_raw <- purrr::pmap_dfr(
  contrasts_all,
  function(block, tag, genotype, ctrl) fit_within_block(block, genotype, ctrl, tag)
)
dplyr::count(dact_raw, tag)
```

## 5e-bis. Attach the REAL percent_max_abund (the WT-atlas presence gate)

`percent_max_abund` is **not** a contrast quantity and cannot be derived from
`get_perturbation_effects()` output. It is a property of the WT reference atlas:

```
percent_max_abund = exp(log_abund) / max_abundance
max_abundance     = that cell type's PEAK abundance over developmental time in WT
```

at grain `(cell_group x timepoint)`. It answers **"does this cell type meaningfully
exist at this age?"** — nothing to do with mutant vs control. (Verified directly
against `wt_abundances.tsv`; mcclintock's `fit_wt_model.R:240` uses the same 0.1
cutoff for its `present_above_thresh` flag.)

This matters: in the mcclintock-output path (`scripts/2_baseline_perturbation_dacts.py`)
this gate removed ~89 spurious negsib DACTs that were cell types with **zero cells in
every embryo** at the timepoint — spline undershoot in data-sparse regions. Our
hand-rolled fits never had the column, so that whole class of artifact was never
removed here.

```{r attach-pct-max-abund}
# Join grain = (cell_group, timepoint). Verified: the atlas covers 3-96 hpf as
# integers, and all 353 cell types in our contrast tables are present in it.
wt_abund <- readr::read_tsv(WT_ABUNDANCES, show_col_types = FALSE) %>%
  dplyr::select(cell_group, timepoint, percent_max_abund)

attach_pct_max_abund <- function(tbl) {
  # mcclintock's own contrast_abundance_tbl.tsv already carries the column.
  if ("percent_max_abund" %in% colnames(tbl)) return(tbl)
  stopifnot(all(c("cell_group", "timepoint") %in% colnames(tbl)))
  n_before <- nrow(tbl)
  out <- dplyr::left_join(tbl, wt_abund, by = c("cell_group", "timepoint"))
  # A silent key mismatch would leave the column all-NA and turn the gate into a
  # no-op — which is exactly the failure mode being fixed here. Fail loudly.
  stopifnot(nrow(out) == n_before)
  n_na <- sum(is.na(out$percent_max_abund))
  if (n_na > 0) {
    stop(sprintf(
      "%d/%d rows got no percent_max_abund - check cell_group naming / timepoint coding",
      n_na, n_before))
  }
  out
}

dact_raw <- attach_pct_max_abund(dact_raw)
cat("percent_max_abund attached to dact_raw:", nrow(dact_raw), "rows\n")
```

## 5f. Inspect the thresholds BEFORE applying them

```{r threshold-diagnostics, fig.width=9, fig.height=6}
Q_THRESH <- 0.05   # delta_q_value cutoff (FDR)
LA_THRESH <- -3    # log_abund_x floor (drop near-empty cell types)
PMA_THRESH <- 0.1  # percent_max_abund floor (cell type actually present at this age)

# --- how many rows each cut removes, and the two cuts combined ------------------
n_all   <- nrow(dact_raw)
n_q     <- sum(dact_raw$delta_q_value < Q_THRESH, na.rm = TRUE)
n_la    <- sum(dact_raw$log_abund_x  >= LA_THRESH, na.rm = TRUE)
n_both  <- sum(dact_raw$delta_q_value < Q_THRESH &
               dact_raw$log_abund_x  >= LA_THRESH, na.rm = TRUE)
cat(sprintf("total rows            : %d\n", n_all))
cat(sprintf("pass q < %.2f          : %d  (%.1f%%)\n", Q_THRESH, n_q,  100*n_q /n_all))
cat(sprintf("pass log_abund_x >= %d : %d  (%.1f%%)\n", LA_THRESH, n_la, 100*n_la/n_all))
cat(sprintf("pass BOTH (= DACTs)   : %d  (%.1f%%)\n", n_both, 100*n_both/n_all))

# --- histograms with the threshold lines drawn ---------------------------------
p1 <- ggplot(dact_raw, aes(delta_q_value)) +
  geom_histogram(bins = 50, fill = "grey70") +
  geom_vline(xintercept = Q_THRESH, colour = "red") +
  labs(title = "delta_q_value (red = 0.05 cutoff)", x = "delta_q_value", y = "rows")

p2 <- ggplot(dact_raw, aes(log_abund_x)) +
  geom_histogram(bins = 50, fill = "grey70") +
  geom_vline(xintercept = LA_THRESH, colour = "red") +
  labs(title = "log_abund_x (red = -3 floor; left of line dropped)",
       x = "log_abund_x", y = "rows")

# --- the key 2D view: WHERE does each cut land, jointly? -----------------------
# color by pass/fail so you see if the log_abund floor is removing significant hits
dact_raw$pass <- with(dact_raw, delta_q_value < Q_THRESH & log_abund_x >= LA_THRESH)
p3 <- ggplot(dact_raw, aes(log_abund_x, -log10(delta_q_value), colour = pass)) +
  geom_point(alpha = 0.3, size = 0.6) +
  geom_vline(xintercept = LA_THRESH, colour = "red") +
  geom_hline(yintercept = -log10(Q_THRESH), colour = "red") +
  scale_colour_manual(values = c("FALSE" = "grey70", "TRUE" = "firebrick")) +
  labs(title = "joint cut: red box (upper-right) = kept DACTs",
       x = "log_abund_x", y = "-log10(delta_q_value)")

# stack them (patchwork if available, else print individually)
if (requireNamespace("patchwork", quietly = TRUE)) {
  print(patchwork::wrap_plots(p1, p2, p3, ncol = 2))
} else { print(p1); print(p2); print(p3) }
```

The key question p3 answers: **is the `log_abund_x >= -3` floor throwing away
SIGNIFICANT hits** (grey points that are high on the y-axis but left of the red
vertical line)? If yes, loosen the floor; if those are all low-significance, the
floor is safely trimming noise.

## 5g. Apply the DACT gate

```{r kr1-dact-gate}
# DACT = significant AND above BOTH abundance-quality floors.
#   delta_q_value      < 0.05   (FDR)
#   log_abund_x       >= -3     (not a near-empty cell type in the control)
#   percent_max_abund >= 0.1    (cell type actually EXISTS at this age; joined
#                                from the WT atlas in 5e-bis — get_perturbation_effects
#                                does not emit it)
dact_hits <- dact_raw %>%
  filter(delta_q_value < Q_THRESH,
         log_abund_x >= LA_THRESH,
         percent_max_abund >= PMA_THRESH)

# What the presence gate alone removed — if this is large, those calls were
# cell types that barely exist at that timepoint (the spline-undershoot class).
n_pre <- dact_raw %>%
  filter(delta_q_value < Q_THRESH, log_abund_x >= LA_THRESH) %>%
  nrow()
cat(sprintf("hits before percent_max_abund gate: %d\n after: %d  (removed %d)\n",
            n_pre, nrow(dact_hits), n_pre - nrow(dact_hits)))

# the KR1 answer: DACT cell types per contrast
dact_hits %>%
  count(tag, name = "n_dacts") %>%
  arrange(desc(n_dacts))
```

## 5h. VERIFY the calls — per-embryo proportion box plots

The DACT gate is a model output. Before trusting it, plot the RAW per-embryo
proportion of each hit cell type, case vs control, and eyeball whether the mutant
really is depleted/enriched. A real DACT = clear separation with each embryo as a
dot; a suspicious one = overlapping clouds driven by one outlier.

```{r dact-proportions-helper}
# Per-embryo proportion of every cell_type, built straight from the CDS colData
# (embryo_ID x cell_type) — transparent, no ccs internals. One row per embryo x type.
embryo_props <- as.data.frame(colData(cds)) %>%
  count(embryo_ID, perturbation, genotype_block, timepoint, cell_type,
        name = "n_cells") %>%
  group_by(embryo_ID) %>%
  mutate(prop = n_cells / sum(n_cells)) %>%   # fraction of THIS embryo's cells
  ungroup()
```

```{r verify-dacts, fig.width=11, fig.height=8}
# Pick ONE contrast to verify (change tag to check others). cep290's 9 is a good
# small start; crispants have 50+ so you'd want to page through them.
VERIFY_TAG <- "cep290_vs_reference"   # <- edit to any tag in dact_hits

# which (block, case, controls) does this tag correspond to? pull from contrasts_all
spec <- contrasts_all %>% filter(tag == VERIFY_TAG)
blk  <- spec$block[[1]]; case <- spec$genotype[[1]]; ctrls <- spec$ctrl[[1]]

# the DACT cell types for this contrast, tagged by direction (loss vs gain)
hits <- dact_hits %>%
  filter(tag == VERIFY_TAG) %>%
  transmute(cell_type = cell_group,
            dir = ifelse(delta_log_abund < 0, "loss (down in mut)", "gain (up in mut)"),
            delta_log_abund, delta_q_value)

# embryos in this contrast: the case + its control(s), fenced to the block
plot_df <- embryo_props %>%
  filter(genotype_block == blk,
         perturbation %in% c(case, ctrls),
         cell_type %in% hits$cell_type) %>%
  mutate(group = ifelse(perturbation == case, "mutant", "control")) %>%
  left_join(hits %>% select(cell_type, dir), by = "cell_type")

# box + every-embryo jitter, faceted by cell type, colored by DACT direction
ggplot(plot_df, aes(group, prop, colour = dir)) +
  geom_boxplot(outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.7) +
  facet_wrap(~ cell_type, scales = "free_y", labeller = label_wrap_gen(28)) +
  scale_colour_manual(values = c("loss (down in mut)" = "#B2182B",
                                 "gain (up in mut)"  = "#2166AC")) +
  labs(title = paste0("Verify DACTs: ", VERIFY_TAG),
       subtitle = "each dot = one embryo's proportion of that cell type",
       x = NULL, y = "proportion of embryo's cells") +
  theme(strip.text = element_text(size = 7),
        legend.position = "top")
```

Read it: for a REAL call the `mutant` box should sit clearly below (loss) or above
(gain) the `control` box, with the embryo dots supporting it. If the boxes overlap
heavily or one dot drags the mean, treat that DACT with suspicion.

## 5f. Save the results (same outputs the grid script writes)

```{r save-kr1}
# Persist so you can reload WITHOUT re-loading the CDS or re-fitting.
#   dact_raw_contrasts.rds  = the combined raw table, exact R object  (reload: readRDS)
#   dact_raw_contrasts.tsv  = the same, as a plain text table (view / share)
#   dact_hits.tsv          = the DACT-gated subset (the answer)
# (Per-contrast dact_contrast_<tag>.tsv + dact_ccm_<tag>.rds written during 5e.)
saveRDS(dact_raw,            file.path(OUT_DIR, "dact_raw_contrasts.rds"))
readr::write_tsv(dact_raw,   file.path(OUT_DIR, "dact_raw_contrasts.tsv"))
readr::write_tsv(dact_hits,  file.path(OUT_DIR, "dact_hits.tsv"))

cat("saved to", OUT_DIR, ":\n",
    " dact_raw_contrasts.rds / .tsv  (all 9 contrasts, raw)\n",
    " dact_hits.tsv                  (DACT-gated)\n",
    " dact_contrast_<tag>.tsv        (per-contrast, written during 5e)\n")
```

To reload just the result tables in a fresh session (no CDS needed):

```{r reload-kr1, eval=FALSE}
dact_raw   <- readRDS(file.path(OUT_DIR, "dact_raw_contrasts.rds"))
dact_hits <- readr::read_tsv(file.path(OUT_DIR, "dact_hits.tsv"))
```

## 5g. Checkpoint — save enough to pick up exactly where you left off

```{r checkpoint}
# A CHECKPOINT bundles several named objects into ONE .RData file, so a fresh
# session can restore the whole working state at once (by name) — no CDS reload,
# no re-fitting.
#   dact_raw, dact_hits  = the results
#   contrasts_all        = the contrast spec (cheap, handy)
#   ccs                  = the cell_count_set — so you can run MORE fits (the
#                          phenotype splits) later WITHOUT rebuilding it from the
#                          7.3G CDS. (You can always rebuild it if you drop this.)
# NOT saved: `cds` — 7.3G / BPCells-backed; reload from CDS_PATH only if you truly
# need cell-level data again.
CKPT <- file.path(OUT_DIR, "dact_checkpoint.RData")
save(dact_raw, dact_hits, contrasts_all, ccs, file = CKPT)
cat("checkpoint ->", CKPT, "\n")
```

Resume in a brand-new R session (restores all objects by their names):

```{r resume, eval=FALSE}
# after library(monocle3); library(hooke); library(platt); library(splines)
load(file.path(OUT_DIR, "dact_checkpoint.RData"))
#   -> dact_raw, dact_hits, contrasts_all, ccs are all back
```

# 6. 48hpf-ONLY drill-in — cep290 & b9d2 with the timepoint spline KILLED

WHY: the KR1 fits (section 5) use `interval_col = "timepoint"`, which fits a natural
spline `ns(timepoint, ...)` over the whole time course. A wiggly spline across sparse
timepoints can MANUFACTURE abundance shifts that read as DACTs but are fit artifacts,
not biology. This section isolates ONE timepoint (48hpf) for cep290 and b9d2 so the
spline has nothing to bend over — any DACT that survives here is a genuine 48hpf
abundance difference.

HOW the spline actually dies (no formula hacking needed): inside
`platt:::fit_genotype_ccm`, the time term is chosen by
`num_knockout_timepoints = length(unique(colData(subset_ccs)[[interval_col]]))`.
When that is > 1 it builds `~ ns(timepoint, knots=...)`; when it is **1** it falls to
the `else` branch and the nuisance formula collapses to `~ 1` — no spline, no time
term. So **subsetting the ccs to a single timepoint removes the spline by the
library's own logic.** We just feed it 48hpf-only cells.

FILTER GRAIN (the thing to keep straight): the DACT threshold in 5g is applied
per **(cell_type × timepoint × contrast)** row of `get_perturbation_effects`, NOT per
embryo. Embryos are collapsed into the cell_count_set at 5c and pooled by the model;
the gate never sees an individual embryo. Killing the timepoint axis here means each
surviving DACT is one (cell_type × contrast) abundance call at 48hpf.

## 6a. Confirm the timepoint coding, then subset the ccs to 48hpf

```{r tp48-subset}
# What timepoint values exist? (confirm 48 is present and how it's coded — numeric 48
# vs "48hpf" string. Adjust TP48 to match.)
print(sort(unique(colData(ccs)$timepoint)))

TP48 <- 48   # <- edit if timepoint is coded as a string like "48" / "48hpf"

# Subset the SHARED ccs to 48hpf cells only. Everything downstream fences per block
# exactly as fit_within_block does — this just narrows the time axis to one value.
ccs_48 <- ccs[, !is.na(colData(ccs)$timepoint) & colData(ccs)$timepoint == TP48]

cat(sprintf("ccs_48: %d embryos at timepoint==%s (of %d total)\n",
            ncol(ccs_48), TP48, ncol(ccs)))
# GUARD: each block we refit must actually have both case + control embryos at 48hpf.
print(table(colData(ccs_48)$genotype_block, colData(ccs_48)$perturbation,
            useNA = "ifany"))
```

## 6b. Refit cep290 & b9d2 at 48hpf only (spline-free)

```{r tp48-fit}
# Same fence-by-subset logic as fit_within_block (5d), but on ccs_48. Because ccs_48
# holds a single timepoint, fit_genotype_ccm's time term collapses to ~ 1 (no spline).
# We refit ONLY the cep290 & b9d2 contrasts (the two you care about), all 3 controls each.
fit_within_block_48 <- function(block, genotype, ctrl_ids, tag) {
  ccs_sub <- ccs_48[, colData(ccs_48)$genotype_block == block]
  stopifnot(ncol(ccs_sub) > 0)
  message("\n=== [48hpf] ", tag, "  (", genotype, " vs ",
          paste(ctrl_ids, collapse = "+"), " | ", ncol(ccs_sub), " embryos) ===")
  ccm <- platt:::fit_genotype_ccm(
    genotype         = genotype,
    ccs              = ccs_sub,
    perturbation_col = "perturbation",
    ctrl_ids         = ctrl_ids,
    interval_col     = "timepoint",     # still named, but single-value => ~1 (no spline)
    batch_col        = "batch_one",
    vhat_method      = "sandwich_var",
    num_threads      = 3
  )
  saveRDS(ccm, file.path(OUT_DIR, paste0("dact48_ccm_", tag, ".rds")))
  tbl <- platt:::get_perturbation_effects(ccm, interval_col = "timepoint")
  tbl$tag <- tag
  readr::write_tsv(tbl, file.path(OUT_DIR, paste0("dact48_contrast_", tag, ".tsv")))
  tbl
}

# Just the cep290 + b9d2 rows from the existing spec (drops crispant + negsib-vs-ref).
contrasts_48 <- contrasts_all %>%
  dplyr::filter(block %in% c("cep290", "b9d2"),
                tag %in% c("cep290_vs_negsib", "cep290_vs_reference", "cep290_vs_pool",
                           "b9d2_vs_negsib",   "b9d2_vs_reference",   "b9d2_vs_pool"))

dact48_raw <- purrr::pmap_dfr(
  contrasts_48,
  function(block, tag, genotype, ctrl) fit_within_block_48(block, genotype, ctrl, tag)
)
dplyr::count(dact48_raw, tag)
```

## 6c. Gate + color by % of max abundance

```{r tp48-gate-pctmax, fig.width=10, fig.height=7}
# Same DACT gate as 5g (per cell_type × contrast — there's only ONE timepoint now).
Q_THRESH   <- 0.05
LA_THRESH  <- -3
PMA_THRESH <- 0.1

# % of max abundance = the REAL percent_max_abund, joined from the WT atlas in
# 5e-bis. It is each cell type's abundance as a fraction of ITS OWN developmental
# peak in wild-type — i.e. "how present is this cell type at 48hpf?".
#
# This is the gate that matters most for THIS section. The spline artifacts we are
# hunting are largely cell types that barely exist at the timepoint; without the
# presence gate you cannot tell "the spline manufactured this" apart from "this
# cell type isn't really there at 48hpf" — both show up as spline-only SUSPECTs.
dact48_hits <- dact48_raw %>%
  attach_pct_max_abund() %>%
  dplyr::filter(delta_q_value < Q_THRESH,
                log_abund_x >= LA_THRESH,
                percent_max_abund >= PMA_THRESH) %>%
  dplyr::mutate(pct_of_max = 100 * percent_max_abund)

cat(sprintf(
  "48hpf DACTs (q<%.2f, log_abund_x>=%d, pct_max_abund>=%.2f): %d rows / %d contrasts\n",
  Q_THRESH, LA_THRESH, PMA_THRESH, nrow(dact48_hits),
  dplyr::n_distinct(dact48_hits$tag)))

# One panel per contrast; each surviving cell type as a point, x = effect size
# (delta_log_abund), colored by its % of max abundance (how "real"/dominant it is).
ggplot(dact48_hits,
       aes(x = delta_log_abund,
           y = reorder(cell_group, delta_log_abund),
           colour = pct_of_max)) +
  geom_vline(xintercept = 0, colour = "grey60", linewidth = 0.3) +
  geom_point(size = 3) +
  facet_wrap(~ tag, scales = "free_y") +
  scale_colour_viridis_c(name = "% of WT peak\nabundance", limits = c(0, 100)) +
  labs(title = "48hpf-only DACTs (spline killed) — cep290 & b9d2",
       subtitle = "x = delta_log_abund (mut vs control); colour = abundance as % of that cell type's own WT developmental peak",
       x = "delta_log_abund  (<0 loss in mutant, >0 gain)", y = NULL) +
  theme(strip.text = element_text(size = 8))
```

## 6d. Does killing the spline change the calls? (48hpf vs full-spline)

```{r tp48-vs-full}
# The payoff comparison: which DACTs were present in the FULL spline fit (any timepoint)
# but VANISH when restricted to a clean 48hpf-only fit — candidates for spline artifacts.
full_48ish <- dact_hits %>%          # from section 5 (spline fits, all timepoints)
  dplyr::filter(tag %in% contrasts_48$tag) %>%
  dplyr::distinct(tag, cell_group) %>%
  dplyr::mutate(in_full_spline = TRUE)

only48 <- dact48_hits %>%
  dplyr::distinct(tag, cell_group) %>%
  dplyr::mutate(in_48hpf_only = TRUE)

compare <- dplyr::full_join(full_48ish, only48, by = c("tag", "cell_group")) %>%
  tidyr::replace_na(list(in_full_spline = FALSE, in_48hpf_only = FALSE)) %>%
  dplyr::mutate(status = dplyr::case_when(
    in_full_spline &  in_48hpf_only ~ "robust (both)",
    in_full_spline & !in_48hpf_only ~ "spline-only (SUSPECT)",
    !in_full_spline & in_48hpf_only ~ "48hpf-only (new)")) %>%
  dplyr::arrange(tag, status)

print(compare, n = Inf)
dplyr::count(compare, tag, status)
```

Read 6d: **`spline-only (SUSPECT)`** rows are DACTs the full-timecourse spline called
that DO NOT survive a clean 48hpf-only refit — your prime candidates for spline-
manufactured abnormalities. **`robust (both)`** survive the honest single-timepoint
test. `48hpf-only (new)` appear only when the spline's borrowing-across-time was
*hiding* a real 48hpf effect.
