---
title: "Foxj1a response across five cilia modules at 48 hpf"
output: html_document
---

This analysis compares the five approved cilia modules in the same foxj1a
versus control-injected embryos. The embryo remains the biological replicate.

```{r setup, message=FALSE, warning=FALSE}
library(igraph)
library(platt)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(ggplot2)
library(ggrepel)
library(stringr)
library(patchwork)

knitr::opts_chunk$set(dev = "ragg_png", fig.bg = "white")

PROJECT_DIR <- paste0(
  "/net/trapnell/vol1/home/mdcolon/proj/morphseq/results/",
  "mcolon/20260727_gene14_clean"
)
MODULE_DIR <- file.path(PROJECT_DIR, "cilia_gene_module")
PARENT_DIR <- file.path(MODULE_DIR, "foxj1a_48hpf_inspection")
HERE <- file.path(PARENT_DIR, "module_comparison_48hpf")
DATA_DIR <- file.path(HERE, "data")
FIGURE_DIR <- file.path(HERE, "figures")

MODEL_RESULT_FILE <- file.path(
  MODULE_DIR,
  "output",
  "differential_module_results",
  "foxj1a_48hpf.tsv"
)
EMBRYO_SCORE_FILE <- file.path(
  MODULE_DIR,
  "output",
  "embryo_celltype_cilia_module_scores.tsv"
)
FOXJ1A_BASELINE_FILE <- file.path(
  PARENT_DIR,
  "data",
  "celltype_baseline_and_motile_response.tsv"
)
CELLTYPE_LINEAGE_FILE <- file.path(
  PROJECT_DIR,
  "0_shared",
  "cell_type_lineage.tsv"
)
STATE_GRAPH_FILE <- paste0(
  "/net/seahub_zfish/vol1/data/graphs/",
  "v3.1.0/combined_graph.rds"
)

MODULES <- c(
  "v1_all",
  "gold_motile_cilia",
  "v1_basal_body",
  "v1_ciliary_tip",
  "v1_transition_zone"
)
MODULE_LABELS <- c(
  v1_all = "Broad cilia",
  gold_motile_cilia = "Motile cilia",
  v1_basal_body = "Basal body",
  v1_ciliary_tip = "Ciliary tip",
  v1_transition_zone = "Transition zone"
)

PERTURBATION <- "foxj1a"
CONTROL <- "ctrl-inj"
TIMEPOINT <- 48
MINIMUM_CELLS <- 10L
STANDARDIZED_EFFECT_LIMIT <- 2

dir.create(DATA_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(FIGURE_DIR, recursive = TRUE, showWarnings = FALSE)

save_plot_pair <- function(file_stem, plot, width, height) {
  ggsave(
    file.path(FIGURE_DIR, paste0(file_stem, ".pdf")),
    plot,
    width = width,
    height = height,
    limitsize = FALSE
  )
  ggsave(
    file.path(FIGURE_DIR, paste0(file_stem, ".png")),
    plot,
    device = ragg::agg_png,
    width = width,
    height = height,
    units = "in",
    dpi = 200,
    bg = "white",
    limitsize = FALSE
  )
}

arm_colors <- c(control = "#4C78A8", perturbation = "#E45756")
```

## Step 1: Read and summarize the five module results

The original q-value corrects across all module and cell-type tests in this
contrast. `q_within_module` is also reported as a descriptive, module-specific
correction; it does not replace the primary correction.

```{r read-results}
module_results <- readr::read_tsv(
  MODEL_RESULT_FILE,
  show_col_types = FALSE
) |>
  dplyr::filter(.data$module_id %in% MODULES) |>
  dplyr::group_by(.data$module_id) |>
  dplyr::mutate(
    q_within_module = p.adjust(.data$p_value, method = "BH")
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    module_id = factor(.data$module_id, levels = MODULES),
    module_label = factor(
      MODULE_LABELS[as.character(.data$module_id)],
      levels = unname(MODULE_LABELS[MODULES])
    )
  )

module_summary <- module_results |>
  dplyr::group_by(.data$module_id, .data$module_label) |>
  dplyr::summarise(
    n_cell_types = dplyr::n(),
    n_decreased = sum(.data$score_difference < 0),
    n_increased = sum(.data$score_difference > 0),
    n_fdr = sum(.data$is_fdr_significant),
    smallest_q = min(.data$q_value),
    minimum_difference = min(.data$score_difference),
    maximum_difference = max(.data$score_difference),
    .groups = "drop"
  )

readr::write_tsv(
  module_results,
  file.path(DATA_DIR, "all_module_celltype_results.tsv")
)
readr::write_tsv(
  module_summary,
  file.path(DATA_DIR, "module_summary.tsv")
)

module_summary
```

## Step 2: Make one shared global graph layout

The graph layout is calculated once. Untested atlas nodes remain visible in
gray so every module is shown in the same developmental context.

```{r graph-layout}
celltype_lineage <- readr::read_tsv(
  CELLTYPE_LINEAGE_FILE,
  show_col_types = FALSE
)
state_graph <- readRDS(STATE_GRAPH_FILE)

node_metadata <- tibble::tibble(id = igraph::V(state_graph)$name) |>
  dplyr::left_join(
    dplyr::select(celltype_lineage, cell_type, sulston_group),
    by = c("id" = "cell_type")
  ) |>
  dplyr::mutate(
    group_nodes_by = dplyr::coalesce(.data$sulston_group, "Unknown")
  )

graph_layout <- platt:::layout_state_graph(
  G = state_graph,
  node_metadata = node_metadata
)

graph_nodes <- as.data.frame(graph_layout$gvizl_coords)
names(graph_nodes) <- c("x", "y")
graph_nodes <- graph_nodes |>
  tibble::rownames_to_column("cell_type")
graph_edges <- graph_layout$bezier_df

graph_module_nodes <- tidyr::crossing(
  module_id = factor(MODULES, levels = MODULES),
  graph_nodes
) |>
  dplyr::left_join(
    module_results,
    by = c("module_id", "cell_type")
  ) |>
  dplyr::mutate(
    module_label = factor(
      MODULE_LABELS[as.character(.data$module_id)],
      levels = unname(MODULE_LABELS[MODULES])
    )
  )

readr::write_tsv(
  graph_module_nodes,
  file.path(DATA_DIR, "global_graph_module_node_metrics.tsv")
)
```

## Step 3: Raw score differences on the global graph

Each module gets its own raw score scale because raw differences cannot be
compared numerically between gene sets. Every scale is symmetrically capped at
that module's 95th percentile absolute difference so one outlier cannot flatten
the rest of its graph.

```{r raw-difference-graphs, fig.width=16, fig.height=15}
draw_raw_difference_graph <- function(module_name) {
  nodes <- graph_module_nodes |>
    dplyr::filter(.data$module_id == module_name)

  color_limit <- stats::quantile(
    abs(nodes$score_difference),
    0.95,
    na.rm = TRUE
  )

  labels <- nodes |>
    dplyr::filter(!is.na(.data$score_difference)) |>
    dplyr::arrange(.data$q_value, dplyr::desc(abs(.data$score_difference))) |>
    dplyr::slice_head(n = 4)

  ggplot() +
    geom_path(
      data = graph_edges,
      aes(x = .data$x, y = .data$y, group = .data$edge_name),
      color = "grey84",
      linewidth = 0.18
    ) +
    geom_point(
      data = nodes,
      aes(x = .data$x, y = .data$y, fill = .data$score_difference),
      shape = 21,
      size = 3,
      stroke = 0.25,
      color = "grey35"
    ) +
    ggrepel::geom_text_repel(
      data = labels,
      aes(x = .data$x, y = .data$y, label = .data$cell_type),
      size = 2.5,
      max.overlaps = Inf,
      min.segment.length = 0,
      box.padding = 0.5,
      seed = 1
    ) +
    scale_fill_gradient2(
      low = "#2166AC",
      mid = "white",
      high = "#B2182B",
      midpoint = 0,
      limits = c(-color_limit, color_limit),
      oob = scales::squish,
      breaks = scales::breaks_pretty(n = 4),
      labels = scales::label_number(accuracy = 0.001),
      na.value = "grey90",
      name = "score change",
      guide = guide_colorbar(
        title.position = "top",
        barwidth = grid::unit(2.8, "in"),
        barheight = grid::unit(0.15, "in")
      )
    ) +
    theme_void(base_size = 10) +
    theme(
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(size = 9),
      legend.position = "bottom",
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8)
    ) +
    labs(
      title = unname(MODULE_LABELS[module_name]),
      subtitle = paste0(
        "foxj1a - control; color capped at |",
        format(color_limit, digits = 2, scientific = TRUE),
        "|"
      )
    )
}

raw_graphs <- lapply(MODULES, draw_raw_difference_graph)
raw_difference_figure <- patchwork::wrap_plots(raw_graphs, ncol = 2) +
  patchwork::plot_annotation(
    title = "Raw cilia-module score changes across the cell-type graph",
    subtitle = "Blue = lower after foxj1a perturbation; red = higher; each module has its own robust scale"
  )

save_plot_pair(
  "01_raw_difference_global_graphs",
  raw_difference_figure,
  width = 16,
  height = 15
)

raw_difference_figure
```

## Step 4: Standardized effects on one common scale

This is the directly comparable view. A standardized effect of -1 means the
perturbation mean is one pooled embryo-level standard deviation below control.

```{r standardized-graph, fig.width=16, fig.height=10}
standardized_edges <- tidyr::crossing(
  module_label = factor(
    unname(MODULE_LABELS[MODULES]),
    levels = unname(MODULE_LABELS[MODULES])
  ),
  graph_edges
)

standardized_graph <- ggplot() +
  geom_path(
    data = standardized_edges,
    aes(
      x = .data$x,
      y = .data$y,
      group = interaction(.data$module_label, .data$edge_name)
    ),
    color = "grey85",
    linewidth = 0.16
  ) +
  geom_point(
    data = graph_module_nodes,
    aes(
      x = .data$x,
      y = .data$y,
      fill = .data$standardized_effect,
      color = .data$is_fdr_significant
    ),
    shape = 21,
    size = 2.5,
    stroke = 0.45
  ) +
  facet_wrap(vars(.data$module_label), ncol = 2) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    limits = c(-STANDARDIZED_EFFECT_LIMIT, STANDARDIZED_EFFECT_LIMIT),
    oob = scales::squish,
    na.value = "grey90",
    name = "standardized effect",
    guide = guide_colorbar(
      title.position = "top",
      barwidth = grid::unit(3.5, "in")
    )
  ) +
  scale_color_manual(
    values = c(`TRUE` = "black", `FALSE` = "grey45"),
    na.value = "grey75",
    guide = "none"
  ) +
  theme_void(base_size = 10) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    legend.position = "bottom"
  ) +
  labs(
    title = "Five modules on a shared standardized-effect scale",
    subtitle = "Blue = lower; red = higher; black outlines pass BH FDR; colors clipped at +/-2 SD"
  )

save_plot_pair(
  "02_standardized_effect_global_graphs",
  standardized_graph,
  width = 16,
  height = 10
)

standardized_graph
```

## Step 5: Effect size and statistical evidence

This plot prevents a visually large color change from being mistaken for a
well-supported result. Every panel uses exactly the same q-value axis. Labels
mark every FDR hit plus the two smallest q-values in modules without a hit.

```{r evidence-plot, fig.width=13, fig.height=9}
evidence_labels <- module_results |>
  dplyr::group_by(.data$module_id) |>
  dplyr::arrange(.data$q_value, .by_group = TRUE) |>
  dplyr::mutate(label_candidate = dplyr::row_number() <= 2) |>
  dplyr::ungroup() |>
  dplyr::filter(.data$is_fdr_significant | .data$label_candidate)

evidence_plot <- ggplot(
  module_results |>
    dplyr::mutate(
      effect_direction = dplyr::if_else(
        .data$score_difference < 0,
        "lower",
        "higher"
      )
    ),
  aes(
    x = .data$standardized_effect,
    y = -log10(.data$q_value),
    color = .data$effect_direction
  )
) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed",
    color = "grey45"
  ) +
  geom_vline(xintercept = 0, color = "grey65") +
  geom_point(size = 2, alpha = 0.8) +
  ggrepel::geom_text_repel(
    data = evidence_labels,
    aes(label = .data$cell_type),
    color = "black",
    size = 2.5,
    max.overlaps = Inf,
    min.segment.length = 0,
    seed = 1,
    show.legend = FALSE
  ) +
  facet_wrap(vars(.data$module_label), ncol = 2) +
  scale_color_manual(
    values = c(lower = "#2166AC", higher = "#B2182B"),
    name = "module score"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  ) +
  labs(
    title = "Effect magnitude and evidence across modules",
    subtitle = "Dashed line marks BH q = 0.05",
    x = "standardized effect (foxj1a - control)",
    y = "-log10(BH q)"
  )

save_plot_pair(
  "03_standardized_effect_and_q",
  evidence_plot,
  width = 13,
  height = 9
)

evidence_plot
```

### One overlaid view of the q-value distributions

Cell types are ranked independently within each module. Because the y-axis is
shared, vertical separation directly represents stronger statistical evidence.

```{r ranked-q-profiles, fig.width=10, fig.height=6}
ranked_q_values <- module_results |>
  dplyr::group_by(.data$module_id, .data$module_label) |>
  dplyr::arrange(.data$q_value, .by_group = TRUE) |>
  dplyr::mutate(q_rank = dplyr::row_number()) |>
  dplyr::ungroup()

readr::write_tsv(
  ranked_q_values |>
    dplyr::select(module_id, module_label, cell_type, q_rank, q_value),
  file.path(DATA_DIR, "ranked_q_values_by_module.tsv")
)

ranked_q_plot <- ggplot(
  ranked_q_values,
  aes(
    x = .data$q_rank,
    y = -log10(.data$q_value),
    color = .data$module_label
  )
) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed",
    color = "grey35"
  ) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  scale_color_manual(
    values = c(
      "Broad cilia" = "#7B3294",
      "Motile cilia" = "#008837",
      "Basal body" = "#E08214",
      "Ciliary tip" = "#2166AC",
      "Transition zone" = "#C51B7D"
    ),
    name = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    title = "Motile-cilia results separate from the other modules",
    subtitle = "Each line ranks the 92 cell types from smallest to largest BH q; dashed line is q = 0.05",
    x = "cell-type rank within module",
    y = "-log10(BH q)"
  )

save_plot_pair(
  "03b_ranked_q_value_profiles",
  ranked_q_plot,
  width = 10,
  height = 6
)

ranked_q_plot
```

## Step 6: Does normal foxj1a expression mark each module?

The motile module should be most active in cell types that normally express
foxj1a. Cell types are divided into three transparent groups: no detected
foxj1a, the lower half of positive expression, and the higher half of positive
expression. The exact Pearson and Spearman correlations are still exported.

```{r baseline-coupling, fig.width=13, fig.height=9}
foxj1a_baseline <- readr::read_tsv(
  FOXJ1A_BASELINE_FILE,
  show_col_types = FALSE
) |>
  dplyr::select(.data$cell_type, .data$control_foxj1a_expression)

baseline_coupling <- module_results |>
  dplyr::left_join(foxj1a_baseline, by = "cell_type")

baseline_correlations <- baseline_coupling |>
  dplyr::group_by(.data$module_id, .data$module_label) |>
  dplyr::summarise(
    pearson = cor(
      .data$control_foxj1a_expression,
      .data$control_mean,
      method = "pearson"
    ),
    spearman = cor(
      .data$control_foxj1a_expression,
      .data$control_mean,
      method = "spearman"
    ),
    .groups = "drop"
  )

readr::write_tsv(
  baseline_correlations,
  file.path(DATA_DIR, "baseline_foxj1a_module_correlations.tsv")
)

baseline_annotations <- baseline_correlations |>
  dplyr::mutate(
    label = paste0(
      "Pearson r = ", sprintf("%.2f", .data$pearson),
      "\nSpearman rho = ", sprintf("%.2f", .data$spearman)
    ),
    x = Inf,
    y = -Inf
  )

positive_expression_median <- stats::median(
  foxj1a_baseline$control_foxj1a_expression[
    foxj1a_baseline$control_foxj1a_expression > 0
  ]
)

baseline_coupling <- baseline_coupling |>
  dplyr::mutate(
    foxj1a_expression_group = dplyr::case_when(
      .data$control_foxj1a_expression == 0 ~ "not detected",
      .data$control_foxj1a_expression <= positive_expression_median ~
        "lower positive",
      TRUE ~ "higher positive"
    ),
    foxj1a_expression_group = factor(
      .data$foxj1a_expression_group,
      levels = c("not detected", "lower positive", "higher positive")
    )
  )

expression_group_counts <- baseline_coupling |>
  dplyr::distinct(.data$cell_type, .data$foxj1a_expression_group) |>
  dplyr::count(.data$foxj1a_expression_group, name = "n_cell_types")

readr::write_tsv(
  expression_group_counts,
  file.path(DATA_DIR, "foxj1a_expression_group_counts.tsv")
)

baseline_plot <- ggplot(
  baseline_coupling,
  aes(
    x = .data$foxj1a_expression_group,
    y = .data$control_mean,
    fill = .data$foxj1a_expression_group
  )
) +
  geom_boxplot(width = 0.65, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(
    aes(color = .data$foxj1a_expression_group),
    width = 0.12,
    height = 0,
    size = 1.6,
    alpha = 0.75,
    show.legend = FALSE
  ) +
  facet_wrap(vars(.data$module_label), ncol = 2, scales = "free_y") +
  scale_fill_manual(
    values = c(
      "not detected" = "#D9D9D9",
      "lower positive" = "#80CDC1",
      "higher positive" = "#018571"
    ),
    guide = "none"
  ) +
  scale_color_manual(
    values = c(
      "not detected" = "#7A7A7A",
      "lower positive" = "#35978F",
      "higher positive" = "#01665E"
    ),
    guide = "none"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 20, hjust = 1)
  ) +
  labs(
    title = "Module activity across foxj1a-expression groups",
    subtitle = paste0(
      "Each point is one cell type; positive expression split at the median (",
      sprintf("%.3g", positive_expression_median),
      ")"
    ),
    x = "mean control foxj1a expression group",
    y = "mean control module score"
  )

save_plot_pair(
  "04_baseline_foxj1a_module_coupling",
  baseline_plot,
  width = 13,
  height = 9
)

baseline_plot
```

## Step 7: Show embryos for the strongest candidates

The three smallest q-values from each module are shown, including all FDR hits.
This is the quickest check for effects driven by one unusual embryo.

```{r candidate-embryos, fig.width=14, fig.height=14}
candidate_results <- module_results |>
  dplyr::group_by(.data$module_id) |>
  dplyr::arrange(.data$q_value, .by_group = TRUE) |>
  dplyr::slice_head(n = 3) |>
  dplyr::ungroup() |>
  dplyr::select(
    module_id,
    module_label,
    cell_type,
    score_difference,
    q_value,
    is_fdr_significant
  )

embryo_scores <- readr::read_tsv(
  EMBRYO_SCORE_FILE,
  show_col_types = FALSE
) |>
  dplyr::filter(
    .data$timepoint == TIMEPOINT,
    .data$perturbation %in% c(PERTURBATION, CONTROL),
    .data$n_cells >= MINIMUM_CELLS
  ) |>
  dplyr::mutate(
    arm = dplyr::if_else(
      .data$perturbation == PERTURBATION,
      "perturbation",
      "control"
    ),
    arm = factor(.data$arm, levels = c("control", "perturbation"))
  )

candidate_embryo_values <- embryo_scores |>
  dplyr::inner_join(
    candidate_results,
    by = c("module_id", "cell_type")
  ) |>
  dplyr::mutate(
    panel_label = paste0(
      MODULE_LABELS[.data$module_id],
      "\n",
      stringr::str_wrap(.data$cell_type, width = 38),
      "\ndelta = ", sprintf("%.3g", .data$score_difference),
      "; q = ", format(.data$q_value, digits = 2, scientific = TRUE)
    )
  )

readr::write_tsv(
  dplyr::select(candidate_embryo_values, -panel_label),
  file.path(DATA_DIR, "top_candidate_embryo_values.tsv")
)

candidate_embryo_plot <- ggplot(
  candidate_embryo_values,
  aes(x = .data$arm, y = .data$mean_module_score, color = .data$arm)
) +
  geom_jitter(width = 0.10, height = 0, size = 1.8, alpha = 0.85) +
  stat_summary(
    fun = mean,
    geom = "crossbar",
    width = 0.5,
    color = "black",
    linewidth = 0.35
  ) +
  facet_wrap(vars(.data$panel_label), scales = "free_y", ncol = 3) +
  scale_color_manual(values = arm_colors, guide = "none") +
  theme_minimal(base_size = 9) +
  theme(
    axis.title.x = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 7.5)
  ) +
  labs(
    title = "Strongest three cell-type results from each module",
    subtitle = "Each point is one embryo; black bar is the group mean",
    y = "mean module score"
  )

save_plot_pair(
  "05_top_candidate_embryo_values",
  candidate_embryo_plot,
  width = 14,
  height = 14
)

candidate_embryo_plot
```

## Step 8: Show which cell types recur across modules

The rows are cilia modules and the columns are cell types. A node means that
the module and cell-type result has raw `p < 0.05`. Repeated nodes within one
column reveal cell types implicated by multiple modules.

```{r module-overlap-upset, fig.width=26, fig.height=13}
is_raw_p_05 <- module_results$p_value < 0.05
is_raw_p_01 <- module_results$p_value < 0.01
is_bh_q_05 <- module_results$q_value < 0.05

if (any(is_bh_q_05 & !is_raw_p_01)) {
  stop("The BH q < 0.05 set is not contained within raw p < 0.01.")
}
if (any(is_raw_p_01 & !is_raw_p_05)) {
  stop("The raw p < 0.01 set is not contained within raw p < 0.05.")
}

threshold_counts <- tibble::tibble(
  threshold = c("BH q < 0.05", "raw p < 0.01", "raw p < 0.05"),
  n_results = c(sum(is_bh_q_05), sum(is_raw_p_01), sum(is_raw_p_05)),
  n_added = c(
    sum(is_bh_q_05),
    sum(is_raw_p_01 & !is_bh_q_05),
    sum(is_raw_p_05 & !is_raw_p_01)
  )
)

readr::write_tsv(
  threshold_counts,
  file.path(DATA_DIR, "significance_threshold_counts.tsv")
)

# Keep every nominal candidate and record its strongest evidence tier.
overlap_candidates <- module_results |>
  dplyr::filter(.data$p_value < 0.05) |>
  dplyr::mutate(
    evidence_tier = dplyr::case_when(
      .data$q_value < 0.05 ~ "BH q < 0.05",
      .data$p_value < 0.01 ~ "raw p < 0.01 only",
      TRUE ~ "raw p < 0.05 only"
    ),
    evidence_tier = factor(
      .data$evidence_tier,
      levels = c(
        "BH q < 0.05",
        "raw p < 0.01 only",
        "raw p < 0.05 only"
      )
    ),
    effect_direction = dplyr::if_else(
      .data$score_difference < 0,
      "decrease",
      "increase"
    )
  )

# Sort by recurrence first, then by the strongest p-value in that cell type.
celltype_order <- overlap_candidates |>
  dplyr::group_by(.data$cell_type) |>
  dplyr::summarise(
    n_modules = dplyr::n_distinct(.data$module_id),
    strongest_p = min(.data$p_value),
    .groups = "drop"
  ) |>
  dplyr::arrange(
    dplyr::desc(.data$n_modules),
    .data$strongest_p,
    .data$cell_type
  )

ordered_celltypes <- celltype_order$cell_type
ordered_module_labels <- rev(unname(MODULE_LABELS[MODULES]))

overlap_candidates <- overlap_candidates |>
  dplyr::mutate(
    cell_type = factor(.data$cell_type, levels = ordered_celltypes),
    module_label = factor(
      as.character(.data$module_label),
      levels = ordered_module_labels
    )
  )

# Small open circles show every possible module and cell-type combination.
matrix_background <- tidyr::expand_grid(
  cell_type = factor(ordered_celltypes, levels = ordered_celltypes),
  module_label = factor(ordered_module_labels, levels = ordered_module_labels)
)

readr::write_tsv(
  overlap_candidates,
  file.path(DATA_DIR, "module_overlap_upset_matrix.tsv")
)

module_count_plot <- ggplot(
  celltype_order |>
    dplyr::mutate(
      cell_type = factor(.data$cell_type, levels = ordered_celltypes)
    ),
  aes(x = .data$cell_type, y = .data$n_modules)
) +
  geom_col(width = 0.72, fill = "#4C78A8") +
  geom_text(
    aes(label = .data$n_modules),
    vjust = -0.30,
    size = 4
  ) +
  scale_y_continuous(
    limits = c(0, 4.7),
    expand = expansion(mult = c(0, 0))
  ) +
  theme_void() +
  theme(
    plot.margin = margin(t = 2, r = 5, b = 0, l = 5)
  )

module_matrix_plot <- ggplot(
  matrix_background,
  aes(x = .data$cell_type, y = .data$module_label)
) +
  geom_point(
    shape = 21,
    size = 1.8,
    stroke = 0.35,
    fill = "white",
    color = "#D0D0D0"
  ) +
  geom_point(
    data = overlap_candidates,
    aes(
      fill = .data$effect_direction,
      size = .data$evidence_tier
    ),
    shape = 21,
    color = "#303030",
    stroke = 0.65
  ) +
  scale_fill_manual(
    values = c(
      decrease = "#2166AC",
      increase = "#B2182B"
    ),
    name = "module-score change"
  ) +
  scale_size_manual(
    values = c(
      "BH q < 0.05" = 9,
      "raw p < 0.01 only" = 6,
      "raw p < 0.05 only" = 3.5
    ),
    name = "evidence"
  ) +
  guides(
    fill = guide_legend(
      override.aes = list(shape = 21, size = 5, color = "#303030")
    ),
    size = guide_legend(
      override.aes = list(shape = 21, fill = "#808080", color = "#303030")
    )
  ) +
  scale_x_discrete(
    labels = function(labels) stringr::str_wrap(labels, width = 20)
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(
      angle = 58,
      hjust = 1,
      vjust = 1,
      size = 10,
      lineheight = 0.92
    ),
    axis.text.y = element_text(size = 11),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "#ECECEC"),
    legend.position = "bottom",
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    plot.margin = margin(t = 0, r = 5, b = 0, l = 5)
  )

module_overlap_plot <- module_count_plot / module_matrix_plot +
  patchwork::plot_layout(heights = c(0.55, 2.0), guides = "collect") +
  patchwork::plot_annotation(
    title = "Cell types implicated across multiple cilia modules",
    subtitle = paste0(
      "Foxj1a at 48 hpf; columns are ordered by module recurrence and then ",
      "statistical evidence. Blue decreases, red increases, and larger nodes ",
      "have stronger evidence."
    ),
    theme = theme(
      plot.title = element_text(size = 18, face = "bold"),
      plot.subtitle = element_text(size = 12)
    )
  ) &
  theme(legend.position = "bottom")

save_plot_pair(
  "06_module_overlap_upset_matrix",
  module_overlap_plot,
  width = 20,
  height = 9.5
)

module_overlap_plot
```

## Step 9: Inspect every nominal raw-p candidate

The next artifact includes every module and cell-type result with raw
`p < 0.05`. These are **nominal candidates**, not FDR-supported discoveries.
The complete PDF has one page per module, and matching PNG pages make the
results easy to inspect without rendering the PDF.

```{r all-raw-p-candidates, fig.show='hide'}
raw_p_candidates <- module_results |>
  dplyr::filter(.data$p_value < 0.05) |>
  dplyr::arrange(.data$module_id, .data$p_value) |>
  dplyr::select(
    module_id,
    module_label,
    cell_type,
    score_difference,
    p_value,
    q_value,
    is_fdr_significant
  )

raw_p_candidate_values <- embryo_scores |>
  dplyr::inner_join(
    raw_p_candidates,
    by = c("module_id", "cell_type")
  ) |>
  dplyr::mutate(
    panel_label = paste0(
      stringr::str_wrap(.data$cell_type, width = 38),
      "\ndelta = ", sprintf("%.3g", .data$score_difference),
      "; p = ", format(.data$p_value, digits = 2, scientific = TRUE),
      "; q = ", format(.data$q_value, digits = 2, scientific = TRUE)
    )
  )

readr::write_tsv(
  dplyr::select(raw_p_candidate_values, -panel_label),
  file.path(DATA_DIR, "raw_p_05_candidate_embryo_values.tsv")
)

raw_candidate_plots <- list()

for (module_name in MODULES) {
  module_values <- raw_p_candidate_values |>
    dplyr::filter(as.character(.data$module_id) == .env$module_name) |>
    dplyr::mutate(
      panel_label = factor(
        .data$panel_label,
        levels = unique(.data$panel_label[order(.data$p_value)])
      )
    )

  module_plot <- ggplot(
    module_values,
    aes(x = .data$arm, y = .data$mean_module_score, color = .data$arm)
  ) +
    geom_jitter(width = 0.10, height = 0, size = 2.0, alpha = 0.85) +
    stat_summary(
      fun = mean,
      geom = "crossbar",
      width = 0.5,
      color = "black",
      linewidth = 0.4
    ) +
    facet_wrap(vars(.data$panel_label), scales = "free_y", ncol = 3) +
    scale_color_manual(values = arm_colors, guide = "none") +
    theme_minimal(base_size = 10) +
    theme(
      axis.title.x = element_blank(),
      panel.grid.minor = element_blank(),
      strip.text = element_text(size = 8)
    ) +
    labs(
      title = paste0(
        MODULE_LABELS[[module_name]],
        ": all nominal candidates with raw p < 0.05"
      ),
      subtitle = paste0(
        nrow(raw_p_candidates |> dplyr::filter(
          as.character(.data$module_id) == .env$module_name
        )),
        " cell types; each point is one embryo; black bar is the group mean"
      ),
      y = "mean module score"
    )

  raw_candidate_plots[[module_name]] <- module_plot

  ggsave(
    file.path(
      FIGURE_DIR,
      paste0("07_raw_p_candidates_", module_name, ".png")
    ),
    module_plot,
    device = ragg::agg_png,
    width = 14,
    height = 11,
    units = "in",
    dpi = 200,
    bg = "white",
    limitsize = FALSE
  )
}

# One PDF page per module keeps all 39 panels readable in one artifact.
grDevices::pdf(
  file.path(FIGURE_DIR, "07_raw_p_candidate_embryo_values.pdf"),
  width = 14,
  height = 11,
  onefile = TRUE
)
for (module_name in MODULES) {
  print(raw_candidate_plots[[module_name]])
}
grDevices::dev.off()

raw_p_candidates |>
  dplyr::count(.data$module_label, name = "n_raw_p_candidates")
```
