---
title: "Are positive motile-cilia responses larger than sampling noise?"
output:
  html_document:
    toc: true
    toc_float: true
---

This is a deliberately separate exploratory analysis. It asks whether cell
types with a higher gold motile-cilia score after foxj1a perturbation look like
plausible compensatory responses or like ordinary embryo-to-embryo variation.

The word **adaptation is not used as a result** here. A positive change must be
stable across embryos, larger than a control-only sampling null, and coherent
across component genes before that interpretation becomes convincing.

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(ggplot2)
library(patchwork)

knitr::opts_chunk$set(
  dev = "ragg_png",
  fig.bg = "white"
)

PROJECT_DIR <- paste0(
  "/net/trapnell/vol1/home/mdcolon/proj/morphseq/results/",
  "mcolon/20260727_gene14_clean"
)
MODULE_DIR <- file.path(PROJECT_DIR, "cilia_gene_module")
PARENT_DIR <- file.path(MODULE_DIR, "foxj1a_48hpf_inspection")
HERE <- file.path(PARENT_DIR, "positive_response_validation")
DATA_DIR <- file.path(HERE, "data")
FIGURE_DIR <- file.path(HERE, "figures")

dir.create(DATA_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(FIGURE_DIR, recursive = TRUE, showWarnings = FALSE)

MODEL_RESULT_FILE <- file.path(
  MODULE_DIR,
  "output",
  "differential_module_results",
  "foxj1a_48hpf.tsv"
)
EMBRYO_SCORE_FILE <- file.path(
  MODULE_DIR,
  "output",
  "embryo_celltype_cilia_module_scores.tsv"
)

PERTURBATION <- "foxj1a"
CONTROL <- "ctrl-inj"
TIMEPOINT <- 48
MODULE <- "gold_motile_cilia"
MINIMUM_CELLS <- 10L
NUMBER_OF_CANDIDATES_TO_SHOW <- 6L

arm_colors <- c(
  control = "#4C78A8",
  perturbation = "#E45756"
)
```

## Step 1: Read the completed model and embryo scores

The fitted model is not rerun. We reuse its effect, interval, and p-value, then
look directly at the embryo means that entered the comparison.

```{r read-data}
model_results <- readr::read_tsv(
  MODEL_RESULT_FILE,
  show_col_types = FALSE
)

embryo_scores <- readr::read_tsv(
  EMBRYO_SCORE_FILE,
  show_col_types = FALSE
) |>
  dplyr::filter(
    .data$timepoint == TIMEPOINT,
    .data$module_id == MODULE,
    .data$perturbation %in% c(PERTURBATION, CONTROL),
    .data$n_cells >= MINIMUM_CELLS
  ) |>
  dplyr::mutate(
    arm = dplyr::if_else(
      .data$perturbation == PERTURBATION,
      "perturbation",
      "control"
    ),
    arm = factor(.data$arm, levels = c("control", "perturbation"))
  )

gold_results <- model_results |>
  dplyr::filter(.data$module_id == MODULE) |>
  dplyr::mutate(
    # The model q-value corrects across all five modules and cell types.
    # This second q-value asks the narrower gold-motile-only question.
    gold_module_q_value = stats::p.adjust(.data$p_value, method = "BH")
  )

positive_results <- gold_results |>
  dplyr::filter(.data$score_difference > 0) |>
  dplyr::arrange(.data$p_value)

cat(nrow(positive_results), "of", nrow(gold_results),
    "tested cell types have a positive point estimate.\n")
```

## Step 2: Ask whether one embryo controls each result

For every positive result, we remove one embryo at a time and recalculate the
plain difference in group means. A candidate is fragile if its direction
changes after removing just one embryo.

```{r leave-one-embryo-out}
leave_one_out_rows <- list()
row_number <- 1L

for (cell_type_name in positive_results$cell_type) {
  cell_type_scores <- embryo_scores |>
    dplyr::filter(.data$cell_type == cell_type_name)

  for (embryo_to_remove in cell_type_scores$embryo_ID) {
    remaining_scores <- cell_type_scores |>
      dplyr::filter(.data$embryo_ID != embryo_to_remove)

    group_means <- remaining_scores |>
      dplyr::group_by(.data$arm) |>
      dplyr::summarise(
        mean_score = mean(.data$mean_module_score),
        .groups = "drop"
      )

    leave_one_out_rows[[row_number]] <- tibble::tibble(
      cell_type = cell_type_name,
      removed_embryo_ID = embryo_to_remove,
      removed_arm = as.character(
        cell_type_scores$arm[cell_type_scores$embryo_ID == embryo_to_remove]
      ),
      score_difference_without_embryo =
        group_means$mean_score[group_means$arm == "perturbation"] -
        group_means$mean_score[group_means$arm == "control"]
    )

    row_number <- row_number + 1L
  }
}

leave_one_out_results <- dplyr::bind_rows(leave_one_out_rows)

leave_one_out_summary <- leave_one_out_results |>
  dplyr::group_by(.data$cell_type) |>
  dplyr::summarise(
    minimum_leave_one_out_difference =
      min(.data$score_difference_without_embryo),
    maximum_leave_one_out_difference =
      max(.data$score_difference_without_embryo),
    fraction_leave_one_out_positive =
      mean(.data$score_difference_without_embryo > 0),
    direction_reverses_without_one_embryo =
      any(.data$score_difference_without_embryo <= 0),
    .groups = "drop"
  )

readr::write_tsv(
  leave_one_out_results,
  file.path(DATA_DIR, "leave_one_embryo_out_results.tsv")
)
```

## Step 3: Build a control-only sampling-noise floor

Within each cell type, the control embryos are divided into every possible pair
of similarly sized pseudo-arms. We then measure the difference between those
pseudo-arms. This answers a narrow question:

> How large a difference can appear when we repeatedly divide control embryos?

This is intentionally conservative because each pseudo-arm is smaller than the
real experimental arms. It captures sampling noise, but **cannot capture the
foxj1a-versus-control plate confounding**, because all pseudo-arms come from the
same control plate.

```{r control-only-null}
control_split_rows <- list()
split_number <- 1L

for (cell_type_name in positive_results$cell_type) {
  control_scores <- embryo_scores |>
    dplyr::filter(
      .data$cell_type == cell_type_name,
      .data$arm == "control"
    ) |>
    dplyr::arrange(.data$embryo_ID)

  n_control <- nrow(control_scores)

  if (n_control < 6L) {
    next
  }

  first_arm_size <- floor(n_control / 2)
  first_arm_combinations <- utils::combn(
    seq_len(n_control),
    first_arm_size,
    simplify = FALSE
  )

  # With equal arm sizes, complementary splits are duplicates. Keeping splits
  # that contain embryo 1 leaves one copy of each comparison.
  if (n_control %% 2L == 0L) {
    first_arm_combinations <- first_arm_combinations[
      vapply(first_arm_combinations, function(rows) 1L %in% rows, logical(1))
    ]
  }

  for (first_arm_rows in first_arm_combinations) {
    second_arm_rows <- setdiff(seq_len(n_control), first_arm_rows)

    control_split_rows[[split_number]] <- tibble::tibble(
      cell_type = cell_type_name,
      split_id = split_number,
      n_first_pseudo_arm = length(first_arm_rows),
      n_second_pseudo_arm = length(second_arm_rows),
      pseudo_arm_difference =
        mean(control_scores$mean_module_score[first_arm_rows]) -
        mean(control_scores$mean_module_score[second_arm_rows])
    )

    split_number <- split_number + 1L
  }
}

control_split_null <- dplyr::bind_rows(control_split_rows)

control_split_summary <- control_split_null |>
  dplyr::group_by(.data$cell_type) |>
  dplyr::summarise(
    n_control_splits = dplyr::n(),
    control_null_95 = stats::quantile(
      abs(.data$pseudo_arm_difference),
      0.95
    ),
    control_null_99 = stats::quantile(
      abs(.data$pseudo_arm_difference),
      0.99
    ),
    .groups = "drop"
  )

candidate_summary <- positive_results |>
  dplyr::left_join(leave_one_out_summary, by = "cell_type") |>
  dplyr::left_join(control_split_summary, by = "cell_type") |>
  dplyr::rowwise() |>
  dplyr::mutate(
    empirical_control_split_p = {
      null_values <- control_split_null$pseudo_arm_difference[
        control_split_null$cell_type == .data$cell_type
      ]

      if (length(null_values) == 0L) {
        NA_real_
      } else {
        (1 + sum(abs(null_values) >= abs(.data$score_difference))) /
          (1 + length(null_values))
      }
    },
    effect_over_control_null_95 =
      abs(.data$score_difference) / .data$control_null_95
  ) |>
  dplyr::ungroup() |>
  dplyr::arrange(.data$p_value) |>
  dplyr::mutate(evidence_rank = dplyr::row_number())

readr::write_tsv(
  control_split_null,
  file.path(DATA_DIR, "control_only_split_null_values.tsv")
)
readr::write_tsv(
  candidate_summary,
  file.path(DATA_DIR, "positive_candidate_summary.tsv")
)

candidate_summary |>
  dplyr::select(
    evidence_rank,
    cell_type,
    score_difference,
    confidence_interval_low,
    confidence_interval_high,
    p_value,
    q_value,
    gold_module_q_value,
    empirical_control_split_p,
    fraction_leave_one_out_positive,
    direction_reverses_without_one_embryo
  ) |>
  head(12)
```

## Step 4: Show the leading positive candidates

The six candidates with the smallest model p-values are shown. This ranking is
for inspection only; it does not turn a nominal p-value into significance.

```{r candidate-plots, fig.width=13, fig.height=10}
top_candidates <- candidate_summary |>
  dplyr::slice_head(n = NUMBER_OF_CANDIDATES_TO_SHOW)

top_candidate_values <- embryo_scores |>
  dplyr::inner_join(
    dplyr::select(top_candidates, cell_type, evidence_rank),
    by = "cell_type"
  ) |>
  dplyr::mutate(
    panel_label = paste0(
      .data$evidence_rank,
      ". ",
      .data$cell_type
    )
  )

readr::write_tsv(
  dplyr::select(top_candidate_values, -panel_label),
  file.path(DATA_DIR, "top_candidate_embryo_values.tsv")
)

raw_embryo_plot <- ggplot(
  top_candidate_values,
  aes(
    x = .data$arm,
    y = .data$mean_module_score,
    color = .data$arm
  )
) +
  geom_jitter(width = 0.10, height = 0, size = 2.2, alpha = 0.85) +
  stat_summary(
    fun = mean,
    geom = "crossbar",
    width = 0.5,
    color = "black",
    linewidth = 0.35
  ) +
  facet_wrap(vars(.data$panel_label), scales = "free_y", ncol = 3) +
  scale_color_manual(values = arm_colors, guide = "none") +
  theme_minimal(base_size = 11) +
  theme(
    axis.title.x = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 8)
  ) +
  labs(
    title = "Leading positive motile-cilia responses",
    subtitle = "Each point is one embryo; the black bar is the group mean",
    y = "mean gold motile-cilia score"
  )

ggsave(
  file.path(FIGURE_DIR, "01_top_candidate_embryo_values.pdf"),
  raw_embryo_plot,
  width = 13,
  height = 8
)
ggsave(
  file.path(FIGURE_DIR, "01_top_candidate_embryo_values.png"),
  raw_embryo_plot,
  device = ragg::agg_png,
  width = 13,
  height = 8,
  dpi = 200,
  bg = "white"
)

raw_embryo_plot
```

## Step 5: Compare each observed gain with its sampling-noise floor

Points above the diagonal are larger than the 95th percentile of their own
cell type's control-only split null. The second panel shows how much the effect
moves when any single embryo is removed.

```{r stability-plots, fig.width=13, fig.height=7}
noise_floor_plot <- candidate_summary |>
  dplyr::filter(!is.na(.data$control_null_95)) |>
  ggplot(
    aes(
      x = .data$control_null_95,
      y = .data$score_difference,
      color = .data$empirical_control_split_p <= 0.05
    )
  ) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray45") +
  geom_point(size = 2.6) +
  scale_color_manual(
    values = c(`FALSE` = "gray60", `TRUE` = "#E45756"),
    name = "above control-only\nnoise floor"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    title = "Observed gain versus control-only sampling noise",
    x = "95th percentile of\n|control pseudo-arm difference|",
    y = "observed score gain"
  )

stability_plot <- candidate_summary |>
  dplyr::slice_head(n = 12) |>
  dplyr::mutate(
    cell_type = factor(.data$cell_type, levels = rev(.data$cell_type))
  ) |>
  ggplot(aes(y = .data$cell_type)) +
  geom_vline(xintercept = 0, color = "gray55") +
  geom_segment(
    aes(
      x = .data$minimum_leave_one_out_difference,
      xend = .data$maximum_leave_one_out_difference,
      yend = .data$cell_type
    ),
    color = "gray45"
  ) +
  geom_point(aes(x = .data$score_difference), color = "#E45756", size = 2.5) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "Effect after removing any one embryo",
    subtitle = "Red point is the full result; line is the leave-one-out range",
    x = "motile-score difference",
    y = NULL
  )

stability_figure <- noise_floor_plot + stability_plot +
  patchwork::plot_layout(widths = c(0.9, 1.1))

ggsave(
  file.path(FIGURE_DIR, "02_noise_floor_and_stability.pdf"),
  stability_figure,
  width = 13,
  height = 7
)
ggsave(
  file.path(FIGURE_DIR, "02_noise_floor_and_stability.png"),
  stability_figure,
  device = ragg::agg_png,
  width = 13,
  height = 7,
  dpi = 200,
  bg = "white"
)

stability_figure
```

## Step 6: State what these compact artifacts cannot test

The existing component-gene artifact contains spinal-cord radial glia only. It
was created to validate the strong **negative** response and cannot establish
gene-level coherence in the positive candidates. The compact cell-score RDS
contains module scores but not individual gene expression, so it also cannot
test `foxj1b` or other compensatory regulators.

```{r data-availability}
data_availability <- tibble::tribble(
  ~question, ~available_without_large_cds, ~reason,
  "Embryo-level positive module scores", TRUE,
  "Available for every tested cell type in the compact embryo-score table.",
  "Control-only sampling null", TRUE,
  "Can be computed from the compact embryo-score table.",
  "Positive-candidate component-gene coherence", FALSE,
  "The existing gene table contains spinal-cord radial glia only.",
  "Leave-one-gene-out module sensitivity", FALSE,
  "Individual component genes are absent for the positive candidates.",
  "foxj1b and compensatory regulator expression", FALSE,
  "The compact cell-score object contains module scores, not gene expression.",
  "Recurrence at another valid foxj1a time point", FALSE,
  "Only the valid 48-hpf foxj1a module contrast has been fitted here."
)

readr::write_tsv(
  data_availability,
  file.path(DATA_DIR, "data_availability.tsv")
)

data_availability
```

## Interpretation rule

A positive candidate becomes more credible when:

1. its confidence interval excludes zero;
2. its FDR-adjusted q-value is small;
3. its gain exceeds the control-only noise floor;
4. removing one embryo never reverses the direction; and
5. many component genes rise together.

The first four can be inspected here. The fifth requires a deliberately scoped
gene-expression extraction for the selected candidates. Even then, the
foxj1a/control plate confounding remains and requires independent replication
or a plate-aware experimental comparison.
