---
title: "Differential cell-type abundance"
output: html_document
params:
  mode: "check_plan"
  contrast_name: !r NULL
---

This notebook asks:

> Does a perturbation change the abundance of a cell type relative to its control?

It can check the contrast plan, fit one contrast, fit all contrasts, or assemble
completed fits.

The default mode is `check_plan`; fitting occurs only when explicitly requested.
`fit_all` fits every missing contrast and then assembles the final table.

```{r setup, message=FALSE, warning=FALSE}
library(monocle3)
library(hooke)
library(platt)
library(dplyr)
library(readr)
library(tibble)
library(tidyr)

PROJECT_DIR <- paste0(
  "/net/trapnell/vol1/home/mdcolon/proj/morphseq/results/",
  "mcolon/20260727_gene14_clean"
)
ABUNDANCE_DIR <- file.path(PROJECT_DIR, "abundance_dact")

CELLTYPE_COUNTS_FILE <- file.path(
  PROJECT_DIR,
  "cell_type_counts",
  "per_embryo_celltype_counts.tsv"
)
MORPHSEQ_MAP_FILE <- file.path(
  PROJECT_DIR,
  "0_shared",
  "morphseq_embryo_sequence_map.csv"
)
MCCLINTOCK_CDS <- paste0(
  "/net/seahub_zfish/vol1/data/seahub_rna_processing/portal_inputs/",
  "v3.1.0/mcclintock/GENE14/run_1/filter_embryos/embryo_filtered_cds"
)

OUTPUT_DIR <- file.path(ABUNDANCE_DIR, "output")
RAW_MODEL_RESULTS_DIR <- file.path(OUTPUT_DIR, "raw_model_results")
FINAL_OUTPUT_FILE <- file.path(OUTPUT_DIR, "differential_abundance.tsv")

source(file.path(ABUNDANCE_DIR, "contrast_plan.R"))
source(file.path(ABUNDANCE_DIR, "cell_type_gate_policy.R"))

allowed_modes <- c("check_plan", "fit_one", "fit_all", "assemble")
if (!params$mode %in% allowed_modes) {
  stop("mode must be one of: ", paste(allowed_modes, collapse = ", "))
}
if (params$mode == "fit_one" && is.null(params$contrast_name)) {
  stop("fit_one mode requires a contrast_name.")
}

dir.create(RAW_MODEL_RESULTS_DIR, recursive = TRUE, showWarnings = FALSE)
```

## Step 1: Load the embryo cell counts

This table contains one row per embryo and cell type. A zero means that no
retained sequenced cells from that embryo were assigned to that cell type.

```{r load-counts}
celltype_counts <- readr::read_tsv(
  CELLTYPE_COUNTS_FILE,
  show_col_types = FALSE
)

# Keep one preferred MorphSeq phenotype label per sequencing embryo.
phenotype_labels <- readr::read_csv(
  MORPHSEQ_MAP_FILE,
  col_select = c(
    pred_seq_embryo_ID,
    predicted_phenotype,
    is_preferred_morphseq_embryo,
    in_mcclintock
  ),
  col_types = readr::cols(
    pred_seq_embryo_ID = readr::col_character(),
    predicted_phenotype = readr::col_character(),
    is_preferred_morphseq_embryo = readr::col_logical(),
    in_mcclintock = readr::col_logical()
  )
) |>
  dplyr::filter(
    .data$is_preferred_morphseq_embryo %in% TRUE,
    .data$in_mcclintock %in% TRUE,
    !is.na(.data$predicted_phenotype)
  ) |>
  dplyr::transmute(
    embryo_ID = .data$pred_seq_embryo_ID,
    predicted_phenotype = .data$predicted_phenotype
  )

if (anyDuplicated(phenotype_labels$embryo_ID)) {
  stop("Each sequencing embryo must have at most one preferred phenotype.")
}

celltype_counts <- celltype_counts |>
  dplyr::left_join(phenotype_labels, by = "embryo_ID") |>
  dplyr::mutate(
    # Controls keep their sequencing perturbation name. Phenotyped mutants use
    # their MorphSeq label, such as High_to_Low or CE.
    phenotype_group = dplyr::coalesce(
      .data$predicted_phenotype,
      .data$perturbation
    )
  )

cat(
  dplyr::n_distinct(celltype_counts$embryo_ID), "embryos x",
  dplyr::n_distinct(celltype_counts$cell_type), "cell types\n"
)
```

## Step 2: Check every planned contrast

For each contrast, we confirm that:

- The perturbation and control both exist.
- Both groups have enough embryos.
- Each group has one RT block.
- The observed cross-block status matches the plan.

The function is defined here because it is used immediately below.

```{r define-contrast-audit}
audit_one_contrast <- function(celltype_counts, contrast) {
  contrast_name <- contrast$contrast_name
  perturbation_name <- contrast$perturbation
  control_name <- contrast$control
  timepoint_hpf <- contrast$timepoint
  group_column <- contrast$group_column

  if (!group_column %in% names(celltype_counts)) {
    stop("The cell-type counts are missing group column: ", group_column)
  }

  embryos <- celltype_counts |>
    dplyr::filter(
      .data$timepoint == .env$timepoint_hpf,
      .data[[group_column]] %in%
        c(.env$perturbation_name, .env$control_name)
    ) |>
    dplyr::distinct(
      .data$embryo_ID,
      analysis_group = .data[[group_column]],
      .data$rt_block
    )

  # After distinct(), multiple rows for one embryo mean that its analysis group
  # or RT-block assignment conflicts across the input table.
  metadata_rows_per_embryo <- embryos |>
    dplyr::count(.data$embryo_ID, name = "n_metadata_rows")
  embryos_with_conflicting_metadata <- metadata_rows_per_embryo |>
    dplyr::filter(.data$n_metadata_rows != 1)

  if (nrow(embryos_with_conflicting_metadata) > 0) {
    stop(
      contrast_name,
      " contains embryos with conflicting group or RT-block metadata."
    )
  }

  perturbation_embryos <- embryos |>
    dplyr::filter(.data$analysis_group == .env$perturbation_name)
  control_embryos <- embryos |>
    dplyr::filter(.data$analysis_group == .env$control_name)

  perturbation_rt_blocks <- unique(
    stats::na.omit(perturbation_embryos$rt_block)
  )
  control_rt_blocks <- unique(
    stats::na.omit(control_embryos$rt_block)
  )

  if (length(perturbation_rt_blocks) != 1L) {
    stop(contrast_name, " must have one perturbation RT block.")
  }
  if (length(control_rt_blocks) != 1L) {
    stop(contrast_name, " must have one control RT block.")
  }

  n_perturbation_embryos <- dplyr::n_distinct(
    perturbation_embryos$embryo_ID
  )
  n_control_embryos <- dplyr::n_distinct(
    control_embryos$embryo_ID
  )

  if (n_perturbation_embryos < MINIMUM_EMBRYOS_PER_ARM) {
    stop(contrast_name, " has too few perturbation embryos.")
  }
  if (n_control_embryos < MINIMUM_EMBRYOS_PER_ARM) {
    stop(contrast_name, " has too few control embryos.")
  }

  perturbation_rt_block <- perturbation_rt_blocks[[1]]
  control_rt_block <- control_rt_blocks[[1]]
  is_cross_rt_block <- perturbation_rt_block != control_rt_block

  if (is_cross_rt_block != contrast$expected_cross_rt_block) {
    stop(
      contrast_name,
      " expected cross-RT-block = ",
      contrast$expected_cross_rt_block,
      ", but observed ",
      perturbation_rt_block,
      " versus ",
      control_rt_block,
      "."
    )
  }

  tibble::tibble(
    contrast_name = contrast_name,
    group_column = group_column,
    perturbation = perturbation_name,
    control = control_name,
    timepoint = timepoint_hpf,
    control_note = contrast$control_note,
    perturbation_rt_block = perturbation_rt_block,
    control_rt_block = control_rt_block,
    is_cross_rt_block = is_cross_rt_block,
    n_perturbation_embryos = n_perturbation_embryos,
    n_control_embryos = n_control_embryos
  )
}
```

Now apply that check to every row of `contrast_plan`.

```{r run-contrast-audit}
contrast_audit_rows <- vector("list", nrow(contrast_plan))

for (i in seq_len(nrow(contrast_plan))) {
  contrast_audit_rows[[i]] <- audit_one_contrast(
    celltype_counts = celltype_counts,
    contrast = contrast_plan[i, ]
  )
}

contrast_audit <- dplyr::bind_rows(contrast_audit_rows)
contrast_audit
```

When `mode = "check_plan"`, the notebook stops here.

## Step 3: Fit differential abundance

This step runs only in `fit_one` or `fit_all` mode.

Existing raw model checkpoints are skipped, so fitting is resumable.

### Step 3A: Choose the contrasts

First, show exactly which contrasts will be fitted.

```{r select-contrasts, eval=params$mode %in% c("fit_one", "fit_all")}
if (params$mode == "fit_one") {
  contrasts_to_fit <- contrast_audit |>
    dplyr::filter(.data$contrast_name == params$contrast_name)

  if (nrow(contrasts_to_fit) != 1L) {
    stop("Unknown contrast_name: ", params$contrast_name)
  }
} else {
  contrasts_to_fit <- contrast_audit
}

contrasts_to_fit |>
  dplyr::select(
    contrast_name,
    group_column,
    perturbation,
    control,
    timepoint
  )
```

### Step 3B: Load the model input

This is the large single-cell dataset. It is loaded only when fitting was
explicitly requested.

```{r load-model-input, eval=params$mode %in% c("fit_one", "fit_all")}
message("Loading McClintock CDS...")
cds <- monocle3::load_monocle_objects(MCCLINTOCK_CDS)

# Attach the same preferred phenotype labels used in the embryo count table.
phenotype_group <- phenotype_labels$predicted_phenotype[
  match(colData(cds)$embryo_ID, phenotype_labels$embryo_ID)
]
phenotype_group <- dplyr::coalesce(
  phenotype_group,
  as.character(colData(cds)$perturbation)
)
colData(cds)$phenotype_group <- phenotype_group

# Platt requires a batch column. We intentionally fit no batch effect here:
# every sample receives the same level, so batch_one is a no-op covariate.
colData(cds)$batch_one <- "one_batch"

message("Building Hooke cell-count set...")
cell_count_set <- hooke::new_cell_count_set(
  cds,
  sample_group = "embryo_ID",
  cell_group = "cell_type"
)

rm(cds)
invisible(gc())

num_threads <- suppressWarnings(
  as.integer(Sys.getenv("NSLOTS", "3"))
)
if (is.na(num_threads) || num_threads < 1L) {
  num_threads <- 3L
}
```

### Step 3C: Fit the requested contrasts

The important model calls remain visible in the loop:

```text
fit_genotype_ccm()
get_perturbation_effects()
```

```{r fit-contrasts, eval=params$mode %in% c("fit_one", "fit_all")}
for (i in seq_len(nrow(contrasts_to_fit))) {
  contrast <- contrasts_to_fit[i, ]
  saved_result <- file.path(
    RAW_MODEL_RESULTS_DIR,
    paste0(contrast$contrast_name, ".tsv")
  )

  if (file.exists(saved_result)) {
    message("Already complete: ", contrast$contrast_name)
    next
  }

  sample_metadata <- colData(cell_count_set)
  keep <- (
    sample_metadata$timepoint == contrast$timepoint &
    sample_metadata[[contrast$group_column]] %in%
      c(contrast$perturbation, contrast$control)
  )
  contrast_ccs <- cell_count_set[, keep]

  message(
    "Fitting ", contrast$contrast_name, ": ",
    contrast$perturbation, " versus ", contrast$control,
    " at ", contrast$timepoint, " hpf"
  )

  model <- platt:::fit_genotype_ccm(
    genotype = contrast$perturbation,
    ccs = contrast_ccs,
    perturbation_col = contrast$group_column,
    ctrl_ids = contrast$control,
    interval_col = "timepoint",
    batch_col = "batch_one",
    vhat_method = "sandwich_var",
    num_threads = num_threads
  )

  model_results <- platt:::get_perturbation_effects(
    model,
    interval_col = "timepoint"
  )

  rm(model)
  invisible(gc())

  required_columns <- c(
    "cell_group",
    "log_abund_x",
    "log_abund_y",
    "delta_log_abund",
    "delta_p_value",
    "delta_q_value"
  )
  missing_columns <- setdiff(required_columns, names(model_results))
  if (length(missing_columns) > 0L) {
    stop(
      contrast$contrast_name,
      " model output is missing: ",
      paste(missing_columns, collapse = ", ")
    )
  }

  # Platt defines x = control, y = perturbation, and delta = y - x.
  model_results <- model_results |>
    dplyr::rename(
      cell_type = cell_group,
      log_abundance_control = log_abund_x,
      log_abundance_perturbation = log_abund_y,
      log_abundance_difference = delta_log_abund,
      p_value = delta_p_value,
      q_value_all = delta_q_value
    ) |>
    dplyr::mutate(
      contrast_name = contrast$contrast_name,
      group_column = contrast$group_column,
      perturbation = contrast$perturbation,
      control = contrast$control,
      timepoint = contrast$timepoint,
      control_note = contrast$control_note,
      perturbation_rt_block = contrast$perturbation_rt_block,
      control_rt_block = contrast$control_rt_block,
      is_cross_rt_block = contrast$is_cross_rt_block,
      n_perturbation_embryos = contrast$n_perturbation_embryos,
      n_control_embryos = contrast$n_control_embryos,
      .before = 1
    )

  readr::write_tsv(model_results, saved_result)
  message("Wrote ungated model result: ", saved_result)
}
```

## Step 4: Add reliability to each completed contrast

This step runs in `assemble` or `fit_all` mode.

The ungated model output stays unchanged on disk. Reliability is added to a copy
that will become the final combined table.

```{r assemble-results, eval=params$mode %in% c("assemble", "fit_all")}
expected_files <- file.path(
  RAW_MODEL_RESULTS_DIR,
  paste0(contrast_audit$contrast_name, ".tsv")
)
missing_files <- expected_files[!file.exists(expected_files)]

if (length(missing_files) > 0L) {
  stop(
    "Cannot assemble the final table. Missing model results:\n  ",
    paste(basename(missing_files), collapse = "\n  ")
  )
}

assembled_results <- vector("list", nrow(contrast_audit))

for (i in seq_len(nrow(contrast_audit))) {
  contrast <- contrast_audit[i, ]
  model_results <- readr::read_tsv(
    expected_files[[i]],
    show_col_types = FALSE
  )

  expected_identity <- list(
    contrast_name = contrast$contrast_name,
    group_column = contrast$group_column,
    perturbation = contrast$perturbation,
    control = contrast$control,
    timepoint = contrast$timepoint
  )
  for (column_name in names(expected_identity)) {
    observed_values <- unique(model_results[[column_name]])
    values_match <- (
      length(observed_values) == 1L &&
      isTRUE(
        all.equal(
          observed_values[[1]],
          expected_identity[[column_name]],
          check.attributes = FALSE
        )
      )
    )
    if (!values_match) {
      stop(
        basename(expected_files[[i]]),
        " does not match the current contrast plan."
      )
    }
  }

  detection_summary <- summarize_celltype_detection(
    celltype_counts = celltype_counts,
    group_column = contrast$group_column,
    perturbation_name = contrast$perturbation,
    control_name = contrast$control,
    timepoint_hpf = contrast$timepoint
  )

  assembled_results[[i]] <- add_celltype_reliability(
    model_results = model_results,
    detection_summary = detection_summary,
    gate_policy = CELLTYPE_GATE_POLICY
  )
}
```

## Step 5: Write the final table

One row represents one contrast and one cell type. No modeled cell types are
removed.

```{r write-final-results, eval=params$mode %in% c("assemble", "fit_all")}
differential_abundance <- dplyr::bind_rows(assembled_results) |>
  dplyr::arrange(.data$contrast_name, .data$cell_type)

readr::write_tsv(
  differential_abundance,
  FINAL_OUTPUT_FILE
)

cat(
  "Wrote", nrow(differential_abundance),
  "contrast x cell-type rows to\n",
  FINAL_OUTPUT_FILE,
  "\n"
)
```
