---
title: "Build embryo-level cilia module scores"
output: html_document
---

This notebook creates the table needed for differential cilia-module testing.

Every cell already has a module score. Here, cells of the same type from the
same embryo are summarized together so that each embryo remains one biological
replicate.

This table is deliberately kept separate from the CDS:

- CDS `colData` has one row per cell.
- This output has one row per embryo, cell type, and module.

No cell type is converted to zero when it is absent from an embryo.

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(readr)

PROJECT_DIR <- paste0(
  "/net/trapnell/vol1/home/mdcolon/proj/morphseq/results/",
  "mcolon/20260727_gene14_clean"
)
MODULE_DIR <- file.path(PROJECT_DIR, "cilia_gene_module")
OUTPUT_DIR <- file.path(MODULE_DIR, "output")

CELL_SCORE_FILE <- file.path(
  OUTPUT_DIR,
  "gene14_cell_cilia_scores.rds"
)
CELLTYPE_COUNT_FILE <- file.path(
  PROJECT_DIR,
  "cell_type_counts",
  "per_embryo_celltype_counts.tsv"
)
EMBRYO_MODULE_FILE <- file.path(
  OUTPUT_DIR,
  "embryo_celltype_cilia_module_scores.tsv"
)
COVERAGE_AUDIT_FILE <- file.path(
  OUTPUT_DIR,
  "embryo_celltype_module_coverage.tsv"
)

source(file.path(MODULE_DIR, "cilia_gene_module_policy.R"))
source(file.path(PROJECT_DIR, "abundance_dact", "contrast_plan.R"))

# We inspect several possible cell-count requirements before choosing one for
# significance testing. This notebook does not silently select a threshold.
CELL_COUNT_THRESHOLDS_TO_AUDIT <- c(1L, 5L, 10L, 20L)
```

## Step 1: Read the existing per-cell scores

The expensive CDS scoring has already been completed. This compact checkpoint
contains cell identity, embryo identity, cell type, and all module scores.

```{r read-cell-scores}
cell_scores <- readRDS(CELL_SCORE_FILE)

required_columns <- c(
  "cell",
  "embryo_ID",
  "cell_type",
  "perturbation",
  "timepoint",
  CHOSEN_CILIA_MODULES
)
missing_columns <- setdiff(required_columns, names(cell_scores))
if (length(missing_columns) > 0L) {
  stop(
    "The per-cell score checkpoint is missing: ",
    paste(missing_columns, collapse = ", ")
  )
}

cell_scores <- cell_scores |>
  dplyr::filter(
    !is.na(.data$embryo_ID),
    !is.na(.data$cell_type)
  )

cat(
  nrow(cell_scores), "cells from",
  dplyr::n_distinct(cell_scores$embryo_ID), "embryos\n"
)
```

## Step 2: Verify embryo identity and attach RT block

Perturbation, collection time, and RT block must each have one value per
embryo. The RT-block value comes from the existing cell-type count table.

```{r embryo-metadata}
score_embryo_metadata <- cell_scores |>
  dplyr::distinct(
    .data$embryo_ID,
    .data$perturbation,
    .data$timepoint
  )

conflicting_score_metadata <- score_embryo_metadata |>
  dplyr::count(.data$embryo_ID, name = "n_metadata_rows") |>
  dplyr::filter(.data$n_metadata_rows != 1L)

if (nrow(conflicting_score_metadata) > 0L) {
  stop("Some embryos have conflicting perturbation or timepoint metadata.")
}

celltype_counts <- readr::read_tsv(
  CELLTYPE_COUNT_FILE,
  show_col_types = FALSE
)

count_embryo_metadata <- celltype_counts |>
  dplyr::distinct(
    .data$embryo_ID,
    .data$perturbation,
    .data$timepoint,
    .data$rt_block
  )

conflicting_count_metadata <- count_embryo_metadata |>
  dplyr::count(.data$embryo_ID, name = "n_metadata_rows") |>
  dplyr::filter(.data$n_metadata_rows != 1L)

if (nrow(conflicting_count_metadata) > 0L) {
  stop("Some embryos have conflicting RT-block metadata.")
}

embryo_metadata <- score_embryo_metadata |>
  dplyr::left_join(
    count_embryo_metadata,
    by = c("embryo_ID", "perturbation", "timepoint")
  )

if (anyNA(embryo_metadata$rt_block)) {
  stop("Some scored embryos are missing an RT-block assignment.")
}

embryo_metadata
```

## Step 3: Summarize cells within each embryo and cell type

For readability and memory efficiency, the five chosen modules are summarized
one at a time. The mean will be the primary differential-module measurement.
The median and fraction of nonzero cells remain available for interpretation
and sensitivity checks.

```{r summarize-modules}
module_tables <- vector("list", length(CHOSEN_CILIA_MODULES))

for (i in seq_along(CHOSEN_CILIA_MODULES)) {
  module_name <- CHOSEN_CILIA_MODULES[[i]]

  module_tables[[i]] <- cell_scores |>
    dplyr::group_by(
      .data$embryo_ID,
      .data$cell_type
    ) |>
    dplyr::summarise(
      module_id = .env$module_name,
      n_cells = dplyr::n(),
      mean_module_score = mean(.data[[module_name]]),
      median_module_score = stats::median(.data[[module_name]]),
      fraction_nonzero = mean(.data[[module_name]] > 0),
      .groups = "drop"
    )
}

embryo_module_scores <- dplyr::bind_rows(module_tables) |>
  dplyr::left_join(embryo_metadata, by = "embryo_ID") |>
  dplyr::select(
    embryo_ID,
    perturbation,
    timepoint,
    rt_block,
    cell_type,
    module_id,
    n_cells,
    mean_module_score,
    median_module_score,
    fraction_nonzero
  ) |>
  dplyr::arrange(
    .data$timepoint,
    .data$perturbation,
    .data$embryo_ID,
    .data$cell_type,
    .data$module_id
  )

n_observed_embryo_celltypes <- cell_scores |>
  dplyr::distinct(.data$embryo_ID, .data$cell_type) |>
  nrow()

expected_rows <- n_observed_embryo_celltypes * length(CHOSEN_CILIA_MODULES)

if (nrow(embryo_module_scores) != expected_rows) {
  stop("The embryo-level table does not have five modules per observed cell type.")
}

readr::write_tsv(embryo_module_scores, EMBRYO_MODULE_FILE)

cat(
  "Wrote", EMBRYO_MODULE_FILE, "with",
  nrow(embryo_module_scores), "rows\n"
)

embryo_module_scores
```

## Step 4: Audit cell-count thresholds before testing

This audit asks how many perturbation and control embryos have enough cells of
each type. Module scores share the same cells, so coverage is calculated once
per contrast and cell type rather than repeated five times.

```{r coverage-audit}
celltype_observations <- embryo_module_scores |>
  dplyr::distinct(
    .data$embryo_ID,
    .data$perturbation,
    .data$timepoint,
    .data$rt_block,
    .data$cell_type,
    .data$n_cells
  )

coverage_rows <- list()
row_number <- 1L

for (contrast_number in seq_len(nrow(perturbation_contrast_plan))) {
  contrast <- perturbation_contrast_plan[contrast_number, ]

  contrast_observations <- celltype_observations |>
    dplyr::filter(
      .data$timepoint == contrast$timepoint,
      .data$perturbation %in%
        c(contrast$perturbation, contrast$control)
    ) |>
    dplyr::mutate(
      arm = dplyr::if_else(
        .data$perturbation == contrast$perturbation,
        "perturbation",
        "control"
      )
    )

  for (minimum_cells in CELL_COUNT_THRESHOLDS_TO_AUDIT) {
    coverage_rows[[row_number]] <- contrast_observations |>
      dplyr::filter(.data$n_cells >= .env$minimum_cells) |>
      dplyr::group_by(.data$cell_type, .data$arm) |>
      dplyr::summarise(
        n_embryos = dplyr::n_distinct(.data$embryo_ID),
        .groups = "drop"
      ) |>
      tidyr::complete(
        cell_type,
        arm = c("perturbation", "control"),
        fill = list(n_embryos = 0L)
      ) |>
      tidyr::pivot_wider(
        names_from = arm,
        values_from = n_embryos,
        names_prefix = "n_"
      ) |>
      dplyr::mutate(
        contrast_name = contrast$contrast_name,
        perturbation = contrast$perturbation,
        control = contrast$control,
        timepoint = contrast$timepoint,
        expected_cross_rt_block = contrast$expected_cross_rt_block,
        minimum_cells_per_embryo = minimum_cells,
        passes_minimum_embryos =
          .data$n_perturbation >= MINIMUM_EMBRYOS_PER_ARM &
          .data$n_control >= MINIMUM_EMBRYOS_PER_ARM,
        .before = 1
      )

    row_number <- row_number + 1L
  }
}

coverage_audit <- dplyr::bind_rows(coverage_rows) |>
  dplyr::arrange(
    .data$contrast_name,
    .data$minimum_cells_per_embryo,
    .data$cell_type
  )

readr::write_tsv(coverage_audit, COVERAGE_AUDIT_FILE)

coverage_summary <- coverage_audit |>
  dplyr::group_by(
    .data$contrast_name,
    .data$minimum_cells_per_embryo
  ) |>
  dplyr::summarise(
    n_testable_cell_types = sum(.data$passes_minimum_embryos),
    .groups = "drop"
  )

coverage_summary
```

## What happens next?

This notebook stops before significance testing. After choosing the minimum
cell count, the next analysis will compare embryo-level mean module scores
between the perturbation and control arms. Individual cells will remain
visible for plotting but will not be treated as independent embryos.
