---
title: "Interpretable cilia gene-module scores"
output: html_document
params:
  mode: "prepare_modules"
---

This notebook asks:

> Which cell types express curated cilia programs, and where do those programs
> appear on the reference and GENE14 UMAPs?

The analysis keeps the biology visible. First we build and inspect the human
gene lists. Then we map them to zebrafish genes. Only after that audit do we
load the large single-cell objects and calculate scores.

The default mode is `prepare_modules`, which does not load a CDS.

```{r setup, message=FALSE, warning=FALSE}
library(monocle3)
library(Matrix)
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(readxl)
library(stringr)
library(ggplot2)
library(jsonlite)

PROJECT_DIR <- paste0(
  "/net/trapnell/vol1/home/mdcolon/proj/morphseq/results/",
  "mcolon/20260727_gene14_clean"
)
MODULE_DIR <- file.path(PROJECT_DIR, "cilia_gene_module")
REPOSITORY_DIR <- normalizePath(file.path(PROJECT_DIR, "..", "..", ".."))

PROJECT_CONFIG <- jsonlite::read_json(
  file.path(PROJECT_DIR, "project_config.json"),
  simplifyVector = TRUE
)

resolve_project_path <- function(path) {
  if (grepl("^/", path)) {
    return(path)
  }
  normalizePath(file.path(PROJECT_DIR, path), mustWork = FALSE)
}

REFERENCE_CDS <- resolve_project_path(PROJECT_CONFIG$reference_cds)
GENE14_CDS <- resolve_project_path(PROJECT_CONFIG$mcclintock_cds)
REFERENCE_EXPRESSION <- resolve_project_path(
  PROJECT_CONFIG$reference_expression
)
ZFIN_ORTHOLOGS <- resolve_project_path(PROJECT_CONFIG$zfin_human_orthologs)

SCGS_V1_FILE <- file.path(
  REPOSITORY_DIR,
  "software",
  "SysCilia",
  "source",
  "scgs.v1.tsv"
)
CILIOGENICS_SOURCE <- file.path(
  REPOSITORY_DIR,
  "software",
  "CilioGenics_Analysis",
  "source"
)
CILIOGENICS_GOLD <- file.path(
  CILIOGENICS_SOURCE,
  "Table_S3_Gold_standard.xlsx"
)
CILIOGENICS_MOTILE <- file.path(
  CILIOGENICS_SOURCE,
  "motile_genes.csv"
)
HGNC_NAMES <- file.path(CILIOGENICS_SOURCE, "hgnc_names.txt")

DATA_DIR <- file.path(MODULE_DIR, "data")
OUTPUT_DIR <- file.path(MODULE_DIR, "output")
FIGURE_DIR <- file.path(MODULE_DIR, "figures")

HUMAN_MODULE_FILE <- file.path(DATA_DIR, "human_cilia_modules.tsv")
ORTHOLOGY_FILE <- file.path(DATA_DIR, "human_to_zebrafish_cilia_map.tsv")
ZEBRAFISH_MODULE_FILE <- file.path(DATA_DIR, "zebrafish_cilia_modules.tsv")
MODULE_COVERAGE_FILE <- file.path(DATA_DIR, "cilia_module_mapping_coverage.tsv")

REFERENCE_SCORE_FILE <- file.path(OUTPUT_DIR, "reference_cell_cilia_scores.rds")
GENE14_SCORE_FILE <- file.path(OUTPUT_DIR, "gene14_cell_cilia_scores.rds")
CELLTYPE_SUMMARY_FILE <- file.path(OUTPUT_DIR, "cell_type_cilia_scores.tsv")

source(file.path(MODULE_DIR, "cilia_gene_module_policy.R"))

allowed_modes <- c(
  "prepare_modules",
  "score_reference",
  "score_gene14",
  "assemble",
  "all"
)
if (!params$mode %in% allowed_modes) {
  stop("mode must be one of: ", paste(allowed_modes, collapse = ", "))
}

dir.create(DATA_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(FIGURE_DIR, recursive = TRUE, showWarnings = FALSE)
```

## Step 1: Read the annotated SysCilia v1 genes

SysCilia v1 is our interpretable foundation. Its table records why each gene
was included, supporting publications, ciliome studies, and protein location.

The file contains 303 rows but 302 unique human genes because `DCDC2` occurs
twice. We retain one row per stable human Ensembl ID and symbol.

```{r load-scgs-v1, eval=params$mode %in% c("prepare_modules", "all")}
scgs_v1_columns <- c(
  "human_ensembl_id_v1",
  "human_symbol_v1",
  "human_gene_description",
  "scgs_note",
  "scgs_pubmed",
  "scgs_ciliomes",
  "scgs_localisation"
)

scgs_v1_raw <- readr::read_tsv(
  SCGS_V1_FILE,
  comment = "#",
  col_names = scgs_v1_columns,
  show_col_types = FALSE,
  trim_ws = TRUE
)

if (nrow(scgs_v1_raw) != 303L) {
  stop("Expected 303 SysCilia v1 rows; found ", nrow(scgs_v1_raw), ".")
}
if (dplyr::n_distinct(scgs_v1_raw$human_symbol_v1) != 302L) {
  stop("Expected 302 unique SysCilia v1 symbols.")
}

duplicated_v1_symbols <- scgs_v1_raw |>
  dplyr::count(.data$human_symbol_v1) |>
  dplyr::filter(.data$n > 1)

print(duplicated_v1_symbols)

scgs_v1 <- scgs_v1_raw |>
  dplyr::distinct(
    .data$human_ensembl_id_v1,
    .data$human_symbol_v1,
    .keep_all = TRUE
  )
```

### Update historical human symbols without losing them

The v1 table used Ensembl release 71. We use its stable Ensembl ID to attach
the approved human symbol distributed with CilioGenics. The old symbol remains
in the audit table, so changes such as `AZI1` to `CEP131` are explicit.

```{r update-human-symbols, eval=params$mode %in% c("prepare_modules", "all")}
hgnc_symbol_map <- readr::read_tsv(
  HGNC_NAMES,
  show_col_types = FALSE,
  progress = FALSE
) |>
  dplyr::transmute(
    human_ensembl_id_v1 = .data$`Ensembl gene ID`,
    approved_human_symbol = .data$`Approved symbol`
  ) |>
  dplyr::filter(
    !is.na(.data$human_ensembl_id_v1),
    .data$human_ensembl_id_v1 != "",
    !is.na(.data$approved_human_symbol),
    .data$approved_human_symbol != ""
  ) |>
  dplyr::distinct(.data$human_ensembl_id_v1, .keep_all = TRUE)

scgs_v1 <- scgs_v1 |>
  dplyr::left_join(hgnc_symbol_map, by = "human_ensembl_id_v1") |>
  dplyr::mutate(
    current_human_symbol = dplyr::coalesce(
      .data$approved_human_symbol,
      .data$human_symbol_v1
    ),
    symbol_was_updated =
      .data$current_human_symbol != .data$human_symbol_v1
  ) |>
  dplyr::select(-.data$approved_human_symbol)

cat(sum(scgs_v1$symbol_was_updated), "v1 symbols were updated.\n")

scgs_v1 |>
  dplyr::filter(.data$symbol_was_updated) |>
  dplyr::select(
    human_ensembl_id_v1,
    human_symbol_v1,
    current_human_symbol,
    scgs_localisation
  )
```

## Step 2: Reconstruct the gold-standard motile list

CilioGenics defines its motile evidence list from five human lung or tracheal
single-cell sources: Reyfman, Carraro, Habermann, Murthy, and Trachea.

For interpretability, we keep only genes that are also in the CilioGenics gold
standard. We calculate that intersection directly instead of relying on the
precomputed Venn table, which separates pair-only and triple-overlap genes.

```{r build-gold-motile, eval=params$mode %in% c("prepare_modules", "all")}
ciliogenics_gold_source <- readxl::read_excel(
  CILIOGENICS_GOLD,
  sheet = "o_gold_standard_list"
)

ciliogenics_gold_rows <- tibble::tibble(
  current_human_symbol = as.character(ciliogenics_gold_source[[1]])
) |>
  dplyr::filter(
    !is.na(.data$current_human_symbol),
    .data$current_human_symbol != "Gene_name",
    .data$current_human_symbol != ""
  )

if (nrow(ciliogenics_gold_rows) != 688L) {
  stop("Expected 688 CilioGenics gold rows.")
}

duplicated_gold_symbols <- ciliogenics_gold_rows |>
  dplyr::count(.data$current_human_symbol) |>
  dplyr::filter(.data$n > 1)
print(duplicated_gold_symbols)

ciliogenics_gold <- ciliogenics_gold_rows |>
  dplyr::distinct(.data$current_human_symbol)

if (nrow(ciliogenics_gold) != 687L) {
  stop("Expected 687 unique CilioGenics gold genes.")
}

ciliogenics_motile <- readr::read_delim(
  CILIOGENICS_MOTILE,
  delim = ";",
  show_col_types = FALSE
) |>
  dplyr::transmute(
    current_human_symbol = .data$Gene_name,
    Reyfman = .data$Reyfman,
    Carraro = .data$Carraro,
    Habermann = .data$Habermann,
    Murthy = .data$Murthy,
    Trachea = .data$Trachea
  ) |>
  dplyr::distinct(.data$current_human_symbol, .keep_all = TRUE)

if (nrow(ciliogenics_motile) != 356L) {
  stop("Expected 356 genes in the CilioGenics motile evidence list.")
}

gold_motile_genes <- ciliogenics_gold |>
  dplyr::inner_join(ciliogenics_motile, by = "current_human_symbol")

if (nrow(gold_motile_genes) != 142L) {
  stop(
    "Expected 142 gold-standard motile genes; found ",
    nrow(gold_motile_genes),
    "."
  )
}

cat(
  nrow(ciliogenics_gold), "unique CilioGenics gold genes\n",
  nrow(ciliogenics_motile), "five-study motile genes\n",
  nrow(gold_motile_genes), "gold-standard motile genes\n"
)
```

## Step 3: Build the interpretable human modules

The broad score is the union of SysCilia v1 and the gold-standard motile list.
The smaller v1 modules come directly from its curated localization column. A
multi-localized protein can—and should—appear in more than one module.

```{r build-human-modules, eval=params$mode %in% c("prepare_modules", "all")}
v1_all_membership <- scgs_v1 |>
  dplyr::transmute(
    module_id = "v1_all",
    current_human_symbol = .data$current_human_symbol,
    module_provenance = "SysCilia v1 curated gold standard (2013)"
  )

gold_motile_membership <- gold_motile_genes |>
  dplyr::transmute(
    module_id = "gold_motile_cilia",
    current_human_symbol = .data$current_human_symbol,
    module_provenance = paste(
      "CilioGenics gold standard intersected with five-study",
      "human lung/tracheal motile evidence"
    )
  )

combined_membership <- dplyr::bind_rows(
  v1_all_membership,
  gold_motile_membership
) |>
  dplyr::distinct(.data$current_human_symbol) |>
  dplyr::mutate(
    module_id = "combined_cilia",
    module_provenance = paste(
      "Union of annotated SysCilia v1 and CilioGenics",
      "gold-standard motile genes"
    ),
    .before = 1
  )

v1_location_membership <- build_v1_location_membership(scgs_v1)

human_module_membership <- dplyr::bind_rows(
  combined_membership,
  gold_motile_membership,
  v1_all_membership,
  v1_location_membership
) |>
  dplyr::distinct(
    .data$module_id,
    .data$current_human_symbol,
    .keep_all = TRUE
  )

human_module_sizes <- human_module_membership |>
  dplyr::count(.data$module_id, name = "n_human_genes") |>
  dplyr::arrange(dplyr::desc(.data$n_human_genes))

human_module_sizes
```

Keep the full v1 evidence beside module membership for auditing.

```{r write-human-modules, eval=params$mode %in% c("prepare_modules", "all")}
human_cilia_modules <- human_module_membership |>
  dplyr::left_join(
    scgs_v1 |>
      dplyr::select(
        current_human_symbol,
        human_ensembl_id_v1,
        human_symbol_v1,
        symbol_was_updated,
        human_gene_description,
        scgs_note,
        scgs_pubmed,
        scgs_ciliomes,
        scgs_localisation
      ),
    by = "current_human_symbol"
  ) |>
  dplyr::arrange(.data$module_id, .data$current_human_symbol)

readr::write_tsv(human_cilia_modules, HUMAN_MODULE_FILE)
cat("Wrote", HUMAN_MODULE_FILE, "\n")
```

## Step 4: Map human module genes to zebrafish Ensembl IDs

ZFIN is the source of human–zebrafish orthology. Its download contains one row
per supporting assertion, so several rows can describe the same gene pair. We
first keep unique human-symbol/ZFIN-gene pairs.

The reference-expression table then connects the ZFIN zebrafish symbol to the
Ensembl gene ID used as the row name of both v3.1.0 CDS objects.

```{r map-to-zebrafish, eval=params$mode %in% c("prepare_modules", "all")}
zfin_columns <- c(
  "zfin_gene_id",
  "zebrafish_symbol",
  "zebrafish_gene_name",
  "current_human_symbol",
  "human_gene_name",
  "human_omim_id",
  "human_ncbi_gene_id",
  "human_hgnc_id",
  "orthology_evidence_code",
  "zfin_publication_id",
  "orthology_evidence_name",
  "eco_id",
  "eco_name"
)

module_human_genes <- human_module_membership |>
  dplyr::distinct(.data$current_human_symbol)

zfin_orthologs <- readr::read_tsv(
  ZFIN_ORTHOLOGS,
  col_names = zfin_columns,
  show_col_types = FALSE,
  progress = FALSE
) |>
  dplyr::filter(
    .data$current_human_symbol %in%
      module_human_genes$current_human_symbol
  ) |>
  dplyr::distinct(
    .data$current_human_symbol,
    .data$zfin_gene_id,
    .data$zebrafish_symbol,
    .data$zebrafish_gene_name
  )

reference_gene_map <- readr::read_tsv(
  REFERENCE_EXPRESSION,
  col_select = c(gene_id, gene_short_name),
  show_col_types = FALSE,
  progress = FALSE
) |>
  dplyr::filter(
    !is.na(.data$gene_id),
    !is.na(.data$gene_short_name)
  ) |>
  dplyr::distinct(
    zebrafish_gene_id = .data$gene_id,
    zebrafish_symbol = .data$gene_short_name
  )

human_to_zebrafish <- module_human_genes |>
  dplyr::left_join(zfin_orthologs, by = "current_human_symbol") |>
  dplyr::left_join(reference_gene_map, by = "zebrafish_symbol") |>
  dplyr::group_by(.data$current_human_symbol) |>
  dplyr::mutate(
    n_zebrafish_orthologs = dplyr::n_distinct(
      .data$zfin_gene_id,
      na.rm = TRUE
    ),
    mapping_status = dplyr::case_when(
      is.na(.data$zfin_gene_id) ~ "no_zfin_ortholog",
      is.na(.data$zebrafish_gene_id) ~ "not_in_reference_gene_map",
      .data$n_zebrafish_orthologs == 1L ~ "one_zebrafish_ortholog",
      TRUE ~ "multiple_zebrafish_orthologs"
    )
  ) |>
  dplyr::ungroup() |>
  dplyr::arrange(.data$current_human_symbol, .data$zebrafish_symbol)

zebrafish_module_membership <- human_module_membership |>
  dplyr::left_join(human_to_zebrafish, by = "current_human_symbol") |>
  dplyr::filter(!is.na(.data$zebrafish_gene_id)) |>
  dplyr::distinct(
    .data$module_id,
    .data$current_human_symbol,
    .data$zfin_gene_id,
    .data$zebrafish_symbol,
    .data$zebrafish_gene_id,
    .keep_all = TRUE
  )

module_mapping_coverage <- human_module_membership |>
  dplyr::distinct(.data$module_id, .data$current_human_symbol) |>
  dplyr::left_join(
    human_to_zebrafish |>
      dplyr::group_by(.data$current_human_symbol) |>
      dplyr::summarise(
        has_zfin_ortholog = any(!is.na(.data$zfin_gene_id)),
        has_zebrafish_gene_id = any(!is.na(.data$zebrafish_gene_id)),
        n_zebrafish_orthologs = max(.data$n_zebrafish_orthologs),
        .groups = "drop"
      ),
    by = "current_human_symbol"
  ) |>
  dplyr::group_by(.data$module_id) |>
  dplyr::summarise(
    n_human_genes = dplyr::n(),
    n_with_zfin_ortholog = sum(.data$has_zfin_ortholog %in% TRUE),
    n_mapped_human_genes = sum(.data$has_zebrafish_gene_id %in% TRUE),
    n_one_to_many = sum(.data$n_zebrafish_orthologs > 1, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    mapping_fraction = .data$n_mapped_human_genes / .data$n_human_genes,
    will_be_scored =
      .data$n_mapped_human_genes >= MINIMUM_MAPPED_HUMAN_GENES
  ) |>
  dplyr::arrange(dplyr::desc(.data$n_human_genes))

readr::write_tsv(human_to_zebrafish, ORTHOLOGY_FILE)
readr::write_tsv(zebrafish_module_membership, ZEBRAFISH_MODULE_FILE)
readr::write_tsv(module_mapping_coverage, MODULE_COVERAGE_FILE)

module_mapping_coverage
```

Check our experiment’s cilia perturbations explicitly.

```{r audit-target-genes, eval=params$mode %in% c("prepare_modules", "all")}
target_gene_audit <- tibble::tibble(
  perturbation_gene = c("cep290", "b9d2", "ift88", "foxj1a"),
  current_human_symbol = c("CEP290", "B9D2", "IFT88", "FOXJ1")
) |>
  dplyr::left_join(human_to_zebrafish, by = "current_human_symbol") |>
  dplyr::left_join(
    human_module_membership |>
      dplyr::group_by(.data$current_human_symbol) |>
      dplyr::summarise(
        modules = paste(sort(.data$module_id), collapse = "; "),
        .groups = "drop"
      ),
    by = "current_human_symbol"
  )

target_gene_audit
```

When `mode = "prepare_modules"`, the notebook stops here.

## Step 5: Score one loaded CDS

This is the only large-data helper. It is defined here because it is used
immediately below.

For each human cilia gene, zebrafish paralogs are summed before scoring. Each
human ortholog therefore receives one equal module weight.

```{r define-score-dataset, eval=params$mode %in% c("score_reference", "score_gene14", "all")}
if (!file.exists(ORTHOLOGY_FILE) || !file.exists(HUMAN_MODULE_FILE)) {
  stop("Run prepare_modules before scoring a CDS.")
}

human_to_zebrafish <- readr::read_tsv(
  ORTHOLOGY_FILE,
  show_col_types = FALSE
)
human_module_membership <- readr::read_tsv(
  HUMAN_MODULE_FILE,
  show_col_types = FALSE
) |>
  dplyr::select(
    module_id,
    current_human_symbol,
    module_provenance
  ) |>
  dplyr::distinct()

score_one_dataset <- function(dataset_name, cds_path, score_file) {
  message("Loading ", dataset_name, " CDS...")
  cds <- monocle3::load_monocle_objects(cds_path)

  if (!"UMAP" %in% SingleCellExperiment::reducedDimNames(cds)) {
    stop(dataset_name, " CDS does not contain a UMAP embedding.")
  }

  scored <- score_cilia_modules(
    cds = cds,
    human_to_zebrafish = human_to_zebrafish,
    human_module_membership = human_module_membership
  )

  umap <- SingleCellExperiment::reducedDim(cds, "UMAP")
  umap <- tibble::tibble(
    cell = colnames(cds),
    UMAP_1 = umap[, 1],
    UMAP_2 = umap[, 2]
  )

  cell_metadata <- as.data.frame(SummarizedExperiment::colData(cds))

  # The CDS row names are the authoritative cell IDs. Some CDS objects also
  # contain a column named `cell`, so assign this column explicitly instead of
  # trying to add a second column with the same name.
  cell_metadata$cell <- rownames(cell_metadata)
  cell_metadata <- tibble::as_tibble(cell_metadata)

  metadata_to_keep <- intersect(
    c(
      "cell",
      "cell_type",
      "projection_group",
      "embryo_ID",
      "perturbation",
      "timepoint"
    ),
    names(cell_metadata)
  )

  cell_scores <- cell_metadata |>
    dplyr::select(dplyr::all_of(metadata_to_keep)) |>
    dplyr::left_join(umap, by = "cell") |>
    dplyr::left_join(scored$scores, by = "cell") |>
    dplyr::mutate(dataset = dataset_name, .after = "cell")

  if (nrow(cell_scores) != ncol(cds)) {
    stop(dataset_name, " score rows do not match the CDS cells.")
  }
  if (anyNA(cell_scores$UMAP_1) || anyNA(cell_scores$UMAP_2)) {
    stop(dataset_name, " has cells without UMAP coordinates.")
  }

  score_columns <- setdiff(
    names(scored$scores),
    "cell"
  )
  finite_scores <- vapply(
    cell_scores[score_columns],
    function(values) all(is.finite(values)),
    logical(1)
  )
  constant_scores <- vapply(
    cell_scores[score_columns],
    function(values) dplyr::n_distinct(values) <= 1L,
    logical(1)
  )
  if (!all(finite_scores)) {
    stop(dataset_name, " contains a non-finite cilia score.")
  }
  if (any(constant_scores)) {
    stop(
      dataset_name,
      " contains constant scores: ",
      paste(names(constant_scores)[constant_scores], collapse = ", ")
    )
  }

  saveRDS(cell_scores, score_file)
  readr::write_tsv(
    scored$mapped_module_sizes |>
      dplyr::mutate(dataset = dataset_name, .before = 1),
    file.path(
      OUTPUT_DIR,
      paste0(dataset_name, "_mapped_module_sizes.tsv")
    )
  )

  message("Wrote ", score_file, ": ", nrow(cell_scores), " cells")

  rm(cds)
  invisible(gc())
  invisible(cell_scores)
}
```

### Step 5A: Score the reference atlas

```{r score-reference, eval=params$mode %in% c("score_reference", "all")}
reference_scores <- score_one_dataset(
  dataset_name = "reference",
  cds_path = REFERENCE_CDS,
  score_file = REFERENCE_SCORE_FILE
)
rm(reference_scores)
invisible(gc())
```

### Step 5B: Score GENE14

```{r score-gene14, eval=params$mode %in% c("score_gene14", "all")}
gene14_scores <- score_one_dataset(
  dataset_name = "gene14",
  cds_path = GENE14_CDS,
  score_file = GENE14_SCORE_FILE
)
rm(gene14_scores)
invisible(gc())
```

## Step 6: Summarize cell types and draw the maps

This step reads the compact score checkpoints, not the large CDS objects.

Every cell contributes to the cell-type summaries. UMAP rendering uses a
deterministic sample of at most 200 cells per cell type so the PDFs remain
responsive while every annotated cell type stays represented.

```{r assemble-results, eval=params$mode %in% c("assemble", "all"), fig.width=13, fig.height=10}
missing_score_files <- c(
  REFERENCE_SCORE_FILE,
  GENE14_SCORE_FILE
)[!file.exists(c(REFERENCE_SCORE_FILE, GENE14_SCORE_FILE))]

if (length(missing_score_files) > 0L) {
  stop(
    "Cannot assemble; missing score checkpoints: ",
    paste(basename(missing_score_files), collapse = ", ")
  )
}

all_cell_scores <- dplyr::bind_rows(
  readRDS(REFERENCE_SCORE_FILE),
  readRDS(GENE14_SCORE_FILE)
)

metadata_columns <- c(
  "cell",
  "dataset",
  "cell_type",
  "projection_group",
  "embryo_ID",
  "perturbation",
  "timepoint",
  "UMAP_1",
  "UMAP_2"
)
score_columns <- setdiff(names(all_cell_scores), metadata_columns)

cell_type_cilia_scores <- all_cell_scores |>
  dplyr::mutate(
    cell_type = dplyr::coalesce(
      as.character(.data$cell_type),
      "unannotated"
    )
  ) |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(score_columns),
    names_to = "module_id",
    values_to = "score"
  ) |>
  dplyr::group_by(.data$dataset, .data$cell_type, .data$module_id) |>
  dplyr::summarise(
    n_cells = dplyr::n(),
    mean_score = mean(.data$score),
    median_score = stats::median(.data$score),
    fraction_nonzero = mean(.data$score > 0),
    .groups = "drop"
  )

readr::write_tsv(cell_type_cilia_scores, CELLTYPE_SUMMARY_FILE)

set.seed(1)
umap_cells <- all_cell_scores |>
  dplyr::mutate(random_order = stats::runif(dplyr::n())) |>
  dplyr::group_by(.data$dataset, .data$cell_type) |>
  dplyr::arrange(.data$random_order, .by_group = TRUE) |>
  dplyr::slice_head(n = 200L) |>
  dplyr::ungroup() |>
  dplyr::select(-.data$random_order) |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(score_columns),
    names_to = "module_id",
    values_to = "score"
  )

display_limits <- all_cell_scores |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(score_columns),
    names_to = "module_id",
    values_to = "score"
  ) |>
  dplyr::group_by(.data$module_id) |>
  dplyr::summarise(
    display_max = stats::quantile(.data$score, 0.99, na.rm = TRUE),
    .groups = "drop"
  )

umap_cells <- umap_cells |>
  dplyr::left_join(display_limits, by = "module_id") |>
  dplyr::mutate(
    display_score = pmin(.data$score, .data$display_max),
    display_fraction = dplyr::if_else(
      .data$display_max > 0,
      .data$display_score / .data$display_max,
      0
    )
  )

for (dataset_name in unique(umap_cells$dataset)) {
  dataset_umap <- umap_cells |>
    dplyr::filter(.data$dataset == .env$dataset_name)

  all_module_plot <- ggplot(
    dataset_umap,
    aes(
      x = .data$UMAP_1,
      y = .data$UMAP_2,
      color = .data$display_fraction
    )
  ) +
    geom_point(size = 0.08, alpha = 0.7) +
    # Every facet is the same UMAP, so all modules must use the same axes.
    facet_wrap(vars(.data$module_id)) +
    scale_color_viridis_c(option = "magma") +
    coord_equal() +
    theme_void(base_size = 10) +
    theme(
      strip.text = element_text(size = 9),
      legend.position = "bottom"
    ) +
    labs(
      title = paste(dataset_name, "cilia gene-module scores"),
      color = "fraction of\nmodule 99th %ile"
    )

  ggsave(
    file.path(FIGURE_DIR, paste0(dataset_name, "_all_cilia_module_umaps.pdf")),
    all_module_plot,
    width = 13,
    height = 10,
    limitsize = FALSE
  )
  ggsave(
    file.path(FIGURE_DIR, paste0(dataset_name, "_all_cilia_module_umaps.png")),
    all_module_plot,
    width = 13,
    height = 10,
    dpi = 200,
    bg = "white",
    limitsize = FALSE
  )

  focused_plot <- dataset_umap |>
    dplyr::filter(
      .data$module_id %in% c("combined_cilia", "gold_motile_cilia")
    ) |>
    ggplot(
      aes(
        x = .data$UMAP_1,
        y = .data$UMAP_2,
        color = .data$display_fraction
      )
    ) +
    geom_point(size = 0.1, alpha = 0.75) +
    facet_wrap(vars(.data$module_id)) +
    scale_color_viridis_c(option = "magma") +
    coord_equal() +
    theme_void(base_size = 11) +
    theme(legend.position = "bottom") +
    labs(
      title = paste(dataset_name, "combined and motile cilia scores"),
      color = "fraction of\nmodule 99th %ile"
    )

  ggsave(
    file.path(FIGURE_DIR, paste0(dataset_name, "_main_cilia_module_umaps.pdf")),
    focused_plot,
    width = 10,
    height = 5.5,
    limitsize = FALSE
  )
  ggsave(
    file.path(FIGURE_DIR, paste0(dataset_name, "_main_cilia_module_umaps.png")),
    focused_plot,
    width = 10,
    height = 5.5,
    dpi = 200,
    bg = "white",
    limitsize = FALSE
  )
}
```

The cell-type heatmap keeps every cell type and is deliberately tall so its
labels remain readable when opened as a PDF.

```{r cell-type-heatmap, eval=params$mode %in% c("assemble", "all"), fig.width=12, fig.height=16}
cell_type_order <- cell_type_cilia_scores |>
  dplyr::filter(
    .data$dataset == "reference",
    .data$module_id == "combined_cilia"
  ) |>
  dplyr::arrange(dplyr::desc(.data$median_score)) |>
  dplyr::pull("cell_type")

# Append any GENE14-only annotation instead of silently dropping it from the
# shared ordering.
cell_type_order <- c(
  cell_type_order,
  setdiff(unique(cell_type_cilia_scores$cell_type), cell_type_order)
)

heatmap_data <- cell_type_cilia_scores |>
  dplyr::group_by(.data$module_id) |>
  dplyr::mutate(
    module_display_max = stats::quantile(
      .data$median_score,
      0.99,
      na.rm = TRUE
    ),
    relative_median_score = dplyr::if_else(
      .data$module_display_max > 0,
      pmin(.data$median_score, .data$module_display_max) /
        .data$module_display_max,
      0
    )
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    cell_type = factor(.data$cell_type, levels = rev(cell_type_order)),
    module_id = factor(.data$module_id, levels = score_columns)
  )

heatmap_plot <- ggplot(
  heatmap_data,
  aes(
    x = .data$module_id,
    y = .data$cell_type,
    fill = .data$relative_median_score
  )
) +
  geom_tile(color = "white", linewidth = 0.05) +
  facet_wrap(vars(.data$dataset), nrow = 1) +
  scale_fill_viridis_c(option = "magma") +
  theme_minimal(base_size = 8) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(
    title = "Median cilia gene-module score by cell type",
    fill = "relative median\nwithin module"
  )

heatmap_height <- max(14, length(cell_type_order) * 0.13)
ggsave(
  file.path(FIGURE_DIR, "cell_type_cilia_module_heatmap.pdf"),
  heatmap_plot,
  width = 14,
  height = heatmap_height,
  limitsize = FALSE
)
ggsave(
  file.path(FIGURE_DIR, "cell_type_cilia_module_heatmap.png"),
  heatmap_plot,
  width = 14,
  height = heatmap_height,
  dpi = 200,
  bg = "white",
  limitsize = FALSE
)

main_module_cell_types <- cell_type_cilia_scores |>
  dplyr::filter(
    .data$module_id %in% c("combined_cilia", "gold_motile_cilia")
  ) |>
  dplyr::mutate(
    cell_type = factor(.data$cell_type, levels = rev(cell_type_order))
  )

ranked_plot <- ggplot(
  main_module_cell_types,
  aes(x = .data$median_score, y = .data$cell_type)
) +
  geom_point(aes(size = .data$n_cells), alpha = 0.8) +
  facet_grid(
    rows = vars(.data$dataset),
    cols = vars(.data$module_id),
    scales = "free_x"
  ) +
  scale_size_area(max_size = 3) +
  theme_minimal(base_size = 8) +
  theme(
    axis.title.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    title = "Cell types ranked by the reference combined-cilia score",
    x = "median score",
    size = "cells"
  )

ggsave(
  file.path(FIGURE_DIR, "ranked_main_cilia_cell_types.pdf"),
  ranked_plot,
  width = 12,
  height = heatmap_height,
  limitsize = FALSE
)
ggsave(
  file.path(FIGURE_DIR, "ranked_main_cilia_cell_types.png"),
  ranked_plot,
  width = 12,
  height = heatmap_height,
  dpi = 200,
  bg = "white",
  limitsize = FALSE
)

cell_type_cilia_scores |>
  dplyr::filter(
    .data$module_id %in% c("combined_cilia", "gold_motile_cilia")
  ) |>
  dplyr::arrange(
    .data$dataset,
    .data$module_id,
    dplyr::desc(.data$median_score)
  ) |>
  dplyr::group_by(.data$dataset, .data$module_id) |>
  dplyr::slice_head(n = 20L)
```
