---
title: "Which cilia modules describe distinct cell-type patterns?"
output: html_document
---

This analysis asks whether the cilia modules provide different information or
mostly rank cell types in the same way.

We use one median score per cell type. This means a cell type with thousands
of cells does not dominate the correlation simply because it is abundant.
Cell types represented by fewer than 50 cells are left out because their
median scores are unstable.

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(ggplot2)

PROJECT_DIR <- paste0(
  "/net/trapnell/vol1/home/mdcolon/proj/morphseq/results/",
  "mcolon/20260727_gene14_clean"
)
MODULE_DIR <- file.path(PROJECT_DIR, "cilia_gene_module")
OUTPUT_DIR <- file.path(MODULE_DIR, "output")
FIGURE_DIR <- file.path(MODULE_DIR, "figures")

CELLTYPE_SCORES_FILE <- file.path(
  OUTPUT_DIR,
  "cell_type_cilia_scores.tsv"
)
MODULE_COVERAGE_FILE <- file.path(
  MODULE_DIR,
  "data",
  "cilia_module_mapping_coverage.tsv"
)

CORRELATION_FILE <- file.path(
  OUTPUT_DIR,
  "module_correlations_by_cell_type.tsv"
)
DATASET_AGREEMENT_FILE <- file.path(
  OUTPUT_DIR,
  "module_reference_gene14_agreement.tsv"
)
MODULE_SUMMARY_FILE <- file.path(
  OUTPUT_DIR,
  "cilia_module_comparison_summary.tsv"
)

MINIMUM_CELLS_PER_CELL_TYPE <- 50L
```

## Step 1: Read the cell-type scores

These scores were produced by `cilia_gene_module.Rmd`. Each row describes one
dataset, cell type, and cilia module.

```{r read-scores}
cell_type_scores <- readr::read_tsv(
  CELLTYPE_SCORES_FILE,
  show_col_types = FALSE
)

required_columns <- c(
  "dataset",
  "cell_type",
  "module_id",
  "n_cells",
  "median_score"
)
missing_columns <- setdiff(required_columns, names(cell_type_scores))
if (length(missing_columns) > 0L) {
  stop(
    "The cell-type score table is missing: ",
    paste(missing_columns, collapse = ", ")
  )
}

cell_types_used <- cell_type_scores |>
  dplyr::filter(.data$n_cells >= MINIMUM_CELLS_PER_CELL_TYPE) |>
  dplyr::distinct(.data$dataset, .data$cell_type) |>
  dplyr::count(.data$dataset, name = "n_cell_types")

cell_types_used
```

## Step 2: Put modules beside one another

The resulting table has one row per cell type and one column per module. This
is the table used for every correlation below.

```{r build-score-matrices}
make_cell_type_matrix <- function(dataset_name) {
  cell_type_scores |>
    dplyr::filter(
      .data$dataset == .env$dataset_name,
      .data$n_cells >= MINIMUM_CELLS_PER_CELL_TYPE
    ) |>
    dplyr::select(
      cell_type,
      module_id,
      median_score
    ) |>
    tidyr::pivot_wider(
      names_from = module_id,
      values_from = median_score
    ) |>
    dplyr::arrange(.data$cell_type)
}

gene14_matrix <- make_cell_type_matrix("gene14")
reference_matrix <- make_cell_type_matrix("reference")

gene14_matrix
```

## Step 3: Correlate module rankings across cell types

Spearman correlation asks whether two modules rank cell types similarly. It
does not require their numerical score scales to match.

```{r calculate-correlations}
correlate_modules <- function(score_matrix, dataset_name) {
  correlation_matrix <- stats::cor(
    dplyr::select(score_matrix, -cell_type),
    method = "spearman",
    use = "pairwise.complete.obs"
  )

  as.data.frame(as.table(correlation_matrix)) |>
    tibble::as_tibble() |>
    dplyr::rename(
      module_1 = Var1,
      module_2 = Var2,
      spearman_correlation = Freq
    ) |>
    dplyr::mutate(dataset = dataset_name, .before = 1)
}

module_correlations <- dplyr::bind_rows(
  correlate_modules(gene14_matrix, "gene14"),
  correlate_modules(reference_matrix, "reference")
)

readr::write_tsv(module_correlations, CORRELATION_FILE)

# Show every unique GENE14 module pair, from most to least correlated.
gene14_module_pairs <- module_correlations |>
  dplyr::filter(
    .data$dataset == "gene14",
    as.character(.data$module_1) < as.character(.data$module_2)
  ) |>
  dplyr::arrange(dplyr::desc(.data$spearman_correlation))

gene14_module_pairs
```

Cluster the modules using GENE14, then apply that same order to both datasets.
This makes disagreements between the two heatmaps easy to see.

```{r correlation-heatmap, fig.width=11, fig.height=5.5}
gene14_correlation_matrix <- stats::cor(
  dplyr::select(gene14_matrix, -cell_type),
  method = "spearman"
)

module_order <- stats::hclust(
  stats::as.dist(1 - gene14_correlation_matrix)
)$labels

correlation_plot_data <- module_correlations |>
  dplyr::mutate(
    module_1 = factor(.data$module_1, levels = module_order),
    module_2 = factor(.data$module_2, levels = rev(module_order))
  )

correlation_plot <- ggplot(
  correlation_plot_data,
  aes(
    x = .data$module_1,
    y = .data$module_2,
    fill = .data$spearman_correlation
  )
) +
  geom_tile(color = "white", linewidth = 0.3) +
  facet_wrap(vars(.data$dataset), nrow = 1) +
  scale_fill_gradient2(
    low = "#3B4CC0",
    mid = "white",
    high = "#B40426",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  coord_equal() +
  theme_minimal(base_size = 9) +
  theme(
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(
    title = "Cilia-module correlation across cell types",
    subtitle = paste(
      "Spearman correlation of cell-type medians;",
      "at least 50 cells per cell type"
    ),
    fill = "Spearman\ncorrelation"
  )

ggsave(
  file.path(FIGURE_DIR, "cilia_module_correlation_heatmap.pdf"),
  correlation_plot,
  width = 11,
  height = 5.5
)
ggsave(
  file.path(FIGURE_DIR, "cilia_module_correlation_heatmap.png"),
  correlation_plot,
  width = 11,
  height = 5.5,
  dpi = 200,
  bg = "white"
)

correlation_plot
```

## Step 4: Ask whether each module is stable across datasets

For each module, we correlate reference and GENE14 scores using only cell
types present in both datasets. A stable module should rank those shared cell
types similarly in both datasets.

```{r compare-datasets}
shared_cell_type_scores <- cell_type_scores |>
  dplyr::filter(.data$n_cells >= MINIMUM_CELLS_PER_CELL_TYPE) |>
  dplyr::select(
    dataset,
    cell_type,
    module_id,
    median_score
  ) |>
  tidyr::pivot_wider(
    names_from = dataset,
    values_from = median_score
  ) |>
  dplyr::filter(
    !is.na(.data$reference),
    !is.na(.data$gene14)
  )

module_dataset_agreement <- shared_cell_type_scores |>
  dplyr::group_by(.data$module_id) |>
  dplyr::summarise(
    n_shared_cell_types = dplyr::n(),
    reference_gene14_correlation = stats::cor(
      .data$reference,
      .data$gene14,
      method = "spearman"
    ),
    .groups = "drop"
  ) |>
  dplyr::arrange(
    dplyr::desc(.data$reference_gene14_correlation)
  )

readr::write_tsv(
  module_dataset_agreement,
  DATASET_AGREEMENT_FILE
)

module_dataset_agreement
```

## Step 5: Put the decision information in one table

Correlation measures redundancy. The interquartile range reports how strongly
a module separates cell types. We also report how many cell types have a
nonzero median and how many distinct median values exist. These checks prevent
a sparse module with many tied zeroes from looking deceptively stable.

The final table retains module size and reference-versus-GENE14 agreement.

```{r module-summary}
module_coverage <- readr::read_tsv(
  MODULE_COVERAGE_FILE,
  show_col_types = FALSE
)

module_score_spread <- cell_type_scores |>
  dplyr::filter(.data$n_cells >= MINIMUM_CELLS_PER_CELL_TYPE) |>
  dplyr::group_by(.data$dataset, .data$module_id) |>
  dplyr::summarise(
    n_cell_types = dplyr::n(),
    fraction_cell_types_nonzero = mean(.data$median_score > 0),
    n_distinct_cell_type_scores = dplyr::n_distinct(.data$median_score),
    median_across_cell_types = stats::median(.data$median_score),
    cell_type_score_iqr = stats::IQR(.data$median_score),
    .groups = "drop"
  ) |>
  tidyr::pivot_wider(
    names_from = dataset,
    values_from = c(
      n_cell_types,
      fraction_cell_types_nonzero,
      n_distinct_cell_type_scores,
      median_across_cell_types,
      cell_type_score_iqr
    ),
    names_glue = "{dataset}_{.value}"
  )

module_comparison_summary <- module_coverage |>
  dplyr::select(
    module_id,
    n_human_genes,
    n_mapped_human_genes,
    mapping_fraction
  ) |>
  dplyr::left_join(module_score_spread, by = "module_id") |>
  dplyr::left_join(module_dataset_agreement, by = "module_id") |>
  dplyr::arrange(
    dplyr::desc(.data$reference_gene14_correlation)
  )

readr::write_tsv(
  module_comparison_summary,
  MODULE_SUMMARY_FILE
)

module_comparison_summary
```

This report describes module redundancy and stability. It does not test any
perturbation. Perturbation testing should happen only after selecting the
modules, using one value per embryo and cell type rather than treating cells
as independent replicates.
