---
title: "Differential cilia-module expression"
output: html_document
params:
  contrast_name: "all"
---

This analysis asks:

> Within a given cell type, do embryos in the perturbation arm have a different
> cilia-module score from embryos in the control arm?

The embryo is the biological replicate. Individual cells contribute to their
embryo's mean score but are never treated as independent embryos.

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(readr)
library(ggplot2)

PROJECT_DIR <- paste0(
  "/net/trapnell/vol1/home/mdcolon/proj/morphseq/results/",
  "mcolon/20260727_gene14_clean"
)
MODULE_DIR <- file.path(PROJECT_DIR, "cilia_gene_module")
OUTPUT_DIR <- file.path(MODULE_DIR, "output")
FIGURE_DIR <- file.path(MODULE_DIR, "figures")

EMBRYO_MODULE_FILE <- file.path(
  OUTPUT_DIR,
  "embryo_celltype_cilia_module_scores.tsv"
)
COVERAGE_AUDIT_FILE <- file.path(
  OUTPUT_DIR,
  "embryo_celltype_module_coverage.tsv"
)
CONTRAST_RESULT_DIR <- file.path(
  OUTPUT_DIR,
  "differential_module_results"
)
dir.create(CONTRAST_RESULT_DIR, recursive = TRUE, showWarnings = FALSE)

source(file.path(MODULE_DIR, "cilia_gene_module_policy.R"))
source(file.path(PROJECT_DIR, "abundance_dact", "contrast_plan.R"))

MINIMUM_CELLS_PER_EMBRYO <- 10L
FDR_THRESHOLD <- 0.05

if (params$contrast_name == "all") {
  contrasts_to_fit <- perturbation_contrast_plan
  RESULT_FILE <- file.path(
    OUTPUT_DIR,
    "differential_cilia_module_expression.tsv"
  )
} else {
  contrasts_to_fit <- perturbation_contrast_plan |>
    dplyr::filter(.data$contrast_name == params$contrast_name)

  if (nrow(contrasts_to_fit) != 1L) {
    stop("Unknown perturbation contrast: ", params$contrast_name)
  }

  RESULT_FILE <- file.path(
    CONTRAST_RESULT_DIR,
    paste0(params$contrast_name, ".tsv")
  )
}

contrasts_to_fit |>
  dplyr::select(
    contrast_name,
    perturbation,
    control,
    timepoint,
    expected_cross_rt_block
  )
```

## Step 1: Read the embryo-level scores

The input was created by `build_embryo_module_scores.Rmd`. It contains one row
per embryo, observed cell type, and chosen module.

```{r read-input}
embryo_module_scores <- readr::read_tsv(
  EMBRYO_MODULE_FILE,
  show_col_types = FALSE
)

required_columns <- c(
  "embryo_ID",
  "perturbation",
  "timepoint",
  "rt_block",
  "cell_type",
  "module_id",
  "n_cells",
  "mean_module_score"
)
missing_columns <- setdiff(required_columns, names(embryo_module_scores))
if (length(missing_columns) > 0L) {
  stop(
    "The embryo-level module table is missing: ",
    paste(missing_columns, collapse = ", ")
  )
}

unexpected_modules <- setdiff(
  unique(embryo_module_scores$module_id),
  CHOSEN_CILIA_MODULES
)
if (length(unexpected_modules) > 0L) {
  stop(
    "The embryo-level table contains unapproved modules: ",
    paste(unexpected_modules, collapse = ", ")
  )
}

duplicated_measurements <- embryo_module_scores |>
  dplyr::count(
    .data$embryo_ID,
    .data$cell_type,
    .data$module_id
  ) |>
  dplyr::filter(.data$n != 1L)

if (nrow(duplicated_measurements) > 0L) {
  stop("Expected one score per embryo, cell type, and module.")
}

cat(
  nrow(embryo_module_scores), "embryo × cell-type × module rows\n"
)
```

## Step 2: Apply the reliability requirements

A cell type is testable only when both arms contain at least three embryos with
at least ten cells of that type. The existing coverage audit makes this
decision before looking at module differences or p-values.

```{r select-tests}
coverage_audit <- readr::read_tsv(
  COVERAGE_AUDIT_FILE,
  show_col_types = FALSE
)

eligible_cell_types <- coverage_audit |>
  dplyr::filter(
    .data$contrast_name %in% contrasts_to_fit$contrast_name,
    .data$minimum_cells_per_embryo == MINIMUM_CELLS_PER_EMBRYO,
    .data$passes_minimum_embryos
  ) |>
  dplyr::select(
    contrast_name,
    cell_type,
    n_control,
    n_perturbation
  )

tests_per_contrast <- eligible_cell_types |>
  dplyr::count(
    .data$contrast_name,
    name = "n_eligible_cell_types"
  ) |>
  dplyr::mutate(
    n_module_tests =
      .data$n_eligible_cell_types * length(CHOSEN_CILIA_MODULES)
  )

tests_per_contrast
```

## Step 3: Fit one embryo-level model

The same calculation is repeated for every eligible cell type and module.

```text
mean_module_score ~ arm
```

`arm` is explicitly ordered as control followed by perturbation. Therefore,
the fitted coefficient is always perturbation minus control.

HC3 standard errors are used because embryo-level variability can differ
between the two arms. The standardized effect divides the raw score difference
by the pooled embryo-level standard deviation.

```{r define-model}
fit_one_module_model <- function(model_data) {
  model_data <- model_data |>
    dplyr::mutate(
      arm = factor(
        .data$arm,
        levels = c("control", "perturbation")
      )
    )

  control_scores <- model_data$mean_module_score[
    model_data$arm == "control"
  ]
  perturbation_scores <- model_data$mean_module_score[
    model_data$arm == "perturbation"
  ]

  n_control <- length(control_scores)
  n_perturbation <- length(perturbation_scores)
  control_mean <- mean(control_scores)
  perturbation_mean <- mean(perturbation_scores)
  score_difference <- perturbation_mean - control_mean

  pooled_variance <- (
    (n_control - 1L) * stats::var(control_scores) +
      (n_perturbation - 1L) * stats::var(perturbation_scores)
  ) / (n_control + n_perturbation - 2L)

  standardized_effect <- dplyr::if_else(
    is.finite(pooled_variance) && pooled_variance > 0,
    score_difference / sqrt(pooled_variance),
    NA_real_
  )

  if (dplyr::n_distinct(model_data$mean_module_score) <= 1L) {
    return(
      tibble::tibble(
        control_mean = control_mean,
        perturbation_mean = perturbation_mean,
        score_difference = score_difference,
        standardized_effect = standardized_effect,
        standard_error = NA_real_,
        confidence_interval_low = NA_real_,
        confidence_interval_high = NA_real_,
        p_value = NA_real_,
        model_status = "constant_score"
      )
    )
  }

  model <- stats::lm(
    mean_module_score ~ arm,
    data = model_data
  )
  robust_variance <- sandwich::vcovHC(model, type = "HC3")
  coefficient_table <- lmtest::coeftest(
    model,
    vcov. = robust_variance
  )

  coefficient_name <- "armperturbation"
  standard_error <- coefficient_table[coefficient_name, "Std. Error"]
  p_value <- coefficient_table[coefficient_name, "Pr(>|t|)"]
  critical_value <- stats::qt(
    0.975,
    df = stats::df.residual(model)
  )

  tibble::tibble(
    control_mean = control_mean,
    perturbation_mean = perturbation_mean,
    score_difference = score_difference,
    standardized_effect = standardized_effect,
    standard_error = standard_error,
    confidence_interval_low =
      score_difference - critical_value * standard_error,
    confidence_interval_high =
      score_difference + critical_value * standard_error,
    p_value = p_value,
    model_status = "fitted"
  )
}
```

## Step 4: Fit every planned perturbation contrast

This first analysis uses the original sequencing perturbations. MorphSeq
phenotype-group contrasts remain separate because they require the preferred
phenotype labels to be joined at the embryo level.

```{r fit-models}
result_rows <- list()
result_number <- 1L

for (contrast_number in seq_len(nrow(contrasts_to_fit))) {
  contrast <- contrasts_to_fit[contrast_number, ]

  contrast_cell_types <- eligible_cell_types |>
    dplyr::filter(
      .data$contrast_name == contrast$contrast_name
    ) |>
    dplyr::pull("cell_type")

  contrast_scores <- embryo_module_scores |>
    dplyr::filter(
      .data$timepoint == contrast$timepoint,
      .data$perturbation %in%
        c(contrast$perturbation, contrast$control),
      .data$cell_type %in% contrast_cell_types,
      .data$module_id %in% CHOSEN_CILIA_MODULES,
      .data$n_cells >= MINIMUM_CELLS_PER_EMBRYO
    ) |>
    dplyr::mutate(
      arm = dplyr::if_else(
        .data$perturbation == contrast$perturbation,
        "perturbation",
        "control"
      )
    )

  perturbation_rt_blocks <- contrast_scores |>
    dplyr::filter(.data$arm == "perturbation") |>
    dplyr::pull("rt_block") |>
    unique() |>
    sort()
  control_rt_blocks <- contrast_scores |>
    dplyr::filter(.data$arm == "control") |>
    dplyr::pull("rt_block") |>
    unique() |>
    sort()
  observed_cross_rt_block <- !identical(
    perturbation_rt_blocks,
    control_rt_blocks
  )

  if (observed_cross_rt_block != contrast$expected_cross_rt_block) {
    stop(
      contrast$contrast_name,
      " does not match its expected cross-RT-block status."
    )
  }

  for (module_name in CHOSEN_CILIA_MODULES) {
    for (cell_type_name in contrast_cell_types) {
      model_data <- contrast_scores |>
        dplyr::filter(
          .data$module_id == .env$module_name,
          .data$cell_type == .env$cell_type_name
        )

      model_result <- fit_one_module_model(model_data)

      arm_summary <- model_data |>
        dplyr::group_by(.data$arm) |>
        dplyr::summarise(
          n_embryos = dplyr::n_distinct(.data$embryo_ID),
          median_cells_per_embryo = stats::median(.data$n_cells),
          minimum_cells_observed = min(.data$n_cells),
          .groups = "drop"
        )

      result_rows[[result_number]] <- model_result |>
        dplyr::mutate(
          contrast_name = contrast$contrast_name,
          perturbation = contrast$perturbation,
          control = contrast$control,
          timepoint = contrast$timepoint,
          module_id = module_name,
          cell_type = cell_type_name,
          expected_cross_rt_block = contrast$expected_cross_rt_block,
          perturbation_rt_blocks = paste(
            perturbation_rt_blocks,
            collapse = ";"
          ),
          control_rt_blocks = paste(
            control_rt_blocks,
            collapse = ";"
          ),
          n_control_embryos = arm_summary$n_embryos[
            arm_summary$arm == "control"
          ],
          n_perturbation_embryos = arm_summary$n_embryos[
            arm_summary$arm == "perturbation"
          ],
          median_control_cells = arm_summary$median_cells_per_embryo[
            arm_summary$arm == "control"
          ],
          median_perturbation_cells = arm_summary$median_cells_per_embryo[
            arm_summary$arm == "perturbation"
          ],
          minimum_cells_per_embryo = MINIMUM_CELLS_PER_EMBRYO,
          .before = 1
        )

      result_number <- result_number + 1L
    }
  }
}

module_results <- dplyr::bind_rows(result_rows)
```

## Step 5: Apply McClintock-style BH correction

Within each contrast, all eligible cell type × module p-values form one testing
family. This uses the same `stats::p.adjust(..., method = "BH")` convention as
the McClintock abundance output.

```{r adjust-p-values}
module_results <- module_results |>
  dplyr::group_by(.data$contrast_name) |>
  dplyr::mutate(
    q_value = stats::p.adjust(
      .data$p_value,
      method = "BH"
    ),
    is_fdr_significant =
      !is.na(.data$q_value) & .data$q_value < FDR_THRESHOLD
  ) |>
  dplyr::ungroup() |>
  dplyr::arrange(
    .data$contrast_name,
    .data$q_value,
    dplyr::desc(abs(.data$standardized_effect)),
    .data$module_id,
    .data$cell_type
  )

readr::write_tsv(module_results, RESULT_FILE)

cat(
  "Wrote", RESULT_FILE, "with", nrow(module_results), "tests\n",
  sum(module_results$is_fdr_significant),
  "tests have BH q <", FDR_THRESHOLD, "\n"
)
```

## Step 6: Inspect the strongest results

The table below is descriptive. Statistical interpretation should consider the
effect size, confidence interval, embryo counts, module biology, and RT-block
status—not the q-value alone.

```{r inspect-results}
significant_results <- module_results |>
  dplyr::filter(.data$is_fdr_significant)

significant_results |>
  dplyr::select(
    contrast_name,
    module_id,
    cell_type,
    control_mean,
    perturbation_mean,
    score_difference,
    standardized_effect,
    confidence_interval_low,
    confidence_interval_high,
    p_value,
    q_value,
    n_control_embryos,
    n_perturbation_embryos,
    expected_cross_rt_block
  )
```

Summarize the number of significant cell types without hiding the complete
result table.

```{r summarize-results}
module_results |>
  dplyr::group_by(
    .data$contrast_name,
    .data$module_id
  ) |>
  dplyr::summarise(
    n_tested_cell_types = dplyr::n(),
    n_fdr_significant = sum(.data$is_fdr_significant),
    .groups = "drop"
  )
```
